### 2021-02-27 doc/article.tex
Rough outline of article latex document and IOP required statements

\documentclass[12pt]{iopart}

\usepackage{graphicx}
\graphicspath{{../fig/}}

\begin{document}

\title[PREPRINT by Doss-Gollin, Farnham, Lall, and Modi]{How unprecedented was the February 2021 Texas cold snap?}

\author{James Doss-Gollin$^1$, David J. Farnham$^2$, Upmanu Lall,$^3$ and Vijay Modi$^4$}
\address{$^1$ Department of Civil and Environmental Engineering, Rice University, Houston, TX}
\address{$^2$ Demo}
\address{$^3$ Demo}
\address{$^4$ Demo}
\ead{jdossgollin@rice.edu}
\vspace{10pt}

\begin{abstract}
  We need to write an abstract!
\end{abstract}
\noindent{\it Keywords: Climate Dynamics, Cold Extremes, Energy Demand, Texas}

%\submitto{\ERL}
\maketitle
\nocite{*}

\section{Introduction}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{historic_events_era5.pdf}
  \caption{CAPTION}\label{fig:historic_era5}
\end{figure}

\section{Methods}

\section{Results}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{local_rt_era5.pdf}
  \caption{CAPTION}\label{fig:local_era5}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{ERCOT_HDD_IDF_MLE_popweighted.pdf}
  \caption{CAPTION}\label{fig:idf_weighted}
\end{figure}

\section{Discussion}

\section{Conclusions}

\clearpage
\section*{Acknowledgements}

The authors thank Adam Massmann and Vivek Srikrishnan for insightful comments on a preliminary draft of this manuscript.
The authors thank\ldots.

\bibliographystyle{unsrt}
\bibliography{references}

\clearpage
\section{Supplemental Results}
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{historic_events_bk.pdf}
  \caption{CAPTION}\label{fig:historic_bk}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{historic_events_era5_TX.pdf}
  \caption{CAPTION}\label{fig:historic_tx}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{local_rt_ghcnd.pdf}
  \caption{CAPTION}\label{fig:local_ghcnd}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{ERCOT_HDD_IDF_MLE_unweighted.pdf}
  \caption{CAPTION}\label{fig:idf_unweighted}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{ERCOT_HDD_IDF_plotpos_popweighted.pdf}
  \caption{CAPTION}\label{fig:idf_lmoments_weighted}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{ERCOT_HDD_IDF_plotpos_unweighted.pdf}
  \caption{CAPTION}\label{fig:idf_lmoments_unweighted}
\end{figure}

\end{document}

### 2021-02-27 scripts/berkeley_earth_aggregate.py
Calculate anomalies based on DJF mean, not calendar month

@@ -8,9 +8,13 @@ import numpy as np
 import pandas as pd
 import xarray as xr
 
+from codebase.calc import calculate_anomaly
+
 
 def parse_file(fname: str) -> xr.Dataset:
     """Read in one of the data files"""
+
+    # step 1: figure out what the underlying temperature value is
     bkds = xr.open_dataset(fname).sel(
         latitude=slice(25, 50), longitude=slice(-125, -65)
     )
@@ -24,7 +28,10 @@ def parse_file(fname: str) -> xr.Dataset:
     anomaly_c = bkds["temperature"]
     temp_c = anomaly_c + climatology_c
 
-    return xr.Dataset(dict(anomaly_f=anomaly_c * 9 / 5, temp_f=temp_c * 9 / 5 + 32))
+    # step 2: re-derive climatology using consistent formulat & calc new anomaly
+    anomaly_c2 = calculate_anomaly(temp_c, groupby_type="time.season")
+
+    return xr.Dataset(dict(anomaly_f=anomaly_c2 * 9 / 5, temp_f=temp_c * 9 / 5 + 32))
 
 
 def main() -> None:

### 2021-02-27 scripts/codebase/calc.py

@@ -43,3 +43,9 @@ def cold_return_period(y: float, x: np.ndarray) -> float:
     are annual MINIMA rather than MAXIMA
     """
     return return_period(-y, -x)
+
+
+def calculate_anomaly(da, groupby_type="time.month"):
+    """Function to calculate anomalies from xarray docs"""
+    clim = da.groupby(groupby_type).mean(dim="time")
+    return da.groupby(groupby_type) - clim

### 2021-02-27 scripts/era5_daily_summaries.py

@@ -8,19 +8,9 @@ import argparse
 from pandas import Timedelta
 import xarray as xr
 
-CLIMATOLOGY_SDATE = "1979-12-31"
-CLIMATOLOGY_EDATE = "2020-02-28"
+from codebase.calc import calculate_anomaly
+
 TIME_ZONE_OFFSET = 6  # group by 1 day in Texas, not UTC
-
-
-def calculate_anomaly(da, groupby_type="time.month"):
-    """From Xarray docs"""
-    clim = (
-        da.sel(time=slice(CLIMATOLOGY_SDATE, CLIMATOLOGY_EDATE))
-        .groupby(groupby_type)
-        .mean(dim="time")
-    )
-    return da.groupby(groupby_type) - clim
 
 
 def main() -> None:
@@ -50,8 +40,8 @@ def main() -> None:
 
     # calculate daily climatology and anomalies
     daily_anomaly = calculate_anomaly(
-        temp.resample(time="1D").mean(), groupby_type="time.month"
-    )
+        temp.resample(time="1D").mean(), groupby_type="time.season"
+    )  # time.season ==> DJF
 
     # save *daily* data
     xr.Dataset(

### 2021-02-28 doc/article.tex
Figures for circulation to coauthors

@@
 
+\usepackage{url}
+\usepackage{siunitx}
+\usepackage{cleveref}
+
 \begin{document}
@@
 
-\author{James Doss-Gollin$^1$, David J. Farnham$^2$, Upmanu Lall,$^3$ and Vijay Modi$^4$}
-\address{$^1$ Department of Civil and Environmental Engineering, Rice University, Houston, TX}
-\address{$^2$ Demo}
-\address{$^3$ Demo}
-\address{$^4$ Demo}
+\author{James Doss-Gollin$^1$, David J. Farnham$^2$, Upmanu Lall,$^{3,4}$ and Vijay Modi$^5$}
+\address{$^1$ Department of Civil and Environmental Engineering, Rice University, Houston, TX, USA (ORCID 0000-0002-3428-2224)}
+\address{$^2$ Department of Global Ecology, Carnegie Institution for Science, Stanford, CA, USA (ORCID 0000-0002-6690-4251)}
+\address{$^3$ Columbia Water Center, Columbia University, New York, NY, USA (ORCID 0000-0003-0529-8128)}
+\address{$^4$ Department of Earth and Environmental Engineering, Columbia University, New York, NY, USA}
+\address{$^4$ Department of Mechanical Engineering, Columbia University, New York, NY, USA (ORCID 0000-0003-2513-0437)}
 \ead{jdossgollin@rice.edu}
@@
 \begin{abstract}
-  We need to write an abstract!
+  Winter storm Uri brought severe cold to the southern United States in February 2021, causing a cascading failure of interdependent systems in Texas where infrastructure was not adequately prepared for such cold temperatures.
+  In particular, the failure of interconnected energy systems limited electricity supply just as heating demands spiked, leaving millions of Texans without heat or electricity, many for several days.
+  This motivates the question: was the cold that caused this infrastructure failure a ``black swan'' that could not have been anticipated, or did historical storms provide a precedent?
+  We compute the population weighted temperature excursion below \SI{68}{\degree F} as a proxy for heating demand and use this inferred demand for heating to explore how rare this event was using historical data.
+  We find that temperatures during a storm in December 1989 were more intense than those recorded during February 2021, and that several other storms were nearly as severe.
+  Although some specific locations experienced record cold, the spatially aggregated temperatures were intense but not unprecedented.
+  Given anticipated population growth, future storms may lead to even greater infrastructure failures if adaptive investments are not made.
+  Further, electricity system managers should anticipate that upward trends in electrification of heating may cause peak annual loads on the Texas interconnect to occur during winter storms.
+
 \end{abstract}
-\noindent{\it Keywords: Climate Dynamics, Cold Extremes, Energy Demand, Texas}
+\noindent{\it Keywords: Energy, Electricity, Texas, Natural Hazards, Climate Resilience}
 
@@
 \maketitle
-\nocite{*}
 
 \section{Introduction}
+
+Between February 14th and 17th, 2021, a northern air mass blanketed much of the continental United States, causing anomalously low surface temperatures across the Great Plains.
+While polar air excursions are not unusual, they do not typically reach as far South as the Mexican border.
+The state of Texas was particularly hard hit, with coincident and cascading failures of natural gas production, power generation, transportation, and water systems leaving millions of Texans without electricity, heat, and water for several days.
+The intensity and widespread nature of this weather event underscore a high vulnerability to cold temperatures in places that did not experience a comparable event in preceding decades.
+Even as energy demand for winter heating in Texas has grown dramatically with population and economic growth, the system’s vulnerability to cold affected lives and property and brought the Texas electricity grid within minutes of collapse.
+
+Since production and distribution of natural gas and electricity is possible under conditions far colder than any Texas experienced in February 2021, energy system failures reflect inadequate preparedness for cold.
+If this were an unprecedented or very rare event then it is possible that the impacts resulted from an event outside the design envelope.
+On the other hand, if similar events were in the record then one would expect a higher level of preparation.
+It is therefore important to assess whether the February 2021 cold snap could have been anticipated.
+To answer this question, we first compute the population weighted temperature excursion below 68F as a proxy for the unknown heating demand, and use this metric to explore how rare this event was using historical data.
+Next, we conduct a spatially distributed analysis to assess how unexpected the cold experienced by local roads, water mains, gas pipelines, energy generation facilities, and critical infrastructure installations was across Texas.
+We conclude by discussing the implications of these findings for long-term electricity systems planning given anticipated population growth and electrification.
+
+\subsection{Previous Cold Snaps in Texas}
+
+It is well documented that Texas has previously experienced severe cold snaps, most notably in 1899, 1951, 1983, 1989, and 2011.
+The specific spatiotemporal structure of a cold event, and its correspondence with population centers, modulates the grid-wide demand for heating (\cref{sec:inferred-demand}).
+The structure of the storm also drives the aggregated hazard to energy infrastructure, which has implications for the costs and benefits of infrastructure hardening.
+It is therefore important to assess the weather conditions that led to these infrastructure failures and answer the question: were the failures of the energy system that occurred during February 2021 cold snap a manifestation of a storm that was largely unprecedented in its spatiotemporal form, or were there similar past storms in the historic temperature records?
+
+\Cref{fig:historic_era5,fig:historic_bk,fig:historic_tx} show historic events that caused severe cold in Texas.
+Although the spatiotemporal structure of each event is distinct, it is apparent that cold extremes in Texas tend to co-occur with cold temperatures across much of the United States, particularly the Great Plains.
+While the 2021 event was severe, daily temperatures in Texas appear comparable to historical events.
+The ``Great Blizzard'' of February 1899, shown in figure \cref{fig:historic_bk}, shows substantially more intense cold.
 
@@
   \includegraphics[width=\textwidth]{historic_events_era5.pdf}
-  \caption{CAPTION}\label{fig:historic_era5}
+  \caption{
+    Severe cold snaps that affect Texas and extend into the Central United States have several precedents in the historical record.
+    Plot shows anomalies of daily mean temperatures for historic major cold events affecting Texas, defined as the departure from the seasonal (December-February) mean to facilitate comparison of events in different calendar months.
+    Temperatures over the Continental United States are plotted to show large-scale patterns, but events are selected based only on Texas temperatures.
+    Hourly temperatures are averaged to 1-day (a,d,g,j,m), 3-day (b,e,h,k,n), and 5-day (c,f,i,l,o) average temperature anomalies.
+  }\label{fig:historic_era5}
 \end{figure}
 
-\section{Methods}
+\section{Data and Methods}
 
-\section{Results}
+We use three temperature datasets to analyze the recent and historical cold snap in Texas through the lens of distributed (each grid cell analyzed separately) and aggregated (weighted averages taken across space) extreme values analysis.
+
+\subsection{Datasets}
+
+We use three temperature datasets to ensure robust findings:
+\begin{enumerate}
+  \item Hourly \SI{2}{\meter} air temperature reanalysis on a \SI{0.25}{\degree} grid from the ERA-5 reanalysis project produced by the European Centre for Medium Range Weather Forecasting \cite{hersbach_era5:2020} and available  from the Copernicus Data Store (\url{https://cds.climate.copernicus.eu}) from 1950 to the present.
+All plots shown in the main text use the ERA-5 data, but supplemental figures use other data sets.
+  \item Daily mean, minimum and maximum temperatures, gridded to \SI{1}{\degree}, produced by Berkeley Earth (\url{http://berkeleyearth.org/data/}).
+This gridded product is based on statistical analysis of station data and is available from 1880 to the present.
+This data is considered an experimental product, so we use it only for comparative purposes.
+  \item To complement blended gridded data products, we use station temperature data from the Global Historical Climatology Network (GHCN) dataset compiled by the National Ocean and Atmospheric Administration \cite{Menne:2012hk} and available at \url{https://www1.ncdc.noaa.gov/pub/data/ghcn/daily/}.
+This dataset provides daily mean, maximum, and minimum temperature.
+These measurements represent point measurements, which can differ in important ways from gridded products describing spatial averages due to spatial heterogeneity of temperature.
+We retain stations within the state of Texas if they provide at least 60 years of data and if they contain observations for the set of historical cold extremes shown in \cref{fig:historic_era5}.
+\end{enumerate}
+We also use population density data from the GPWv4 dataset \cite{ciesin_gpwv4:2016}, a list of power generation facilities from the US Energy Information Administration \cite{useia_generators:2021}, and a map of the Texas Interconnect \cite{useia_regions:2021}.
+
+\subsection{Inferred Demand for Heating}\label{sec:inferred-demand}
+
+Most space heating in Texas is either electric or gas \cite{waite_heating:2020} and the majority of power generation in the Texas Interconnect depends on natural gas \cite{everhart_iea:2021}.
+Stress on natural gas production and delivery was therefore just as important as the more visible stress on the electric system.
+
+The hourly or daily thermal energy requirement for space heating is primarily driven by how much lower the ambient temperature is from an indoor comfort temperature of \SI{68}{\degree F}.
+This relationship is often expressed in terms of heating degree days or hours.
+We therefore consider the temperature excursion from \SI{68}{\degree F} as a proxy for thermal heating demand.
+We compute this value each hour for the ERA5 data, defining heating demand at each grid cell as $\text{HD}_t = \max (68 - T_t, 0)$, where $T_t$ is the temperature at hour $t$ in \si{\degree F}.
+The Berkeley Earth and GHCN datasets provide daily minimum and maximum temperatures, so we define heating demand at each grid cell or station as $\text{HD}_d = \max (68 -\frac{T_{\text{min},d} + T_{\text{max},d}}{2}, 0)$, where $T_{\text{min},d}$ is the minimum temperature recorded on day $d$ and $T_{\text{max},d}$ is the maximum temperature recorded on day $d$, both in \si{\degree F}.
+
+To assess how spatially correlated cold spells might affect the Texas electric grid, we average heating demand in space over the Texas Interconnect domain \cite{useia_regions:2021}, weighting each grid cell by 2020 population density \cite{ciesin_gpwv4:2016}.
+We refer to this spatially aggregated metric, which has the straightforward interpretation as the average temperature below \SI{68}{\degree F} experienced by Texas residents, as ``Inferred Demand for Heating.''
+
+\subsection{Return Period}
+
+For each event duration considered, return periods are obtained by fitting a stationary generalized extreme value (GEV) distribution to the time series of annual minimum December-February (DJF) temperatures.
+Events that occur in December are coded to the following year.
+The 2021 winter season was excluded from return period estimates, allowing interpretation of return periods as \emph{a priori} estimates.
+
+\subsection{Cold Duration}
+
+The effect of cold temperatures on energy demand and critical infrastructure depends on how long the cold persists.
+Short duration cold snaps can kill plants, freeze exposed pipes, and contribute to dangerous roadway conditions.
+Longer duration cold spells contribute to demand for heating and energy and cause pipes to burst even if they have some insulation.
+We calculate demand for heating by taking temporal averages over a range of durations from 1 hour to 4 days.
+
+\subsection{Code and Data}
+
+We are committed to open science.
+Our open source code is freely available in a live repository at \url{https://github.com/jdossgollin/2021-TXtreme} and in an archived repository at TODO.
+Reproducibility is facilitated through the Snakemake library \cite{koster_snakemake:2012}.
+
+\section{How extreme was inferred heating demand over the Texas interconnect?}
+
+The total shock to Texas heating demand is partially determined by the extent to which cold snaps impact multiple population centers simultaneously.
+As such, understanding whether this cold snap had precedent with respect to its feature of impacting several populated regions of Texas' grid concurrently is critical.
+We therefore use our measure of inferred demand for heating (see \cref{sec:inferred-demand}) to represent the aggregate heating demand induced by cold temperatures.
+Aggregating historic temperature fields in space using the 2020 population, we answer the question ``what would the aggregate demand for heating have been had historic cold snaps occurred today?''
+
+\Cref{fig:idf_weighted} shows that the intensity, duration, and recurrence intervals of the February 2021 storm are severe but not unprecedented in the historical record.
+For example, at the 6 hour duration the December 1989 storm was substantially more intense and other storms including February 1951 are nearly as intense.
+At the two day duration, the 2021 and 1989 events were approximately equally intense and other storms including December 1983 are nearly as intense.
+The 2011 storm, which caused rolling blackouts and motivated research into energy system resilience to cold \cite{ferc_outages:2011}, was quite modest by comparison.
+The right panel shows statistical return periods for these extreme events, though these estimates are sensitive to small changes in demand for heating.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{ERCOT_HDD_IDF_MLE_popweighted.pdf}
+  \caption{
+    The aggregated demand for heating induced by the February 2021 cold snap is not unprecedented.
+    For the worst 6 hours, the 1989 event was more severe than the 2021 event, while they are comparable for the longer durations.
+    (a): time series of spatially averaged annual maximum heating degree days weighted by 2020 population for 6 hour and 2 day durations.
+    December extremes, including the December 1989 storm, are coded to the following year so that one maximum per December-February winter season is taken.
+    (b): the intensity-duration-frequency intervals estimated using 1950-2020 data (i.e., not using the 2021 event), overlaid by the annual maxima from the 1989, 2011, and 2021 events.
+    Gray dashed lines indicate 2, 5, 10, 25, 50, and 100 year return levels.
+  }\label{fig:idf_weighted}
+\end{figure}
+
+\section{Spatially distributed temperature extremes}
+
+It is difficult to establish a spatially aggregated proxy for supply-side risk given complex interlinkages between natural gas, electric, and other systems which create the possibility for cascading failures as observed in February 2021.
+Water treatment and distribution systems, as well as other essential services, also rely on electricity, further increasing vulnerabilities.
+Instead of aggregating this risk in space, we estimate the exceedance probability of the February 2021 temperatures at each grid cell separately to shed light on the severity of cold experienced by specific installations across the region.
+
+\Cref{fig:local_era5} shows local return periods for the February 2021 cold snap at 6 hour, 1 day, 2 day, and 4 day durations.
+Other than a band from south-central to south-east Texas, nearly all regions of the Texas Interconnect (gray outline in \cref{fig:local_era5}a,d) experienced cold with a return period below 50 years.
+Results are similar using station data (\cref{fig:local_ghcnd}).
+Importantly for the energy system, the band experiencing cold with return period greater than 50 years includes a substantial fraction of Texas’s population (\cref{fig:local_ghcnd}a) and natural gas generation (\cref{fig:local_ghcnd}d).
+Outside the Texas Interconnect region, much of the Texas Panhandle and central Oklahoma experienced intense cold with a return period greater than 100 years.
+Although Oklahoma recorded some water main breaks \cite{crum_water:2021} and rolling blackouts for several hours \cite{money_oklahoma:2021}, infrastructure impacts were less severe than in Texas despite recording more extreme (relative to its own historical record) temperatures.
 
@@
   \includegraphics[width=\textwidth]{local_rt_era5.pdf}
-  \caption{CAPTION}\label{fig:local_era5}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{ERCOT_HDD_IDF_MLE_popweighted.pdf}
-  \caption{CAPTION}\label{fig:idf_weighted}
+  \caption{
+    Return periods for the February 2021 event, calculated using stationary estimates of annual extremes over the period 1950-2020.
+    Return periods are calculated separately for each cell.
+    (a): estimates of 2020 population density \cite{ciesin_gpwv4:2016}.
+    (d): energy generation facilities in Texas \cite{useia_generators:2021}.
+    (b,c,e,f): local return periods for 6 hour, 24 hour, 48 hour, and 96 hour cold durations, respectively.
+    Contours enclose regions that recorded 50 and 100 year return levels.
+    Gray region with dotted black line in (a) and (d) shows boundaries of the Texas Interconnect \cite{useia_regions:2021}.
+  }\label{fig:local_era5}
 \end{figure}
@@
 
+Our spatially aggregated metric of inferred demand for heating shows that the February 2021 event was significant but not unforeseeable (\cref{fig:idf_weighted}).
+Although specific locations experienced very intense ($>100$ year return period) temperatures, we find that for most locations in Texas the temperatures recorded during the February 2021 cold snap had precedent in the historical record and could have been anticipated given available data.
+
+A proximate cause of load shedding in the Texas Interconnect during February 2021 was the vulnerability of the electricity generation system to the severe cold temperatures \cite{everhart_iea:2021}.
+\Cref{fig:local_era5,fig:local_ghcnd} show that this happened even though most parts of the state had previously experienced similarly intense cold, notably in 1989.
+While our analysis neglects other meteorological factors, like freezing rain, that may have impeded operations of specific facilities, we find that the electricity and energy infrastructure in the Texas Interconnect were collectively vulnerable to the temperatures observed in 1989.
+Similarly, water mains broke in places like Houston where temperatures did not exceed 100-year return levels, underscoring deep vulnerability of critical infrastructure systems \cite{chester_reliable:2020}.
+
+Another cause of load shedding was the high demand for electricity that low temperatures induced.
+Currently over 50\% of Texas households use electricity for space heating over the majority of census tracts of the state \cite{waite_heating:2020,white_txresidential:2019} and further electrification is a central element of many plans to decarbonize our energy sector \cite{williams_decarbonization:2012,davis:2018} in an effort to mitigate the climate risks associated with greenhouse gas emissions.
+While summer peak loads have been a central planning concern on the Texas grid in the past, it is likely that winter peak loads will become a greater concern in the coming decades.
+In fact, the peak demand during this event was estimated at around \SI{74.5}{\giga\watt}, just shy of the all time peak for any season of \SI{74.8}{\giga\watt} in August 2019 \cite{everhart_iea:2021}.
+As electrification of heating increases, severe cold snaps may represent the annual peak demand on the Texas Interconnect.
+
+Our primary findings hold for an alternative gridded dataset and station data (see supplemental material).
+However, calculated return periods are sensitive to the method of estimation (\cref{fig:idf_lmoments_unweighted,fig:idf_lmoments_weighted}).
+Future analysis could address parametric uncertainty, model structure uncertainty \cite{wong_floodrisk:2018}, non-stationarity \cite{Milly:2008dg}, or regime-like modes of climate variability \cite{DossGollin:2019}.
+More fundamentally, an assessment of exposure to cold extremes over the next decades should consider the deeply uncertain distribution of future climate change, and the induced effect on cold extremes in Texas.
+In general, the IPCC anticipates that global warming will lead to more hot extremes and fewer cold extremes \cite{ipcc_ar5:2014}, but the effects of Arctic amplification on mid-latitude regional temperature extremes remains an area of active research \cite{Barnes:2013fp,Cohen:2014gx,Screen:2013ho}.
+However, the effect of climate change on peak demand for heating is likely to be small compared to the effect of rapid population growth (for example, the Texas Water Development Board \cite{texaswaterplan:2012} anticipates at least 40\% growth from 2020-2050).
+
+Our analysis quantifies the probability with which temperature extremes could have been anticipated \emph{a priori}.
+Other factors also govern infrastructure performance and failure, including precipitation, the demand for natural gas in adjacent regions, and complex connections within and between regional systems.
+Similarly, decisions at multiple time scales, including disaster preparedness and risk communication, contribute to the human consequences of physical infrastructure failure.
+Thus, the exact chain of events that led to the blackouts and water system disruptions during February 2021 should be sorted out only after thorough investigations by parties on the ground in Texas.
+
 \section{Conclusions}
 
-\clearpage
+The February 2021 cold snap was the most intense in 30 years, but was not without precedent in the full historical record.
+In addition to the record cold conditions of 1899 (\cref{fig:historic_bk}), we estimate that the weather of December 1989 would have resulted in slightly higher 6-hour and 2-day aggregate heating demands over the Texas Interconnect compared to February 2021.
+Several other storms since 1950 were nearly as severe.
+Given upward trends in the electrification of heating, it is likely that future cold snaps will cause peak annual loads on the Texas interconnect to occur during the winter season.
+Infrastructure expansion necessitated by a rapidly growing population offers Texas the opportunity to develop a more resilient energy system.
+
 \section*{Acknowledgements}
 
-The authors thank Adam Massmann and Vivek Srikrishnan for insightful comments on a preliminary draft of this manuscript.
-The authors thank\ldots.
+DJF was supported by a gift from Gates Ventures LLC to the Carnegie Institution for Science.
+The authors thank Adam Massmann, Elisabeth Gawthrop, and Vivek Srikrishnan for insightful comments on a preliminary draft of this manuscript.
+We also thank the developers and maintainers of the open source packages we used, in particular the pydata and pangeo communities, and the many journalists and academics whose timely reporting and commentary has informed our thinking.
 
+\section*{Bibliography}
 \bibliographystyle{unsrt}
@@
 
+\subsection{Historic Extreme Temperatures}
+
+To complement \cref{fig:historic_era5}, we plot extreme cold temperatures using alternate data.
+First, \cref{fig:historic_bk} shows historic events over the Continental United States.
+The set of events is slightly different than that of \cref{fig:historic_era5}: the data set does not include the 2021 event, but does include the 1899 ``Great Blizzard.''
+The 1899 event shows more intense and persistent cold than the other events in the dataset.
+Next, \cref{fig:historic_tx} shows the same data as \cref{fig:historic_era5} but zooms in on Texas.
+The 1989 (g-i) and 2021 (m-o) appear to be the most severe events in this data set, but the 1-day cold extremes in the 1989 event are more intense than in the February 2021 event.
+
 \begin{figure}
@@
   \includegraphics[width=\textwidth]{historic_events_bk.pdf}
-  \caption{CAPTION}\label{fig:historic_bk}
+  \caption{
+    As \cref{fig:historic_bk} but the Berkely Earth temperature data is used.
+    The dataset does not contain the 2021 event, but the ``Great Blizzard'' of February 1899 is included.
+    Spatial patterns of cold from this dataset are qualitatively similar to Figure 1.
+    The 1899 event emphasizes that the modern historical record does not yield a full sample from the full distribution of possible hazards.
+  }\label{fig:historic_bk}
 \end{figure}
@@
   \includegraphics[width=\textwidth]{historic_events_era5_TX.pdf}
-  \caption{CAPTION}\label{fig:historic_tx}
+  \caption{
+    As \cref{fig:historic_bk} but only Texas is shown.
+  }\label{fig:historic_tx}
 \end{figure}
+
+\subsection{Spatially distributed temperature extremes}
+
+To complement \cref{fig:local_era5}, we compute local return periods using station data from the GHCN \cite{Menne:2012hk}.
+\Cref{fig:local_ghcnd} shows the return periods of the February 2021 cold snap for 1, 2, 3, and 4 day durations.
+Only stations with at least 60 years of data are considered, and since the locations of these stations are not chosen at random, this does not constitute a representative sample of all points across Texas.
+However, the spatial pattern matches that of \cref{fig:local_era5}, with a band of severe cold stretching from South Central to East Texas and in the Texas Panhandle.
+
+\subsection{Inferred demand for heating}
+
+To complement our analysis of inferred demand for heating, we consider how results change when two variables are changed.
+First, we consider what happens if the spatial field demand for heating is aggregated using grid cell area rather than population density.
+Next, we compute return periods using an estimator based on the method of $L$-moments.
+Although $L$-moment estimators for the generalized extreme value distribution are not unbiased, they are popular in the statistical hydrology literature for their stability \cite{hosking_gev:1985,martins_gev:2001,morrison_gev:2002}.
+
+We draw two conclusions from these plots.
+First, the 2021 event appears more severe if grid cells are weighted by population density (\cref{fig:idf_weighted,fig:idf_lmoments_weighted}) than if they are weighted only by area \cref{fig:idf_unweighted,fig:idf_lmoments_unweighted}).
+This is consistent with our observation of a correspondence between the most extreme temperatures in February 2021 and population density (\cref{fig:local_era5}).
+By contrast, the 2011 event appears more extreme when grid cells are weighted by area, which is consistent with \cref{fig:historic_tx,fig:historic_era5,fig:historic_bk} showing the coldest temperatures in relatively less populated West Texas.
+Second, the $L$-moment estimators (\cref{fig:idf_lmoments_weighted,fig:idf_lmoments_unweighted}) assign a lower return period to the 1989 and 2021 events than the maximum likelihood estimators (\cref{fig:idf_unweighted,fig:idf_weighted}).
+
+To provide some context for our inferred demand for heating metric, \cref{fig:hdd_ts} plots its time series during the peak of the February 2021 cold snap.
 
@@
   \includegraphics[width=\textwidth]{local_rt_ghcnd.pdf}
-  \caption{CAPTION}\label{fig:local_ghcnd}
+  \caption{
+    As \cref{fig:local_era5} but return periods are calculated using station data from the GHCN data set \cite{Menne:2012hk}.
+    Black circles indicate that a station exceeded its own record for a particular duration.
+  }\label{fig:local_ghcnd}
 \end{figure}
@@
   \includegraphics[width=\textwidth]{ERCOT_HDD_IDF_MLE_unweighted.pdf}
-  \caption{CAPTION}\label{fig:idf_unweighted}
+  \caption{
+    As \cref{fig:idf_unweighted} but grid cells are weighted by area $A=\cos(\phi)$ where $\phi$ is latitude.
+  }\label{fig:idf_unweighted}
 \end{figure}
@@
   \includegraphics[width=\textwidth]{ERCOT_HDD_IDF_plotpos_popweighted.pdf}
-  \caption{CAPTION}\label{fig:idf_lmoments_weighted}
+  \caption{
+    As \cref{fig:idf_unweighted} but return periods are calculated using the $L$-moments estimator.
+  }\label{fig:idf_lmoments_weighted}
 \end{figure}
@@
   \includegraphics[width=\textwidth]{ERCOT_HDD_IDF_plotpos_unweighted.pdf}
-  \caption{CAPTION}\label{fig:idf_lmoments_unweighted}
+  \caption{
+    As \cref{fig:idf_unweighted} but return periods are calculated using the $L$-moments estimator.
+  }\label{fig:idf_lmoments_unweighted}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{HDD_pop_weighted_ts.pdf}
+  \caption{
+    A time series of population weighted temperature excursion from \SI{68}{\degree F} over the Texas Interconnect during the most
+  }\label{fig:hdd_ts}
 \end{figure}

### 2021-02-28 doc/coverletter.tex
Figures for circulation to coauthors

%----------------------------------------------------------------------------------------
%	DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------
\documentclass{scrartcl}
\usepackage[
	backaddress=off,
	fromalign=right,
]{scrletter}

\usepackage[english]{babel}
\usepackage{libertine}
\usepackage{natbib}
\usepackage[letterpaper]{geometry}

\makeatletter
\@setplength{backaddrheight}{0pt}
\let\@texttop\relax
\makeatother

%----------------------------------------------------------------------------------------
%	YOUR INFORMATION AND LETTER DATE
%----------------------------------------------------------------------------------------

\setkomavar{fromname}{Dr. James Doss-Gollin} % Your name used in the from address
\setkomavar{fromaddress}{Department of Civil and Environmental Engineering\\Rice University\\Houston, TX} % Your address
\setkomavar{signature}{James Doss-Gollin, on behalf of coauthors David J. Farnham, Upmanu Lall, and Vijay Modi} % Your name used in the signature

\date{\today} % Date of the letter

%----------------------------------------------------------------------------------------

\begin{document}

%----------------------------------------------------------------------------------------
%	ADDRESSEE
%----------------------------------------------------------------------------------------

\begin{letter}{Dr. Daniel M Kammen\\Editor-in-Chief\\Environmental Research Letters} % Addressee name and address

	%----------------------------------------------------------------------------------------
	%	LETTER CONTENT
	%----------------------------------------------------------------------------------------

	\opening{Dear Dr. Kammen and Co-Editors,}

	I am writing to submit our manuscript entitled ```how unprecedented was the February 2021 Texas cold snap?'' for consideration to Environmental Research Letters.

	In light of the recent cold snap that caused cascading infrastructure failures in Texas, we sought to understand the extent to which the temperatures were unprecedented in the historical record.
	We have tried to do some rapid yet rigorous work.
	Specifically, we deploy an empirical methodology to assess whether the February 2021 temperatures had precedent in the historical record.
	We find that:
	\begin{enumerate}
		\item the February 2021 cold snap in Texas was severe but not unprecedented: notably, an event in December 1989 was colder at most locations;
		\item the coldest extremes in February 2021 coincided with Texas's population centers, increasing net demand for heating across the Texas Interconnect;
		\item anticipated population growth and electrification of heating will likely increase electricity demand during cold spells, posing a challenge for grid reliability; and
		\item design and operation of interdependent critical infrastructure should consider the full distribution of climate hazard, not merely the most recent events.
	\end{enumerate}
	As the Texas state legislature and other regulatory bodies investigate the causes of this event with an eye to preventing future disasters, a credible assessment of the degree to which these temperatures could have been anticipated offers a significant and relevant contribution to public and scientific discourse.

	Our findings also have implications for the broader literature around electrification and decarbonization.
	Deep decarbonization is critical to avoiding severe climate change scenarios.
	\citet{williams_decarbonization:2012} argue, as others have done, that widespread electrification is needed to achieve decarbonization
	More recently, \citet{waite_heating:2020} quantified the load effects of electrifying U.S. residential and commercial space heating using ten years of weather data.
	Their analysis indicated that most regions of Texas could expand up to 70 to 100\% electric heating without impacting local peak annual load, which generally occurs during the summer cooling season.
	This recent cold snap indicates that Texas may be closer to experiencing peak electricity loads during the winter than we had previously anticipated.
	Our results suggest that electrification of building heating could increase peak electricity demands, requiring some combination of increased generation, storage, and/or long-distance transmission to achieve the high reliability levels that are typically considered acceptable.

	Each named author has substantially contributed to conducting the underlying research and drafting this manuscript.
	Additionally, to the best of our knowledge, the named authors have no conflict of interest, financial or otherwise.

	Thank you for your consideration.

	\closing{Sincerely,}

	\bibliographystyle{agu}
	\bibliography{references}

	%----------------------------------------------------------------------------------------

\end{letter}

\end{document}

### 2021-02-28 doc/article.tex
My edits

@@
 
-\title[PREPRINT by Doss-Gollin, Farnham, Lall, and Modi]{How unprecedented was the February 2021 Texas cold snap?}
+\title[PREPRINT: Doss-Gollin, Farnham, Lall, and Modi]{How unprecedented was the February 2021 Texas cold snap?}
 
@@
 \begin{abstract}
-  Winter storm Uri brought severe cold to the southern United States in February 2021, causing a cascading failure of interdependent systems in Texas where infrastructure was not adequately prepared for such cold temperatures.
+  Winter storm Uri brought severe cold to the southern United States in February 2021, causing a cascading failure of interdependent systems in Texas where infrastructure was not adequately prepared for such cold.
   In particular, the failure of interconnected energy systems limited electricity supply just as heating demands spiked, leaving millions of Texans without heat or electricity, many for several days.
-  This motivates the question: was the cold that caused this infrastructure failure a ``black swan'' that could not have been anticipated, or did historical storms provide a precedent?
-  We compute the population weighted temperature excursion below \SI{68}{\degree F} as a proxy for heating demand and use this inferred demand for heating to explore how rare this event was using historical data.
-  We find that temperatures during a storm in December 1989 were more intense than those recorded during February 2021, and that several other storms were nearly as severe.
-  Although some specific locations experienced record cold, the spatially aggregated temperatures were intense but not unprecedented.
+  This motivates the question: was the cold that contributed to this infrastructure failure a ``black swan'' that could not have been anticipated, or did historical storms provide a precedent?
+  We compute the population weighted temperature excursion below \SI{68}{\degree F} as a proxy for heating demand and use this inferred demand for heating to explore how rare the temperatures recorded during the February 2021 cold snap were using historical data.
+  We find that temperatures during a storm in December 1989 were more intense than those recorded during February 2021, and that several other storms were nearly as intense.
+  Although some specific locations experienced record cold, temperatures at most locations were cold but not unprecedented.
   Given anticipated population growth, future storms may lead to even greater infrastructure failures if adaptive investments are not made.
-  Further, electricity system managers should anticipate that upward trends in electrification of heating may cause peak annual loads on the Texas interconnect to occur during winter storms.
+  Further, electricity system managers should anticipate that upward trends in electrification of heating may cause peak annual loads on the Texas Interconnect to occur during winter storms.
 
@@
 The intensity and widespread nature of this weather event underscore a high vulnerability to cold temperatures in places that did not experience a comparable event in preceding decades.
-Even as energy demand for winter heating in Texas has grown dramatically with population and economic growth, the system’s vulnerability to cold affected lives and property and brought the Texas electricity grid within minutes of collapse.
+Even as energy demand for winter heating in Texas has grown dramatically with population and economic development, the system’s vulnerability to cold jeopardized lives and property and brought the Texas electricity grid within minutes of collapse.
 
@@
 It is therefore important to assess whether the February 2021 cold snap could have been anticipated.
-To answer this question, we first compute the population weighted temperature excursion below 68F as a proxy for the unknown heating demand, and use this metric to explore how rare this event was using historical data.
+To answer this question, we first compute the population weighted temperature excursion below \SI{68}{\degree F} as a proxy for the unknown heating demand, and use this metric to explore how rare this event was using historical data.
 Next, we conduct a spatially distributed analysis to assess how unexpected the cold experienced by local roads, water mains, gas pipelines, energy generation facilities, and critical infrastructure installations was across Texas.
@@
 
-It is well documented that Texas has previously experienced severe cold snaps, most notably in 1899, 1951, 1983, 1989, and 2011.
+It is well documented that Texas has previously experienced severe cold, most notably in 1899, 1951, 1983, 1989, and 2011.
 The specific spatiotemporal structure of a cold event, and its correspondence with population centers, modulates the grid-wide demand for heating (\cref{sec:inferred-demand}).
 The structure of the storm also drives the aggregated hazard to energy infrastructure, which has implications for the costs and benefits of infrastructure hardening.
-It is therefore important to assess the weather conditions that led to these infrastructure failures and answer the question: were the failures of the energy system that occurred during February 2021 cold snap a manifestation of a storm that was largely unprecedented in its spatiotemporal form, or were there similar past storms in the historic temperature records?
+It is therefore important to assess the weather conditions that led to these infrastructure failures and identify whether they had historical precedent.
 
-\Cref{fig:historic_era5,fig:historic_bk,fig:historic_tx} show historic events that caused severe cold in Texas.
+\Cref{fig:historic_era5} and supplemental \cref{fig:historic_bk,fig:historic_tx} show historic cold snaps in Texas.
 Although the spatiotemporal structure of each event is distinct, it is apparent that cold extremes in Texas tend to co-occur with cold temperatures across much of the United States, particularly the Great Plains.
-While the 2021 event was severe, daily temperatures in Texas appear comparable to historical events.
-The ``Great Blizzard'' of February 1899, shown in figure \cref{fig:historic_bk}, shows substantially more intense cold.
+While the 2021 event was severe, daily temperature extrema in Texas appear comparable to historical events.
+The ``Great Blizzard'' of February 1899, shown in \cref{fig:historic_bk}, caused even more intense cold.
 
@@
   \item Hourly \SI{2}{\meter} air temperature reanalysis on a \SI{0.25}{\degree} grid from the ERA-5 reanalysis project produced by the European Centre for Medium Range Weather Forecasting \cite{hersbach_era5:2020} and available  from the Copernicus Data Store (\url{https://cds.climate.copernicus.eu}) from 1950 to the present.
+        The period from 1950 to 1979 is released as a preliminary back extension.
         All plots shown in the main text use the ERA-5 data, but supplemental figures use other data sets.
@@
         This gridded product is based on statistical analysis of station data and is available from 1880 to the present.
-This data is considered an experimental product, so we use it only for comparative purposes.
+        This dataset is considered an experimental product, so we use it only for comparative purposes.
   \item To complement blended gridded data products, we use station temperature data from the Global Historical Climatology Network (GHCN) dataset compiled by the National Ocean and Atmospheric Administration \cite{Menne:2012hk} and available at \url{https://www1.ncdc.noaa.gov/pub/data/ghcn/daily/}.
-This dataset provides daily mean, maximum, and minimum temperature.
+        This dataset provides daily mean, maximum, and minimum temperature observations.
         These measurements represent point measurements, which can differ in important ways from gridded products describing spatial averages due to spatial heterogeneity of temperature.
@@
 
-The hourly or daily thermal energy requirement for space heating is primarily driven by how much lower the ambient temperature is from an indoor comfort temperature of \SI{68}{\degree F}.
+The hourly or daily thermal energy requirement for space heating is primarily driven by how much lower the ambient temperature is than an indoor comfort temperature of \SI{68}{\degree F}.
 This relationship is often expressed in terms of heating degree days or hours.
@@
 For each event duration considered, return periods are obtained by fitting a stationary generalized extreme value (GEV) distribution to the time series of annual minimum December-February (DJF) temperatures.
+This is a standard analysis in many engineering textbooks.
 Events that occur in December are coded to the following year.
@@
 
-\section{How extreme was inferred heating demand over the Texas interconnect?}
+\section{How extreme was inferred heating demand over the Texas Interconnect?}
 
 The total shock to Texas heating demand is partially determined by the extent to which cold snaps impact multiple population centers simultaneously.
-As such, understanding whether this cold snap had precedent with respect to its feature of impacting several populated regions of Texas' grid concurrently is critical.
+As such, understanding whether there was precedent for a cold snap simultaneously affecting several regions of Texas's grid that today have high population density is critical.
 We therefore use our measure of inferred demand for heating (see \cref{sec:inferred-demand}) to represent the aggregate heating demand induced by cold temperatures.
@@
 At the two day duration, the 2021 and 1989 events were approximately equally intense and other storms including December 1983 are nearly as intense.
-The 2011 storm, which caused rolling blackouts and motivated research into energy system resilience to cold \cite{ferc_outages:2011}, was quite modest by comparison.
+The 2011 storm, which caused rolling blackouts and motivated research into the energy system's vulnerability to cold \cite{ferc_outages:2011}, was quite modest by comparison.
 The right panel shows statistical return periods for these extreme events, though these estimates are sensitive to small changes in demand for heating.
@@
   \caption{
-    The aggregated demand for heating induced by the February 2021 cold snap is not unprecedented.
+    The inferred demand for heating induced by the February 2021 cold snap is not unprecedented.
     For the worst 6 hours, the 1989 event was more severe than the 2021 event, while they are comparable for the longer durations.
-    (a): time series of spatially averaged annual maximum heating degree days weighted by 2020 population for 6 hour and 2 day durations.
+    (a): time series of annual maximum inferred demand for heating (\cref{sec:inferred-demand}) at 6 hour and 2 day durations.
     December extremes, including the December 1989 storm, are coded to the following year so that one maximum per December-February winter season is taken.
@@
 Results are similar using station data (\cref{fig:local_ghcnd}).
-Importantly for the energy system, the band experiencing cold with return period greater than 50 years includes a substantial fraction of Texas’s population (\cref{fig:local_ghcnd}a) and natural gas generation (\cref{fig:local_ghcnd}d).
-Outside the Texas Interconnect region, much of the Texas Panhandle and central Oklahoma experienced intense cold with a return period greater than 100 years.
-Although Oklahoma recorded some water main breaks \cite{crum_water:2021} and rolling blackouts for several hours \cite{money_oklahoma:2021}, infrastructure impacts were less severe than in Texas despite recording more extreme (relative to its own historical record) temperatures.
+Importantly for the energy system, the band experiencing cold with return period greater than 50 years includes a substantial fraction of Texas’s population (\cref{fig:local_era5}a) and natural gas generation (\cref{fig:local_era5}d).
+Outside the Texas Interconnect region, much of the Texas Panhandle and central Oklahoma experienced intense cold with a return period greater than 100 years at multiple durations.
+Although Oklahoma recorded some water main breaks \cite{crum_water:2021} and rolling blackouts for several hours \cite{money_oklahoma:2021}, infrastructure impacts were less severe than in Texas despite recording more extreme (relative both to its own historical record and in absolute terms) temperatures.
 
@@
 \Cref{fig:local_era5,fig:local_ghcnd} show that this happened even though most parts of the state had previously experienced similarly intense cold, notably in 1989.
-While our analysis neglects other meteorological factors, like freezing rain, that may have impeded operations of specific facilities, we find that the electricity and energy infrastructure in the Texas Interconnect were collectively vulnerable to the temperatures observed in 1989.
-Similarly, water mains broke in places like Houston where temperatures did not exceed 100-year return levels, underscoring deep vulnerability of critical infrastructure systems \cite{chester_reliable:2020}.
+While our analysis neglects other meteorological factors, like freezing rain, that may have impeded operation of specific facilities, we find that the electricity and energy infrastructure in the Texas Interconnect was collectively vulnerable to the temperatures observed in 1989.
+Similarly, water mains broke in places like Houston where temperatures did not exceed 100-year return levels, underscoring characteristic vulnerability of critical infrastructure systems \cite{chester_reliable:2020}.
 
 Another cause of load shedding was the high demand for electricity that low temperatures induced.
-Currently over 50\% of Texas households use electricity for space heating over the majority of census tracts of the state \cite{waite_heating:2020,white_txresidential:2019} and further electrification is a central element of many plans to decarbonize our energy sector \cite{williams_decarbonization:2012,davis:2018} in an effort to mitigate the climate risks associated with greenhouse gas emissions.
+Currently over 50\% of Texas households use electricity for space heating over the majority of census tracts of the state \cite{waite_heating:2020,white_txresidential:2019} and further electrification is a central element of many plans to decarbonize our energy sector \cite{williams_decarbonization:2012,davis:2018}.
 While summer peak loads have been a central planning concern on the Texas grid in the past, it is likely that winter peak loads will become a greater concern in the coming decades.
 In fact, the peak demand during this event was estimated at around \SI{74.5}{\giga\watt}, just shy of the all time peak for any season of \SI{74.8}{\giga\watt} in August 2019 \cite{everhart_iea:2021}.
-As electrification of heating increases, severe cold snaps may represent the annual peak demand on the Texas Interconnect.
+As electrification of heating increases, severe cold snaps may represent the peak demand on the Texas Interconnect.
 
@@
 More fundamentally, an assessment of exposure to cold extremes over the next decades should consider the deeply uncertain distribution of future climate change, and the induced effect on cold extremes in Texas.
-In general, the IPCC anticipates that global warming will lead to more hot extremes and fewer cold extremes \cite{ipcc_ar5:2014}, but the effects of Arctic amplification on mid-latitude regional temperature extremes remains an area of active research \cite{Barnes:2013fp,Cohen:2014gx,Screen:2013ho}.
-However, the effect of climate change on peak demand for heating is likely to be small compared to the effect of rapid population growth (for example, the Texas Water Development Board \cite{texaswaterplan:2012} anticipates at least 40\% growth from 2020-2050).
+Although a broad scientific consensus suggests the frequency of cold extremes should decrease under warming in most places \cite{ipcc_ar5:2014}, possible links between North-South temperature gradients and mid-latitude cold extremes remains an area of active research \cite{Barnes:2013fp,Cohen:2014gx,Screen:2013ho}.
+Regardless, the effect of climate change on peak demand for heating is likely to be small compared to the effect of rapid population growth (for example, the Texas Water Development Board \cite{texaswaterplan:2012} anticipates at least 40\% growth from 2020-2050).
 
@@
 The February 2021 cold snap was the most intense in 30 years, but was not without precedent in the full historical record.
-In addition to the record cold conditions of 1899 (\cref{fig:historic_bk}), we estimate that the weather of December 1989 would have resulted in slightly higher 6-hour and 2-day aggregate heating demands over the Texas Interconnect compared to February 2021.
-Several other storms since 1950 were nearly as severe.
-Given upward trends in the electrification of heating, it is likely that future cold snaps will cause peak annual loads on the Texas interconnect to occur during the winter season.
-Infrastructure expansion necessitated by a rapidly growing population offers Texas the opportunity to develop a more resilient energy system.
+In addition to the record cold conditions of 1899 (\cref{fig:historic_bk}), we estimate that the weather of December 1989 would have resulted in slightly higher 6-hour and 2-day aggregate heating demands over the Texas Interconnect had it occurred in February 2021.
+Several other storms since 1950 would have produced nearly as much demand for heating.
+Given upward trends in the electrification of heating, it is likely that future cold snaps will cause peak annual loads on the Texas Interconnect to occur during the winter season.
+Infrastructure expansion necessitated by a rapidly growing population offers Texas the opportunity to invest in a more resilient energy system.
 
@@
 DJF was supported by a gift from Gates Ventures LLC to the Carnegie Institution for Science.
-The authors thank Adam Massmann, Elisabeth Gawthrop, and Vivek Srikrishnan for insightful comments on a preliminary draft of this manuscript.
-We also thank the developers and maintainers of the open source packages we used, in particular the pydata and pangeo communities, and the many journalists and academics whose timely reporting and commentary has informed our thinking.
+The authors thank
+Sylvia Dee,
+Elisabeth Gawthrop,
+Adam Massmann,
+and
+Vivek Srikrishnan
+for comments on a preliminary draft of this manuscript.
+The authors also thank the developers and maintainers of the open source packages we used, in particular the pydata and pangeo communities, and the many journalists and academics whose timely reporting and commentary has informed our thinking.
 
@@
 \clearpage
+\appendix
 \section{Supplemental Results}
@@
 Next, \cref{fig:historic_tx} shows the same data as \cref{fig:historic_era5} but zooms in on Texas.
-The 1989 (g-i) and 2021 (m-o) appear to be the most severe events in this data set, but the 1-day cold extremes in the 1989 event are more intense than in the February 2021 event.
+The 1989 (g-i) and 2021 (m-o) appear to be the most severe events in this data set, and the 1-day cold extremes in the 1989 event are more intense than in the February 2021 event, consistent with other results.
 
@@
 
-To complement our analysis of inferred demand for heating, we consider how results change when two variables are changed.
+To complement our analysis of inferred demand for heating, we consider how results change as a function of two modeling decisions.
 First, we consider what happens if the spatial field demand for heating is aggregated using grid cell area rather than population density.
@@
 To provide some context for our inferred demand for heating metric, \cref{fig:hdd_ts} plots its time series during the peak of the February 2021 cold snap.
+This reveals a rise from approximately \SI{10}{\degree F} to nearly \SI{60}{\degree F} during the peak of the February 2021 cold snap.
 

### 2021-02-28 doc/coverletter.tex
DJF edits

@@
 
-	I am writing to submit our manuscript entitled ```how unprecedented was the February 2021 Texas cold snap?'' for consideration to Environmental Research Letters.
-
+	I am writing to submit our manuscript entitled ``How unprecedented was the February 2021 Texas cold snap?'' for consideration for publication in Environmental Research Letters.
+	​
 	In light of the recent cold snap that caused cascading infrastructure failures in Texas, we sought to understand the extent to which the temperatures were unprecedented in the historical record.
@@
 	As the Texas state legislature and other regulatory bodies investigate the causes of this event with an eye to preventing future disasters, a credible assessment of the degree to which these temperatures could have been anticipated offers a significant and relevant contribution to public and scientific discourse.
-
+	​
 	Our findings also have implications for the broader literature around electrification and decarbonization.
 	Deep decarbonization is critical to avoiding severe climate change scenarios.
-	\citet{williams_decarbonization:2012} argue, as others have done, that widespread electrification is needed to achieve decarbonization
+	\citet{williams_decarbonization:2012} argue, as others have done, that widespread electrification is needed to achieve decarbonization.
 	More recently, \citet{waite_heating:2020} quantified the load effects of electrifying U.S. residential and commercial space heating using ten years of weather data.
-	Their analysis indicated that most regions of Texas could expand up to 70 to 100\% electric heating without impacting local peak annual load, which generally occurs during the summer cooling season.
-	This recent cold snap indicates that Texas may be closer to experiencing peak electricity loads during the winter than we had previously anticipated.
-	Our results suggest that electrification of building heating could increase peak electricity demands, requiring some combination of increased generation, storage, and/or long-distance transmission to achieve the high reliability levels that are typically considered acceptable.
+	Their analysis indicated that most regions of Texas could expand to up to 70 to 100\% electric heating without impacting local peak annual load, which generally occurs during the summer cooling season.
+	However, this recent cold snap indicates that Texas may be closer to experiencing peak electricity loads during the winter than we had previously anticipated.
+	Our results suggest that even small increases in space heating electrification in Texas could increase peak electricity demands, requiring some combination of increased generation, storage, and/or long-distance transmission.
 

### 2021-02-28 doc/article.tex
DJF & JDG last edits

@@
 
-\title[PREPRINT: Doss-Gollin, Farnham, Lall, and Modi]{How unprecedented was the February 2021 Texas cold snap?}
+%\title[PREPRINT: Doss-Gollin, Farnham, Lall, and Modi]{How unprecedented was the February 2021 Texas cold snap?}
+\title{How unprecedented was the February 2021 Texas cold snap?}
 
@@
   Winter storm Uri brought severe cold to the southern United States in February 2021, causing a cascading failure of interdependent systems in Texas where infrastructure was not adequately prepared for such cold.
-  In particular, the failure of interconnected energy systems limited electricity supply just as heating demands spiked, leaving millions of Texans without heat or electricity, many for several days.
+  In particular, the failure of interconnected energy systems reduced electricity supply just as heating demands spiked, leaving millions of Texans without heat or electricity, many for several days.
   This motivates the question: was the cold that contributed to this infrastructure failure a ``black swan'' that could not have been anticipated, or did historical storms provide a precedent?
-  We compute the population weighted temperature excursion below \SI{68}{\degree F} as a proxy for heating demand and use this inferred demand for heating to explore how rare the temperatures recorded during the February 2021 cold snap were using historical data.
-  We find that temperatures during a storm in December 1989 were more intense than those recorded during February 2021, and that several other storms were nearly as intense.
-  Although some specific locations experienced record cold, temperatures at most locations were cold but not unprecedented.
+  We compute the population weighted temperature excursion below \SI{68}{\degree F} as a proxy for heating demand and use this metric to answer the question ``what would the aggregate demand for heating have been had historic cold snaps occurred today?''.
+  We find that local temperatures and the inferred demand for heating across the Texas Interconnect during a storm in December 1989 were more intense than those recorded during February 2021, and that several other storms in the modern era were comparable.
   Given anticipated population growth, future storms may lead to even greater infrastructure failures if adaptive investments are not made.
@@
 
-%\submitto{\ERL}
+\submitto{\ERL}
 \maketitle
@@
 Between February 14th and 17th, 2021, a northern air mass blanketed much of the continental United States, causing anomalously low surface temperatures across the Great Plains.
-While polar air excursions are not unusual, they do not typically reach as far South as the Mexican border.
-The state of Texas was particularly hard hit, with coincident and cascading failures of natural gas production, power generation, transportation, and water systems leaving millions of Texans without electricity, heat, and water for several days.
+While polar air excursions are not unusual, they do not typically reach as far south as the Mexican border.
+The state of Texas was particularly hard hit, with coincident and cascading failures of natural gas production, power generation, transportation, and water systems leaving millions of Texans without electricity, heat, and water, many for several days.
 The intensity and widespread nature of this weather event underscore a high vulnerability to cold temperatures in places that did not experience a comparable event in preceding decades.
@@
   \caption{
-    Severe cold snaps that affect Texas and extend into the Central United States have several precedents in the historical record.
+    Severe cold snaps that affect Texas and extend into the central United States have several precedents in the historical record.
     Plot shows anomalies of daily mean temperatures for historic major cold events affecting Texas, defined as the departure from the seasonal (December-February) mean to facilitate comparison of events in different calendar months.
@@
 
-We use three temperature datasets to analyze the recent and historical cold snap in Texas through the lens of distributed (each grid cell analyzed separately) and aggregated (weighted averages taken across space) extreme values analysis.
+We use three distinct datasets to analyze temperature minima in Texas through the lens of distributed (each grid cell analyzed separately) and aggregated (weighted averages taken across space) extreme values analysis.
 
@@
   \item Daily mean, minimum and maximum temperatures, gridded to \SI{1}{\degree}, produced by Berkeley Earth (\url{http://berkeleyearth.org/data/}).
-        This gridded product is based on statistical analysis of station data and is available from 1880 to the present.
+        This gridded product is based on statistical analysis of station data and is available from 1880 to 2019.
         This dataset is considered an experimental product, so we use it only for comparative purposes.
@@
         This dataset provides daily mean, maximum, and minimum temperature observations.
-        These measurements represent point measurements, which can differ in important ways from gridded products describing spatial averages due to spatial heterogeneity of temperature.
+        These measurements represent point measurements, which can differ in important ways from gridded products describing spatial averages due to the spatial heterogeneity of temperature fields.
         We retain stations within the state of Texas if they provide at least 60 years of data and if they contain observations for the set of historical cold extremes shown in \cref{fig:historic_era5}.
@@
 To assess how spatially correlated cold spells might affect the Texas electric grid, we average heating demand in space over the Texas Interconnect domain \cite{useia_regions:2021}, weighting each grid cell by 2020 population density \cite{ciesin_gpwv4:2016}.
-We refer to this spatially aggregated metric, which has the straightforward interpretation as the average temperature below \SI{68}{\degree F} experienced by Texas residents, as ``Inferred Demand for Heating.''
+We refer to this spatially aggregated time series, which has the straightforward interpretation as the average heating demand experienced by Texas residents, as ``Inferred Demand for Heating.''
 
@@
 
-For each event duration considered, return periods are obtained by fitting a stationary generalized extreme value (GEV) distribution to the time series of annual minimum December-February (DJF) temperatures.
-This is a standard analysis in many engineering textbooks.
+For each event duration considered, return periods are obtained by fitting a stationary generalized extreme value (GEV) distribution to the time series of annual minimum December-February (DJF) temperatures, or to the time series of annual maxima of inferred demand for heating (\cref{sec:inferred-demand}).
 Events that occur in December are coded to the following year.
-The 2021 winter season was excluded from return period estimates, allowing interpretation of return periods as \emph{a priori} estimates.
+The 2021 winter season was excluded from return period estimates, allowing us to interpret return periods for the February 2021 event as \emph{a priori} estimates.
 
@@
 The effect of cold temperatures on energy demand and critical infrastructure depends on how long the cold persists.
-Short duration cold snaps can kill plants, freeze exposed pipes, and contribute to dangerous roadway conditions.
+Short duration cold snaps can kill plants, freeze exposed pipes, freeze wind turbines, and contribute to dangerous roadway conditions.
 Longer duration cold spells contribute to demand for heating and energy and cause pipes to burst even if they have some insulation.
@@
 \Cref{fig:idf_weighted} shows that the intensity, duration, and recurrence intervals of the February 2021 storm are severe but not unprecedented in the historical record.
-For example, at the 6 hour duration the December 1989 storm was substantially more intense and other storms including February 1951 are nearly as intense.
-At the two day duration, the 2021 and 1989 events were approximately equally intense and other storms including December 1983 are nearly as intense.
+For example, at the 6 hour duration the December 1989 storm was substantially more intense and other storms including February 1951 were nearly as intense.
+At the two day duration, the 2021 and 1989 events were approximately equally intense and other storms including December 1983 were nearly as intense.
 The 2011 storm, which caused rolling blackouts and motivated research into the energy system's vulnerability to cold \cite{ferc_outages:2011}, was quite modest by comparison.
-The right panel shows statistical return periods for these extreme events, though these estimates are sensitive to small changes in demand for heating.
+The right panel shows statistical return periods for these extreme events, though these estimates are sensitive to small changes in inferred demand for heating.
 
@@
 Water treatment and distribution systems, as well as other essential services, also rely on electricity, further increasing vulnerabilities.
-Instead of aggregating this risk in space, we estimate the exceedance probability of the February 2021 temperatures at each grid cell separately to shed light on the severity of cold experienced by specific installations across the region.
+Instead of aggregating this risk in space, we estimate the exceedance probability of the February 2021 temperatures at each grid cell separately to shed light on the severity of cold experienced by installations across the region.
 
@@
     (d): energy generation facilities in Texas \cite{useia_generators:2021}.
-    (b,c,e,f): local return periods for 6 hour, 24 hour, 48 hour, and 96 hour cold durations, respectively.
+    (b,c,e,f): local return periods for 6 hour, 1 day, 2 day, and 4 day durations, respectively.
     Contours enclose regions that recorded 50 and 100 year return levels.
-    Gray region with dotted black line in (a) and (d) shows boundaries of the Texas Interconnect \cite{useia_regions:2021}.
+    The gray region in panels (a) and (d) shows boundaries of the Texas Interconnect \cite{useia_regions:2021}.
   }\label{fig:local_era5}
@@
 
-Our spatially aggregated metric of inferred demand for heating shows that the February 2021 event was significant but not unforeseeable (\cref{fig:idf_weighted}).
+Our spatially aggregated metric of inferred demand for heating shows that the February 2021 event was intense but not unforeseeable (\cref{fig:idf_weighted}).
 Although specific locations experienced very intense ($>100$ year return period) temperatures, we find that for most locations in Texas the temperatures recorded during the February 2021 cold snap had precedent in the historical record and could have been anticipated given available data.
@@
 \Cref{fig:local_era5,fig:local_ghcnd} show that this happened even though most parts of the state had previously experienced similarly intense cold, notably in 1989.
-While our analysis neglects other meteorological factors, like freezing rain, that may have impeded operation of specific facilities, we find that the electricity and energy infrastructure in the Texas Interconnect was collectively vulnerable to the temperatures observed in 1989.
+While our analysis neglects other meteorological factors, like freezing rain, that may have impeded operations at specific facilities, our analysis suggests that the February 2021 failures of energy and electricity systems in the Texas Interconnect took place during temperatures comparable to those previously recorded.
 Similarly, water mains broke in places like Houston where temperatures did not exceed 100-year return levels, underscoring characteristic vulnerability of critical infrastructure systems \cite{chester_reliable:2020}.
@@
 Another cause of load shedding was the high demand for electricity that low temperatures induced.
-Currently over 50\% of Texas households use electricity for space heating over the majority of census tracts of the state \cite{waite_heating:2020,white_txresidential:2019} and further electrification is a central element of many plans to decarbonize our energy sector \cite{williams_decarbonization:2012,davis:2018}.
+Currently over 50\% of Texas households use electricity for space heating over the majority of census tracts of the state \cite{waite_heating:2020} and further electrification is a central element of many plans to decarbonize our energy sector \cite{williams_decarbonization:2012,davis:2018,white_txresidential:2019}.
 While summer peak loads have been a central planning concern on the Texas grid in the past, it is likely that winter peak loads will become a greater concern in the coming decades.
@@
 More fundamentally, an assessment of exposure to cold extremes over the next decades should consider the deeply uncertain distribution of future climate change, and the induced effect on cold extremes in Texas.
-Although a broad scientific consensus suggests the frequency of cold extremes should decrease under warming in most places \cite{ipcc_ar5:2014}, possible links between North-South temperature gradients and mid-latitude cold extremes remains an area of active research \cite{Barnes:2013fp,Cohen:2014gx,Screen:2013ho}.
-Regardless, the effect of climate change on peak demand for heating is likely to be small compared to the effect of rapid population growth (for example, the Texas Water Development Board \cite{texaswaterplan:2012} anticipates at least 40\% growth from 2020-2050).
+Although a broad scientific consensus suggests the frequency of cold extremes should decrease under warming in most places \cite{ipcc_ar5:2014}, possible links between North-South temperature gradients and mid-latitude temperature extremes remains an area of active research \cite{Barnes:2013fp,Cohen:2014gx,Screen:2013ho}.
+Regardless, the effect of climate change on peak demand for heating is likely to be small compared to the effect of rapid population growth; for example, the Texas Water Development Board anticipates at least 40\% growth from 2020 to 2050 \cite{texaswaterplan:2012}.
 
@@
 Next, \cref{fig:historic_tx} shows the same data as \cref{fig:historic_era5} but zooms in on Texas.
-The 1989 (g-i) and 2021 (m-o) appear to be the most severe events in this data set, and the 1-day cold extremes in the 1989 event are more intense than in the February 2021 event, consistent with other results.
+The 1989 (g-i) and 2021 (m-o) appear to be the most severe events in this data set, and the 1-day cold extremes in the 1989 event are more intense than in the February 2021 event, consistent with results in the main text.
 
@@
 Only stations with at least 60 years of data are considered, and since the locations of these stations are not chosen at random, this does not constitute a representative sample of all points across Texas.
-However, the spatial pattern matches that of \cref{fig:local_era5}, with a band of severe cold stretching from South Central to East Texas and in the Texas Panhandle.
+However, the spatial pattern matches that of \cref{fig:local_era5}, with a band of severe cold stretching from south-central to eastern Texas and in the Texas Panhandle.
 
@@
   \caption{
-    A time series of population weighted temperature excursion from \SI{68}{\degree F} over the Texas Interconnect during the most
+    A time series of inferred demand for heating over the Texas Interconnect during the February 2021 cold snap.
   }\label{fig:hdd_ts}

### 2021-02-28 doc/article.tex
Add zenodo DOI

@@
 We are committed to open science.
-Our open source code is freely available in a live repository at \url{https://github.com/jdossgollin/2021-TXtreme} and in an archived repository at TODO.
+Our open source code is freely available in a live repository at \url{https://github.com/jdossgollin/2021-TXtreme} and in an archived repository at \url{https://dx.doi.org/10.5281/zenodo.4568923}.
 Reproducibility is facilitated through the Snakemake library \cite{koster_snakemake:2012}.

### 2021-02-28 doc/article.tex
Improve README

@@
 
+\vspace*{1in}
+
+{\LARGE How unprecedented was the February 2021 Texas cold snap?}
+
+\vspace*{1in}
+
+This is a non-peer reviewed preprint submitted to EarthArXiv by James Doss-Gollin, David J. Farnham, Upmanu Lall, and Vijay Modi.
+This manuscript has been submitted to \emph{Environmental Research Letters}.
+If accepted, a link to the published DOI will be provided here.
+
+\clearpage
+
 %\title[PREPRINT: Doss-Gollin, Farnham, Lall, and Modi]{How unprecedented was the February 2021 Texas cold snap?}
-\title{How unprecedented was the February 2021 Texas cold snap?}
+\title[PREPRINT by Doss-Gollin, Farnham, Modi, and Lall]{How unprecedented was the February 2021 Texas cold snap?}
 

### 2021-03-23 doc/article.tex
EarthArXiV text

@@
 
-This is a non-peer reviewed preprint submitted to EarthArXiv by James Doss-Gollin, David J. Farnham, Upmanu Lall, and Vijay Modi.
-This manuscript has been submitted to \emph{Environmental Research Letters}.
-If accepted, a link to the published DOI will be provided here.
+This is a non-peer reviewed manuscript by James Doss-Gollin, David J. Farnham, Upmanu Lall, and Vijay Modi.
+You can find a permanent and updated preprint on EarthArXiV at \url{https://doi.org/10.31223/X5003J}.
+This manuscript is still under review; if accepted we will provide a link to the published article here.
 

### 2021-04-23 doc/review_response.tex
work on response to reviewers

\documentclass{ar2rc}

\usepackage{url}
\usepackage{siunitx}

\usepackage[colorinlistoftodos]{todonotes}
\usepackage{color}
\newcommand{\jdg}[1]{\todo[color=red!30]{#1}}
\newcommand{\all}[1]{\todo[color=blue!30]{#1}}

\hypersetup{hidelinks}

\title{How unprecedented was the February 2021 Texas cold snap?}
\author{James Doss-Gollin, David J. Farnham, Upmanu Lall, and Vijay Modi}
\journal{Environmental Research Letters}

\begin{document}

\maketitle

\listoftodos

\section{Referee \#1}

\RC  This is an excellent, timely analysis. The authors are right to point to the need for more in depth studies on how cascading failures occurred during the 2021 Texas freeze, and the steps required to minimize their impacts in the future. But this is a much needed, early contribution, and very well done. I don't have any major criticisms of the work, which is well organized and rigorous.

\AR Thank you for your comments. We agree that this points to a need for more in depth studies, particularly related to cascading failures of natural gas and electricity systems but also to failure of water and other systems.

\subsection{Some minor comments:}

\RC Considering I think this paper will be widely read and cited, one general comment is that I felt like the citations were a little light in the introduction, especially regarding the 2021 freeze itself.

\AR This is a point that reviewer \#2 has also brought up.\all{add some citations}

\RC `Texas interconnect' (which I assume is the Texas Interconnection bulk electric power system administered by ERCOT) is used several times but (I don't think) ever defined; many folks will not know what this refers to. If you added a brief definition it would help. If this does refer to the grid, then an additional comment is that I've never heard it referred to as the Interconnect, but rather the Interconnection.

\AR This is a helpful comment. The Interconnection is a better term and we have revised it throughout.\jdg{Change language to always say TX Interconnection}\jdg{define Texas Interconnection}. This will clarify the document and facilitate understanding.

\RC Regarding the definition of a HDD/CDD: More frequently 65 degrees is used, I think.

\AR Do we need to do any analysis, or just reply that it doesn't matter?\all{Assess}

\RC On the difficulty of estimating supply side risk -- is it possible at this point to map the locations of plants that lost functionality?

\AR This is a helpful suggestion. When we submitted the manuscript, this data was not available. However, ERCOT has since released data on generation failure.\jdg{Download this data and update the map}

\RC In the discussion, you mention the 1983 event as being one of comparable severity, and then talk about the grid's failure in 2021. I don't know if this was your intention, but  when I read this sentence the first time I took it to mean the grid is *more* vulnerable now. I think you could reasonably make that case with greater electrification of heating. But I also wondered about coal-to-gas switching, and whether 1983 would have likely been a different fleet of power plants that were perhaps less likely to lose useable capacity during cold weather.

\AR Thank you for the interesting and thought-provoking comment.\all{Discuss: do we have thoughts?}

\section{Referee \#2}

\RC This paper investigates whether the Texas deep freeze event was a black swan event, or if it should have been predicted from previous cold snaps. The authors compute the population weighted temperature excursion below \SI{68}{\degree F} as a proxy for heating demand and use that to determine aggregate heating demand for 2021. Overall the authors find that demand for previous storms was more intense than the 2021 disaster.

\AR Thank you for this helpful summary.

\RC The introduction is seriously lacking in citations and references. For example, this line ``While polar air excursions are not unusual, they do not typically reach as far south as the Mexican border.'' And a number of others need citations and would benefit from concrete numbers. A question that arises is how many storms have reached the Mexican border in the last 10 years? Line 45-46 on page 1 also needs a citation.

\AR Thank you for this comment. We note that referee \#1 made a similar comment. We have now added refererences\ldots\all{Add more references as noted above}

\RC In the introduction the ``population weighted temperature excursion below \SI{68}{\degree F}'' needs a definition for temperature excursion. Figure 1 is missing citations for the data. What do you mean by days ending? Were these temperatures recorded at a specific time of day? Figure 1 would be improved by adding column headers for the averages (1-day, 3 day, and 5 day).

\AR Thank you for calling to our attention these unclear points. First, we have changed the subplot titles of Figure 1 from ``N days ending YYYY-MM-DD'' to ``YYYY-MM-DD to YYYY-MM-DD'' for clarity. We have supplemented this with column titles. We have also added a reference to the ERA-5 data in the caption.\jdg{update the figure as described}

\RC A weakness of this paper is the lack of comment about the capacity and size of the Texas grid. How does the inferred demand compare to the size of the ERCOT power grid and the interconnection capability over time? We know that the ERCOT grid is larger in 2021 than in 1989, but from Figure 2 there should have been more demand in 1989 than in 2021. Would most of the people have wood stoves? Was the blackout and number of people impacted as large in the past, as it was in 2021?

\AR Thank you for bringing this point to our attention. To minimize confusion, we have clarified our terminology so that ``inferred heating demand'' is not ``inferred heating demand per capita.'' This is reflected throughout the document.\jdg{Update language throughout} However, comparing the actual infrastructure failures across different events (such as 1989) would require understanding and modeling cascading failures of interdependent infrastructure systems, which is beyond the scope of our analysis.

\RC On page 7 line 51 the authors say the Oklahoma recorded some water main breaks and rolling blackouts, but infrastructure impacts were less severe. To strengthen this paper, the authors should provide insights as to why this was the case for OK, and not TX.

\AR Perhaps ``it gets cold there'' but I'd hate to say things without any evidence?\all{Comment}

\RC I understand that a big takeaway from the paper is that the cold temperatures and heat demand could be predicted since there were similar demand numbers from 1989, but one year of similar demand does not eliminate this as a black swan event. From the discussion period it seems that over a 30-year time period this type of event has only happened twice. That seems very rare in my opinion.

\AR Thank you for this comment. Whether it was a ``black swan'' or not depends on a definition of black swan. We have provided a definition and citation for this term.\jdg{Cite Taleb and add definition} Beyond this one instance, we have sought to clarify our language to ask only whether the temperatures were ``unprecedented.'' Since we find that, particularly at the aggregate scale, there were historical events of comparable intensity and duration, we can state that the event ``was not unprecedented.'' We also attempt to quantify return periods. However, we have removed references to terms like ``rare'' or ``surprising'' \jdg{check for these terms} that do not have specific interpretations to minimize any confusion.

\RC The conclusions says that several other storms would have resulted in similar demand. How many?

\AR Thank you for pushing us to be more specific with our claims. The text now reads (\ldots).

\RC This paper needs to tie in the size and capacity of the Texas grid to the demand projections. Without it I believe the claims are overstated.

\AR Thank you for this comment. The first step we have taken is, as noted above, to refer to ``inferred demand for heating per capita'' to emphasize that we are not accounting for population change.\all{anything else we should include?}

\end{document}

### 2021-04-23 doc/submission_02.tex
address some reviewer comments -- discussion still needed for others

--- 2021-02-28 doc/article.tex
@@
 
-%\title[PREPRINT: Doss-Gollin, Farnham, Lall, and Modi]{How unprecedented was the February 2021 Texas cold snap?}
 \title{How unprecedented was the February 2021 Texas cold snap?}
@@
   In particular, the failure of interconnected energy systems reduced electricity supply just as heating demands spiked, leaving millions of Texans without heat or electricity, many for several days.
-  This motivates the question: was the cold that contributed to this infrastructure failure a ``black swan'' that could not have been anticipated, or did historical storms provide a precedent?
+  This motivates the question: were the temperatures that contributed to this infrastructure failure beyond what was known to be possible, or did historical storms offer a precedent?
   We compute the population weighted temperature excursion below \SI{68}{\degree F} as a proxy for heating demand and use this metric to answer the question ``what would the aggregate demand for heating have been had historic cold snaps occurred today?''.
-  We find that local temperatures and the inferred demand for heating across the Texas Interconnect during a storm in December 1989 were more intense than those recorded during February 2021, and that several other storms in the modern era were comparable.
+  We find that local temperatures and the inferred demand for heating across the region served by the Texas Interconnection during a storm in December 1989 were more severe than those recorded during February 2021, and that several other storms in the modern era were comparable.
   Given anticipated population growth, future storms may lead to even greater infrastructure failures if adaptive investments are not made.
-  Further, electricity system managers should anticipate that upward trends in electrification of heating may cause peak annual loads on the Texas Interconnect to occur during winter storms.
+  Further, electricity system managers should anticipate that upward trends in electrification of heating may cause peak annual loads on the Texas Interconnection to occur during winter storms.
 
@@
 Since production and distribution of natural gas and electricity is possible under conditions far colder than any Texas experienced in February 2021, energy system failures reflect inadequate preparedness for cold.
-If this were an unprecedented or very rare event then it is possible that the impacts resulted from an event outside the design envelope.
+If these temperatures were unprecedented in the region covered by the Texas Interconnection then it is possible that the impacts resulted from an event outside the design envelope.
 On the other hand, if similar events were in the record then one would expect a higher level of preparation.
-It is therefore important to assess whether the February 2021 cold snap could have been anticipated.
-To answer this question, we first compute the population weighted temperature excursion below \SI{68}{\degree F} as a proxy for the unknown heating demand, and use this metric to explore how rare this event was using historical data.
+It is therefore important to assess whether historical data offered a precedent for the temperatures observed during February 2021.
+To answer this question, we first compute the population weighted temperature excursion below \SI{68}{\degree F} as a proxy for the unknown heating demand, and use this metric to explore the frequency with which this event could have been expected to occur using historical data.
 Next, we conduct a spatially distributed analysis to assess how unexpected the cold experienced by local roads, water mains, gas pipelines, energy generation facilities, and critical infrastructure installations was across Texas.
@@
 
-We use three distinct datasets to analyze temperature minima in Texas through the lens of distributed (each grid cell analyzed separately) and aggregated (weighted averages taken across space) extreme values analysis.
+We use three distinct datasets to analyze temperature minima in the region covered by the Texas Interconnection bulk electric power system administered by the Electric Reliability Council of Texas (hence ``Texas Interconnection'') through the lens of distributed (each grid cell analyzed separately) and aggregated (weighted averages taken across space) extreme values analysis.
 
@@
 \end{enumerate}
-We also use population density data from the GPWv4 dataset \cite{ciesin_gpwv4:2016}, a list of power generation facilities from the US Energy Information Administration \cite{useia_generators:2021}, and a map of the Texas Interconnect \cite{useia_regions:2021}.
+We also use population density data from the GPWv4 dataset \cite{ciesin_gpwv4:2016}, a list of power generation facilities from the US Energy Information Administration \cite{useia_generators:2021}, and a map of the Texas Interconnection \cite{useia_regions:2021}.%TODO: add and cite information on the generators that did and did not go out.
 
-\subsection{Inferred Demand for Heating}\label{sec:inferred-demand}
+\subsection{Inferred per Capita Demand for Heating}\label{sec:inferred-demand}
 
-Most space heating in Texas is either electric or gas \cite{waite_heating:2020} and the majority of power generation in the Texas Interconnect depends on natural gas \cite{everhart_iea:2021}.
+Most space heating in Texas is either electric or gas \cite{waite_heating:2020} and the majority of power generation in the Texas Interconnection depends on natural gas \cite{everhart_iea:2021}.
 Stress on natural gas production and delivery was therefore just as important as the more visible stress on the electric system.
@@
 
-To assess how spatially correlated cold spells might affect the Texas electric grid, we average heating demand in space over the Texas Interconnect domain \cite{useia_regions:2021}, weighting each grid cell by 2020 population density \cite{ciesin_gpwv4:2016}.
-We refer to this spatially aggregated time series, which has the straightforward interpretation as the average heating demand experienced by Texas residents, as ``Inferred Demand for Heating.''
+To assess how spatially correlated cold spells might affect the Texas electric grid, we average heating demand in space over the Texas Interconnection domain \cite{useia_regions:2021}, weighting each grid cell by 2020 population density \cite{ciesin_gpwv4:2016}.
+We refer to this spatially aggregated time series, which has the straightforward interpretation as the average heating demand experienced by Texas residents, as ``Inferred per Capita Demand for Heating.''
 
@@
 
-For each event duration considered, return periods are obtained by fitting a stationary generalized extreme value (GEV) distribution to the time series of annual minimum December-February (DJF) temperatures, or to the time series of annual maxima of inferred demand for heating (\cref{sec:inferred-demand}).
+For each event duration considered, return periods are obtained by fitting a stationary generalized extreme value (GEV) distribution to the time series of annual minimum December-February (DJF) temperatures, or to the time series of annual maxima of inferred per capita demand for heating (\cref{sec:inferred-demand}).
 Events that occur in December are coded to the following year.
@@
 
-\section{How extreme was inferred heating demand over the Texas Interconnect?}
+\section{How extreme was inferred per capita heating demand over the Texas Interconnect?}
 
@@
 As such, understanding whether there was precedent for a cold snap simultaneously affecting several regions of Texas's grid that today have high population density is critical.
-We therefore use our measure of inferred demand for heating (see \cref{sec:inferred-demand}) to represent the aggregate heating demand induced by cold temperatures.
+We therefore use our measure of inferred per capita demand for heating (see \cref{sec:inferred-demand}) to represent the aggregate heating demand induced by cold temperatures.
 Aggregating historic temperature fields in space using the 2020 population, we answer the question ``what would the aggregate demand for heating have been had historic cold snaps occurred today?''
@@
 The 2011 storm, which caused rolling blackouts and motivated research into the energy system's vulnerability to cold \cite{ferc_outages:2011}, was quite modest by comparison.
-The right panel shows statistical return periods for these extreme events, though these estimates are sensitive to small changes in inferred demand for heating.
+The right panel shows statistical return periods for these extreme events.
 
@@
   \caption{
-    The inferred demand for heating induced by the February 2021 cold snap is not unprecedented.
+    The inferred per capita demand for heating induced by the February 2021 cold snap is not unprecedented.
     For the worst 6 hours, the 1989 event was more severe than the 2021 event, while they are comparable for the longer durations.
-    (a): time series of annual maximum inferred demand for heating (\cref{sec:inferred-demand}) at 6 hour and 2 day durations.
+    (a): time series of annual maximum inferred per capita demand for heating (\cref{sec:inferred-demand}) at 6 hour and 2 day durations.
     December extremes, including the December 1989 storm, are coded to the following year so that one maximum per December-February winter season is taken.
@@
 \Cref{fig:local_era5} shows local return periods for the February 2021 cold snap at 6 hour, 1 day, 2 day, and 4 day durations.
-Other than a band from south-central to south-east Texas, nearly all regions of the Texas Interconnect (gray outline in \cref{fig:local_era5}a,d) experienced cold with a return period below 50 years.
+Other than a band from south-central to south-east Texas, nearly all regions of the Texas Interconnection (gray outline in \cref{fig:local_era5}a,d) experienced cold with a return period below 50 years.
 Results are similar using station data (\cref{fig:local_ghcnd}).
 Importantly for the energy system, the band experiencing cold with return period greater than 50 years includes a substantial fraction of Texas’s population (\cref{fig:local_era5}a) and natural gas generation (\cref{fig:local_era5}d).
-Outside the Texas Interconnect region, much of the Texas Panhandle and central Oklahoma experienced intense cold with a return period greater than 100 years at multiple durations.
+Outside the Texas Interconnection region, much of the Texas Panhandle and central Oklahoma experienced intense cold with a return period greater than 100 years at multiple durations.
 Although Oklahoma recorded some water main breaks \cite{crum_water:2021} and rolling blackouts for several hours \cite{money_oklahoma:2021}, infrastructure impacts were less severe than in Texas despite recording more extreme (relative both to its own historical record and in absolute terms) temperatures.
@@
     Contours enclose regions that recorded 50 and 100 year return levels.
-    The gray region in panels (a) and (d) shows boundaries of the Texas Interconnect \cite{useia_regions:2021}.
+    The gray region in panels (a) and (d) shows boundaries of the Texas Interconnection \cite{useia_regions:2021}.
   }\label{fig:local_era5}
@@
 
-Our spatially aggregated metric of inferred demand for heating shows that the February 2021 event was intense but not unforeseeable (\cref{fig:idf_weighted}).
-Although specific locations experienced very intense ($>100$ year return period) temperatures, we find that for most locations in Texas the temperatures recorded during the February 2021 cold snap had precedent in the historical record and could have been anticipated given available data.
+Our spatially aggregated metric of inferred per capita demand for heating shows that the February 2021 event was intense but not unforeseeable (\cref{fig:idf_weighted}).
+Although specific locations experienced very intense ($>100$ year return period) temperatures, we find that for most locations in Texas the temperatures recorded during the February 2021 cold snap had precedent in the historical record.
 
-A proximate cause of load shedding in the Texas Interconnect during February 2021 was the vulnerability of the electricity generation system to the severe cold temperatures \cite{everhart_iea:2021}.
+A proximate cause of load shedding in the Texas Interconnection during February 2021 was the vulnerability of the electricity generation system to the severe cold temperatures \cite{everhart_iea:2021}.
 \Cref{fig:local_era5,fig:local_ghcnd} show that this happened even though most parts of the state had previously experienced similarly intense cold, notably in 1989.
-While our analysis neglects other meteorological factors, like freezing rain, that may have impeded operations at specific facilities, our analysis suggests that the February 2021 failures of energy and electricity systems in the Texas Interconnect took place during temperatures comparable to those previously recorded.
+While our analysis neglects other meteorological factors, like freezing rain, that may have impeded operations at specific facilities, our analysis suggests that the February 2021 failures of energy and electricity systems in the Texas Interconnection took place during temperatures comparable to those previously recorded.
 Similarly, water mains broke in places like Houston where temperatures did not exceed 100-year return levels, underscoring characteristic vulnerability of critical infrastructure systems \cite{chester_reliable:2020}.
@@
 
-Our analysis quantifies the probability with which temperature extremes could have been anticipated \emph{a priori}.
+Our analysis quantifies the frequency with which the temperatures observed during February 2021 could have been expected to occur \emph{a priori}.
 Other factors also govern infrastructure performance and failure, including precipitation, the demand for natural gas in adjacent regions, and complex connections within and between regional systems.
@@
 The February 2021 cold snap was the most intense in 30 years, but was not without precedent in the full historical record.
-In addition to the record cold conditions of 1899 (\cref{fig:historic_bk}), we estimate that the weather of December 1989 would have resulted in slightly higher 6-hour and 2-day aggregate heating demands over the Texas Interconnect had it occurred in February 2021.
+In addition to the record cold conditions of 1899 (\cref{fig:historic_bk}), we estimate that the weather of December 1989 would have resulted in slightly higher 6-hour and 2-day aggregate heating demands over the Texas Interconnection had it occurred in February 2021.
 Several other storms since 1950 would have produced nearly as much demand for heating.
-Given upward trends in the electrification of heating, it is likely that future cold snaps will cause peak annual loads on the Texas Interconnect to occur during the winter season.
+Given upward trends in the electrification of heating, it is likely that future cold snaps will cause peak annual loads on the Texas Interconnection to occur during the winter season.
 Infrastructure expansion necessitated by a rapidly growing population offers Texas the opportunity to invest in a more resilient energy system.
@@
 
-\subsection{Inferred demand for heating}
+\subsection{Inferred per capita demand for heating}
 
-To complement our analysis of inferred demand for heating, we consider how results change as a function of two modeling decisions.
+To complement our analysis of inferred per capita demand for heating, we consider how results change as a function of two modeling decisions.
 First, we consider what happens if the spatial field demand for heating is aggregated using grid cell area rather than population density.
@@
   \caption{
-    A time series of inferred demand for heating over the Texas Interconnect during the February 2021 cold snap.
+    A time series of inferred per capita demand for heating over the Texas Interconnection during the February 2021 cold snap.
   }\label{fig:hdd_ts}

### 2021-05-03 scripts/berkeley_earth_hdd.py
use 65F not 68F (code only)

@@ -7,7 +7,7 @@ import argparse
 import numpy as np
 import xarray as xr
 
-BASE_TEMP = 68
+BASE_TEMP = 65
 
 
 def main() -> None:

### 2021-05-03 scripts/era5_hdd.py

@@ -8,7 +8,7 @@ import argparse
 import numpy as np
 import xarray as xr
 
-BASE_TEMP = 68
+BASE_TEMP = 65
 GPWV4_YEARS = [2000, 2005, 2010, 2015, 2020]
 
 

### 2021-05-03 scripts/era5_return.py

@@ -11,7 +11,7 @@ import xarray as xr
 
 from codebase.calc import cold_return_period
 
-BASE_TEMP = 68
+BASE_TEMP = 65
 DURATIONS = [6, 24, 48, 96]
 
 

### 2021-05-03 doc/submission_02.tex
still working through referee comments

@@
   In particular, the failure of interconnected energy systems reduced electricity supply just as heating demands spiked, leaving millions of Texans without heat or electricity, many for several days.
-  This motivates the question: were the temperatures that contributed to this infrastructure failure beyond what was known to be possible, or did historical storms offer a precedent?
-  We compute the population weighted temperature excursion below \SI{68}{\degree F} as a proxy for heating demand and use this metric to answer the question ``what would the aggregate demand for heating have been had historic cold snaps occurred today?''.
-  We find that local temperatures and the inferred demand for heating across the region served by the Texas Interconnection during a storm in December 1989 were more severe than those recorded during February 2021, and that several other storms in the modern era were comparable.
+  This motivates the question: did historical storms suggest that such temperatures could be anticipated, and if so with what frequency?
+  We compute the a temperature-based proxy for heating demand and use this metric to answer the question ``what would the aggregate demand for heating have been had historic cold snaps occurred with today's population?''.
+  We find that local temperatures and the inferred demand for heating across the region served by the Texas Interconnection during a storm in December 1989 were more severe than those recorded during February 2021, and that cold snaps in 1951 and 1983 were nearly as severe.
   Given anticipated population growth, future storms may lead to even greater infrastructure failures if adaptive investments are not made.
-  Further, electricity system managers should anticipate that upward trends in electrification of heating may cause peak annual loads on the Texas Interconnection to occur during winter storms.
-
+  Further, electricity system managers should anticipate that trends in electrification of heating may cause peak annual loads on the Texas Interconnection to occur during winter storms in some years.
 \end{abstract}
@@
 
-Between February 14th and 17th, 2021, a northern air mass blanketed much of the continental United States, causing anomalously low surface temperatures across the Great Plains.
-While polar air excursions are not unusual, they do not typically reach as far south as the Mexican border.
-The state of Texas was particularly hard hit, with coincident and cascading failures of natural gas production, power generation, transportation, and water systems leaving millions of Texans without electricity, heat, and water, many for several days.
-The intensity and widespread nature of this weather event underscore a high vulnerability to cold temperatures in places that did not experience a comparable event in preceding decades.
-Even as energy demand for winter heating in Texas has grown dramatically with population and economic development, the system’s vulnerability to cold jeopardized lives and property and brought the Texas electricity grid within minutes of collapse.
+Between February 14th and 17th, 2021, a  northern air mass blanketed much of thecontinental United States, causing anomalously low surface temperatures across the Great Plains.
+The state of Texas was particularly hard hit, with coincident and cascading failures of natural gas production, power generation, transportation, and water systems leaving millions of Texans without electricity, heat, and water, many for several days \cite{ceser_winterupdate:2021,clack_uri:2021,smead_eyesoftx:2021}.
+These failures disproportionately affected vulnerable populations \cite{dobbins_blackoutdisparity:2021}, left at least 111 Texans dead \cite{mulcahy_urideath:2021}, and brought the Texas electricity grid within minutes of collapse \cite{magness_review:2021}.
 
-Since production and distribution of natural gas and electricity is possible under conditions far colder than any Texas experienced in February 2021, energy system failures reflect inadequate preparedness for cold.
-If these temperatures were unprecedented in the region covered by the Texas Interconnection then it is possible that the impacts resulted from an event outside the design envelope.
-On the other hand, if similar events were in the record then one would expect a higher level of preparation.
+Since production and distribution of electricity is possible under conditions far colder than any Texas experienced in February 2021, energy system failures reflect inadequate preparedness for cold.
+These failures occurred both because electricity demand exceeded projections, and because electricity supply failed to meet them.
+On the demand side, the Electric Reliability Council of Texas (ERCOT), which operates the Texas Interconnection bulk electric power system (hence ``Texas Interconnection''), estimated that the peak demand would have been \SI{76819}{\mega\watt} without load shedding.
+This surpassed ERCOT's ``extreme winter forecast'' of \SI{67208}{\mega\watt} in its seasonal assessment of resource adequacy \cite{ercotpublic_sarawinter:2020}.
+On the supply side, ERCOT experienced over \SI{30000}{\mega\watt} of lost output over two days due to outages and derates caused by cold temperatures \cite{ercotpublic_outagesv2:2021}.
+A large fraction of this supply shortfall, which exceeded ERCOT's worst-case scenario for forced outages, came from diverse failure modes in the natural gas supply chain \cite{ercotpublic_outagesv2:2021,smead_eyesoftx:2021,ceser_winterupdate:2021}.
+
+If temperatures experienced in the region served by the Texas Interconnection were unprecedented, then it might be expected that engineering designs derived from  historical data would exhibit substantial vulnerability.
+On the other hand, if the region had experienced similar events previously then a higher level of preparation would be expected.
 It is therefore important to assess whether historical data offered a precedent for the temperatures observed during February 2021.
-To answer this question, we first compute the population weighted temperature excursion below \SI{68}{\degree F} as a proxy for the unknown heating demand, and use this metric to explore the frequency with which this event could have been expected to occur using historical data.
+
+In this article we focus on the demand side, recognizing the complex interdependencies of electricity supply.
+We first compute the population weighted difference between observed temperatures and a standard indoor temperature of \SI{65}{\degree F} as a proxy for the unknown heating demand, and use this metric to explore the frequency with which this event could have been expected to occur using historical data.
 Next, we conduct a spatially distributed analysis to assess how unexpected the cold experienced by local roads, water mains, gas pipelines, energy generation facilities, and critical infrastructure installations was across Texas.
-We conclude by discussing the implications of these findings for long-term electricity systems planning given anticipated population growth and electrification.
+We conclude by discussing the implications of these findings for long-term electricity systems planning given anticipated growth of population and electrification.
 
@@
 
-It is well documented that Texas has previously experienced severe cold, most notably in 1899, 1951, 1983, 1989, and 2011.
-The specific spatiotemporal structure of a cold event, and its correspondence with population centers, modulates the grid-wide demand for heating (\cref{sec:inferred-demand}).
+Texas state climatologist John Nielsen-Gammon wrote in 2011 that ``winter weather is a danger to TX in part because it is so rare,'' \cite{nielsen-gammon_txacc:2011}.
+Previous cold snaps in Texas, notably in 1899, 1951, 1983, 1989, and 2011, have affected not only humans but also ecological systems.
+For example, the 1951 cold event caused a  significant die-off of fish life in the shallow Gulf Coast \cite{gunter_fishes:1951}.
+The specific spatiotemporal structure of a cold event, and its correspondence with population centers, determines the grid-wide demand for heating (see \cref{sec:inferred-demand}).
 The structure of the storm also drives the aggregated hazard to energy infrastructure, which has implications for the costs and benefits of infrastructure hardening.
@@
     Severe cold snaps that affect Texas and extend into the central United States have several precedents in the historical record.
-    Plot shows anomalies of daily mean temperatures for historic major cold events affecting Texas, defined as the departure from the seasonal (December-February) mean to facilitate comparison of events in different calendar months.
+    Plot shows anomalies of daily mean temperatures from the ERA-5 reanalysis \cite{hersbach_era5:2020} for historic major cold events affecting Texas, defined as the departure from the seasonal (December-February) mean to facilitate comparison of events in different calendar months.
     Temperatures over the Continental United States are plotted to show large-scale patterns, but events are selected based only on Texas temperatures.
@@
 
-We use three distinct datasets to analyze temperature minima in the region covered by the Texas Interconnection bulk electric power system administered by the Electric Reliability Council of Texas (hence ``Texas Interconnection'') through the lens of distributed (each grid cell analyzed separately) and aggregated (weighted averages taken across space) extreme values analysis.
+We use three distinct datasets to analyze temperature minima in the region covered by the Texas Interconnection through the lens of distributed (each grid cell analyzed separately) and aggregated (weighted averages taken across space) extreme values analysis.
 
@@
 \section{Discussion}
+
+\begin{itemize}
+  \item \cite{nerc_previous:2013} examined the 1983 and 1989 cold snaps, among others, and found that ``two common themes observed in the reports of both the severe and lesser cold weather incidents were: Constraints on natural gas fuel supplies to generating plants; and Generating unit trips, derates, or failure to start due to weather related causes (e.g.
+frozen sensing lines, etc.)''
+  \item \cite{wu_tx:2021} present a model of the electricity system and suggests that there may have been a possibility for corrective operational actions to lessen the impacts of generation outages.
+  \item \cite{cartwright_lightsout:2021} calls on Texas officials to examine other business and regulatory models in light of the failures that occurred in February 2021.
+  \item \cite{nerc_operation:1989} and \cite{osborne_twofreezes:2021} catalog the effects of the December 21-23, 1989 cold snap on the Texas electricity system, notably that there were approximately two hours of rolling blackouts.
+\end{itemize}
 

### 2021-05-03 codebase/calc.py
reorganize the `codebase`. then delete data and re-run from scratch

@@ -45,6 +45,31 @@ def cold_return_period(y: float, x: np.ndarray) -> float:
     return return_period(-y, -x)
 
 
+def calc_plotposition_return_level(X, T):
+    """
+    Use an empirical formula for plot position rather than MLE GEV fit
+
+    https://www.itl.nist.gov/div898/handbook/eda/section3/eda366g.htm
+    https://www.engr.colostate.edu/~ramirez/ce_old/classes/cive322-Ramirez/IDF-Procedure.pdf
+    """
+    KT = -np.sqrt(6) / np.pi * (0.5772 + np.log(np.log(T / (T - 1))))
+    Xbar = np.mean(X)
+    S = np.std(X)
+    return Xbar + KT * S
+
+
+def calc_mle_return_level(X, T):
+    """Calculate the GEV return level using MLE"""
+    theta = gev.fit(
+        X,
+        0,  # guess for shape
+        loc=40,  # guess for location
+        scale=10,  # guess for scale
+        optimizer=optimizer,
+    )
+    return gev(*theta).isf(1 / T)
+
+
 def calculate_anomaly(da, groupby_type="time.month"):
     """Function to calculate anomalies from xarray docs"""
     clim = da.groupby(groupby_type).mean(dim="time")

### 2021-05-03 codebase/data.py

@@ -6,28 +6,23 @@ Functions to read and parse temperature data from three sources:
 3. GHCN Stations
 """
 
-import os
-
 import pandas as pd
 import xarray as xr
 
 from .read_ghcn import read_ghcn_data_file
-
-PARDIR = os.path.abspath(os.path.join(os.path.dirname(__file__), os.pardir, os.pardir))
+from .path import datadir
 
 
 def get_berkeley_earth() -> xr.Dataset:
     """Get the Berkeley Earth data"""
     temp = xr.Dataset(
         dict(
-            tmax=xr.open_dataarray(
-                os.path.join(PARDIR, "data", "processed", "berkeleyearth", "TMAX.nc")
-            ),
+            tmax=xr.open_dataarray(datadir("processed", "berkeleyearth", "TMAX.nc")),
             tmin=xr.open_dataarray(
-                os.path.join(PARDIR, "data", "processed", "berkeleyearth", "TMIN.nc")
+                datadir("processed", "processed", "berkeleyearth", "TMIN.nc")
             ),
             tavg=xr.open_dataarray(
-                os.path.join(PARDIR, "data", "processed", "berkeleyearth", "TAVG.nc")
+                datadir("processed", "processed", "berkeleyearth", "TAVG.nc")
             ),
         )
     )
@@ -36,7 +31,7 @@ def get_berkeley_earth() -> xr.Dataset:
 
 def get_ghcn_data(stnid: str) -> pd.DataFrame:
     """Get the GHCN data for a particular file"""
-    fname = os.path.join(PARDIR, "data", "raw", "ghcnd_all", f"{stnid}.dly")
+    fname = datadir("raw", "ghcnd_all", f"{stnid}.dly")
     data = read_ghcn_data_file(fname, variables=["TMIN", "TMAX"]).apply(pd.to_numeric)
     if len(data.columns) >= 2:
         data = data.loc[lambda df: df["TMAX"] > df["TMIN"]]  # otherwise NOT credible

### 2021-05-03 codebase/fig.py

@@ -5,18 +5,13 @@ from string import ascii_lowercase as letters
 
 import matplotlib.pyplot as plt
 
-from .data import PARDIR
-
-FIGDIR = os.path.join(PARDIR, "fig")
+from .path import figdir
 
 
 def smart_save(name: str) -> None:
     """Get the path to figures"""
-    if not os.path.isdir(FIGDIR):
-        os.makedirs(FIGDIR)
-
-    plt.savefig(os.path.join(FIGDIR, f"{name}.pdf"))
-    plt.savefig(os.path.join(FIGDIR, f"{name}.jpg"), dpi=300)
+    plt.savefig(figdir(f"{name}.pdf"))
+    plt.savefig(figdir(f"{name}.jpg"), dpi=300)
 
 
 def add_panel_text(

### 2021-05-03 codebase/path.py

@@ -1,0 +1,34 @@
+"""
+Make it easy to work with the directory structure
+
+Syntax roughly follows that of DrWatson.jl:
+https://juliadynamics.github.io/DrWatson.jl/dev/
+"""
+
+import os
+
+PARDIR = os.path.abspath(os.path.join(os.path.dirname(__file__), os.path.pardir))
+FIGDIR = os.path.join(PARDIR, "fig")
+DATADIR = os.path.join(PARDIR, "data")
+SCRIPTDIR = os.path.join(PARDIR, "scripts")
+
+
+def datadir(*args):
+    """Get the directory of something in data"""
+    if not os.path.isdir(DATADIR):
+        os.makedirs(DATADIR)
+    return os.path.join(DATADIR, *args)
+
+
+def scriptdir(*args):
+    """Get the directory of something in the scripts folder"""
+    if not os.path.isdir(SCRIPTDIR):
+        os.makedirs(SCRIPTDIR)
+    return os.path.join(SCRIPTDIR, *args)
+
+
+def figdir(*args):
+    """Get the directory of a figure"""
+    if not os.path.isdir(FIGDIR):
+        os.makedirs(FIGDIR)
+    return os.path.join(FIGDIR, *args)

### 2021-05-03 scripts/era5_return.py

@@ -24,7 +24,7 @@ def main() -> None:
     parser.add_argument("-o", "--outfile", type=str)
     args = parser.parse_args()
 
-    temp = xr.open_mfdataset(args.infiles)["t2m"].mean(dim="expver").compute()
+    temp = xr.open_mfdataset(args.infiles)["t2m"].compute()
     temp_roll = xr.concat(
         [
             temp.rolling(time=dur).mean().assign_coords({"duration": dur})

### 2021-05-03 setup.py

@@ -1,0 +1,10 @@
+from distutils.core import setup
+
+setup(
+    name="codebase",
+    version="0.1",
+    packages=[
+        "codebase",
+    ],
+    long_description="helpful tools for this analysis",
+)

### 2021-05-03 doc/submission_02.tex
first stab at addressing reviewer comments

@@
 
-Between February 14th and 17th, 2021, a  northern air mass blanketed much of thecontinental United States, causing anomalously low surface temperatures across the Great Plains.
+Between February 14th and 17th, 2021, a  northern air mass blanketed much of the continental United States, causing anomalously low surface temperatures across the Great Plains.
 The state of Texas was particularly hard hit, with coincident and cascading failures of natural gas production, power generation, transportation, and water systems leaving millions of Texans without electricity, heat, and water, many for several days \cite{ceser_winterupdate:2021,clack_uri:2021,smead_eyesoftx:2021}.
@@
 These failures occurred both because electricity demand exceeded projections, and because electricity supply failed to meet them.
-On the demand side, the Electric Reliability Council of Texas (ERCOT), which operates the Texas Interconnection bulk electric power system (hence ``Texas Interconnection''), estimated that the peak demand would have been \SI{76819}{\mega\watt} without load shedding.
+On the demand side, the Electric Reliability Council of Texas (ERCOT), which operates the Texas Interconnection bulk electric power system (hence ``Texas Interconnection''), estimated that the peak demand would have been \SI{76819}{\mega\watt} without load shedding \cite{magness_review:2021}.
 This surpassed ERCOT's ``extreme winter forecast'' of \SI{67208}{\mega\watt} in its seasonal assessment of resource adequacy \cite{ercotpublic_sarawinter:2020}.
@@
 
-If temperatures experienced in the region served by the Texas Interconnection were unprecedented, then it might be expected that engineering designs derived from  historical data would exhibit substantial vulnerability.
-On the other hand, if the region had experienced similar events previously then a higher level of preparation would be expected.
+If temperatures experienced in the region served by the Texas Interconnection were unprecedented, then engineering designs derived from  historical data might exhibit substantial vulnerability.
+On the other hand, if the region had experienced similar events previously then it would suggest that these vulnerabilities could have been anticipated.
 It is therefore important to assess whether historical data offered a precedent for the temperatures observed during February 2021.
 
-In this article we focus on the demand side, recognizing the complex interdependencies of electricity supply.
-We first compute the population weighted difference between observed temperatures and a standard indoor temperature of \SI{65}{\degree F} as a proxy for the unknown heating demand, and use this metric to explore the frequency with which this event could have been expected to occur using historical data.
-Next, we conduct a spatially distributed analysis to assess how unexpected the cold experienced by local roads, water mains, gas pipelines, energy generation facilities, and critical infrastructure installations was across Texas.
+To answer this question, we first compute the population weighted difference between observed temperatures and a standard indoor temperature of \SI{65}{\degree F} as a proxy for the unknown heating demand, then use standard statistical procedures to assess the probability with which the temperatures observed during February 2021 might have been expected to occur \emph{a priori}.
+We then supplement this with a spatially distributed analysis of how unexpected the cold experienced by local roads, water mains, gas pipelines, energy generation facilities, and critical infrastructure installations was across Texas.
 We conclude by discussing the implications of these findings for long-term electricity systems planning given anticipated growth of population and electrification.
@@
 Texas state climatologist John Nielsen-Gammon wrote in 2011 that ``winter weather is a danger to TX in part because it is so rare,'' \cite{nielsen-gammon_txacc:2011}.
-Previous cold snaps in Texas, notably in 1899, 1951, 1983, 1989, and 2011, have affected not only humans but also ecological systems.
+Previous cold snaps in Texas, notably in 1899, 1951, 1983, 1989, and 2011 (see \cref{fig:historic_era5,fig:historic_bk}), have affected both human and ecological systems.
 For example, the 1951 cold event caused a  significant die-off of fish life in the shallow Gulf Coast \cite{gunter_fishes:1951}.
+
 The specific spatiotemporal structure of a cold event, and its correspondence with population centers, determines the grid-wide demand for heating (see \cref{sec:inferred-demand}).
@@
 It is therefore important to assess the weather conditions that led to these infrastructure failures and identify whether they had historical precedent.
-
-\Cref{fig:historic_era5} and supplemental \cref{fig:historic_bk,fig:historic_tx} show historic cold snaps in Texas.
+The spatiotemporal patterns of historical cold snaps in Texas are illustrated in \cref{fig:historic_era5} and supplemental \cref{fig:historic_bk,fig:historic_tx}.
 Although the spatiotemporal structure of each event is distinct, it is apparent that cold extremes in Texas tend to co-occur with cold temperatures across much of the United States, particularly the Great Plains.
-While the 2021 event was severe, daily temperature extrema in Texas appear comparable to historical events.
+While the 2021 event was severe, daily temperature extrema in Texas appear qualitatively comparable to historical events.
 The ``Great Blizzard'' of February 1899, shown in \cref{fig:historic_bk}, caused even more intense cold.
@@
     Severe cold snaps that affect Texas and extend into the central United States have several precedents in the historical record.
-    Plot shows anomalies of daily mean temperatures from the ERA-5 reanalysis \cite{hersbach_era5:2020} for historic major cold events affecting Texas, defined as the departure from the seasonal (December-February) mean to facilitate comparison of events in different calendar months.
-    Temperatures over the Continental United States are plotted to show large-scale patterns, but events are selected based only on Texas temperatures.
+    Plot shows anomalies of daily mean temperatures from the ERA-5 reanalysis \cite{hersbach_era5:2020} for historic major cold events affecting Texas, defined as the departure from the seasonal (December-February) mean of the observational record.
+    Anomalies facilitate identification of large-scale weather patterns superimposed on long-term climatological averages.
     Hourly temperatures are averaged to 1-day (a,d,g,j,m), 3-day (b,e,h,k,n), and 5-day (c,f,i,l,o) average temperature anomalies.
@@
 
-For each event duration considered, return periods are obtained by fitting a stationary generalized extreme value (GEV) distribution to the time series of annual minimum December-February (DJF) temperatures, or to the time series of annual maxima of inferred per capita demand for heating (\cref{sec:inferred-demand}).
-Events that occur in December are coded to the following year.
+Return periods define the probability with which a particular event can be expected to occur.
+Thus, a $T$-year event has a $\frac{1}{T}$ probability of occurring in a given year.
+For each event duration considered, we calculated return periods by fitting a stationary generalized extreme value (GEV) distribution to the time series of annual maxima of inferred per capita demand for heating (see \cref{sec:inferred-demand}).
+Events that occur in December are coded to the following year so that a single winter season is grouped together.
 The 2021 winter season was excluded from return period estimates, allowing us to interpret return periods for the February 2021 event as \emph{a priori} estimates.
@@
 Importantly for the energy system, the band experiencing cold with return period greater than 50 years includes a substantial fraction of Texas’s population (\cref{fig:local_era5}a) and natural gas generation (\cref{fig:local_era5}d).
-Outside the Texas Interconnection region, much of the Texas Panhandle and central Oklahoma experienced intense cold with a return period greater than 100 years at multiple durations.
-Although Oklahoma recorded some water main breaks \cite{crum_water:2021} and rolling blackouts for several hours \cite{money_oklahoma:2021}, infrastructure impacts were less severe than in Texas despite recording more extreme (relative both to its own historical record and in absolute terms) temperatures.
+Outside the Texas Interconnection region, the Midcontinent Independent System Operator and Southwest Power Pool instructed utilities to shed firm load, but of the approximately 4.89 million customers without power in Texas, Louisiana, and Oklahoma, an estimated 4.5 million were in Texas \cite{ceser_winterupdate:2021}.
 
@@
 
-\begin{itemize}
-  \item \cite{nerc_previous:2013} examined the 1983 and 1989 cold snaps, among others, and found that ``two common themes observed in the reports of both the severe and lesser cold weather incidents were: Constraints on natural gas fuel supplies to generating plants; and Generating unit trips, derates, or failure to start due to weather related causes (e.g.
-frozen sensing lines, etc.)''
-  \item \cite{wu_tx:2021} present a model of the electricity system and suggests that there may have been a possibility for corrective operational actions to lessen the impacts of generation outages.
-  \item \cite{cartwright_lightsout:2021} calls on Texas officials to examine other business and regulatory models in light of the failures that occurred in February 2021.
-  \item \cite{nerc_operation:1989} and \cite{osborne_twofreezes:2021} catalog the effects of the December 21-23, 1989 cold snap on the Texas electricity system, notably that there were approximately two hours of rolling blackouts.
-\end{itemize}
-
 Our spatially aggregated metric of inferred per capita demand for heating shows that the February 2021 event was intense but not unforeseeable (\cref{fig:idf_weighted}).
@@
 \Cref{fig:local_era5,fig:local_ghcnd} show that this happened even though most parts of the state had previously experienced similarly intense cold, notably in 1989.
-While our analysis neglects other meteorological factors, like freezing rain, that may have impeded operations at specific facilities, our analysis suggests that the February 2021 failures of energy and electricity systems in the Texas Interconnection took place during temperatures comparable to those previously recorded.
-Similarly, water mains broke in places like Houston where temperatures did not exceed 100-year return levels, underscoring characteristic vulnerability of critical infrastructure systems \cite{chester_reliable:2020}.
+Yet despite temperatures that were, in aggregate, more intense, the electricity system experienced fewer than three hours of rolling blackouts from December 21-23, 1989 \cite{nerc_operation:1989,osborne_twofreezes:2021}.
+Following the 1983, 1989, and 2011 cold snaps, the North American Electric Reliability Corporation (NERC) warned that ``constraints on natural gas fuel supplies to generating plants'' and ``generating unit trips, derates, or failure to start due to weather related causes'' were common themes across severe and moderate cold extremes \cite{nerc_previous:2013}, foreshadowing many of the causes of February 2021 energy system failures identified by ERCOT \cite{ercotpublic_outagesv2:2021,magness_review:2021}.
+While our analysis neglects other meteorological factors, like freezing rain, that may have impeded operations at specific facilities, we find that the February 2021 failures of energy and electricity systems in the Texas Interconnection took place during temperatures with precedent in the historical record.
 
 Another cause of load shedding was the high demand for electricity that low temperatures induced.
-Currently over 50\% of Texas households use electricity for space heating over the majority of census tracts of the state \cite{waite_heating:2020} and further electrification is a central element of many plans to decarbonize our energy sector \cite{williams_decarbonization:2012,davis:2018,white_txresidential:2019}.
+Currently at 50\% of Texas households use electricity for space heating over the majority of census tracts of the state \cite{waite_heating:2020} and further electrification is a central element of many plans to decarbonize our energy sector \cite{williams_decarbonization:2012,davis:2018,white_txresidential:2019}.
 While summer peak loads have been a central planning concern on the Texas grid in the past, it is likely that winter peak loads will become a greater concern in the coming decades.
-In fact, the peak demand during this event was estimated at around \SI{74.5}{\giga\watt}, just shy of the all time peak for any season of \SI{74.8}{\giga\watt} in August 2019 \cite{everhart_iea:2021}.
-As electrification of heating increases, severe cold snaps may represent the peak demand on the Texas Interconnect.
+In fact, the estimated \SI{76819}{\mega\watt} of peak demand without load shedding during this event \cite{magness_review:2021} exceeded not only the previous winter demand record of \SI{65900}{\mega\watt} recorded on January 17, 2018 but also the all-season record actual demand of \SI{74800}{\mega\watt} recorded on August 19, 2019 \cite{everhart_iea:2021}.
+As electrification of heating continues, severe cold snaps may consistently represent the peak demand on the Texas Interconnect.
 
@@
 More fundamentally, an assessment of exposure to cold extremes over the next decades should consider the deeply uncertain distribution of future climate change, and the induced effect on cold extremes in Texas.
-Although a broad scientific consensus suggests the frequency of cold extremes should decrease under warming in most places \cite{ipcc_ar5:2014}, possible links between North-South temperature gradients and mid-latitude temperature extremes remains an area of active research \cite{Barnes:2013fp,Cohen:2014gx,Screen:2013ho}.
-Regardless, the effect of climate change on peak demand for heating is likely to be small compared to the effect of rapid population growth; for example, the Texas Water Development Board anticipates at least 40\% growth from 2020 to 2050 \cite{texaswaterplan:2012}.
+Although a broad scientific consensus suggests the frequency of cold extremes should decrease under warming in most places \cite{ipcc_ar5:2014}, possible links between North-South temperature gradients and mid-latitude temperature extremes remains an area of active research \cite{Barnes:2013fp,Cohen:2014gx,Screen:2013ho,romanowsky_stratosphere:2019}.
+Regardless, the effect of climate change on peak demand for heating is likely to be small compared to the effect of rapid population growth which the Texas Water Development Board, for example, anticipates to be 40\% from 2020 to 2050 \cite{texaswaterplan:2012}.
 
@@
 The February 2021 cold snap was the most intense in 30 years, but was not without precedent in the full historical record.
-In addition to the record cold conditions of 1899 (\cref{fig:historic_bk}), we estimate that the weather of December 1989 would have resulted in slightly higher 6-hour and 2-day aggregate heating demands over the Texas Interconnection had it occurred in February 2021.
-Several other storms since 1950 would have produced nearly as much demand for heating.
+In addition to the record cold conditions of 1899 (\cref{fig:historic_bk}), we estimate that the weather of December 1989 would have resulted in higher 6-hour and 2-day aggregate per capita demand for heating over the Texas Interconnection than the February 2021 event had it occurred with today's population.
+Storms in February 1951, January 1962, and December 1983 would have resulted in at least 90\% as much per capita demand for heating had they occurred today.
 Given upward trends in the electrification of heating, it is likely that future cold snaps will cause peak annual loads on the Texas Interconnection to occur during the winter season.
@@
 The authors thank
+Daniel Cohan,
 Sylvia Dee,
 Elisabeth Gawthrop,
+Pedram Hassanzadeh,
+Fred Heutte,
 Adam Massmann,
+Alison Silverstein,
 and
 Vivek Srikrishnan
-for comments on a preliminary draft of this manuscript.
+for discussion of this work.
 The authors also thank the developers and maintainers of the open source packages we used, in particular the pydata and pangeo communities, and the many journalists and academics whose timely reporting and commentary has informed our thinking.

### 2021-05-05 doc/submission_02.tex
revisions: copy to circulate w/ coauthors

@@
   This motivates the question: did historical storms suggest that such temperatures could be anticipated, and if so with what frequency?
-  We compute the a temperature-based proxy for heating demand and use this metric to answer the question ``what would the aggregate demand for heating have been had historic cold snaps occurred with today's population?''.
+  We compute a temperature-based proxy for heating demand and use this metric to answer the question ``what would the aggregate demand for heating have been had historic cold snaps occurred with today's population?''.
   We find that local temperatures and the inferred demand for heating across the region served by the Texas Interconnection during a storm in December 1989 were more severe than those recorded during February 2021, and that cold snaps in 1951 and 1983 were nearly as severe.
@@
 The structure of the storm also drives the aggregated hazard to energy infrastructure, which has implications for the costs and benefits of infrastructure hardening.
-It is therefore important to assess the weather conditions that led to these infrastructure failures and identify whether they had historical precedent.
 The spatiotemporal patterns of historical cold snaps in Texas are illustrated in \cref{fig:historic_era5} and supplemental \cref{fig:historic_bk,fig:historic_tx}.
@@
 \end{enumerate}
-We also use population density data from the GPWv4 dataset \cite{ciesin_gpwv4:2016}, a list of power generation facilities from the US Energy Information Administration \cite{useia_generators:2021}, and a map of the Texas Interconnection \cite{useia_regions:2021}.%TODO: add and cite information on the generators that did and did not go out.
+We also use population density data from the GPWv4 dataset \cite{ciesin_gpwv4:2016}, a list of power generation facilities from the US Energy Information Administration \cite{useia_generators:2021}, and a map of the Texas Interconnection \cite{useia_regions:2021}.
 
-\subsection{Inferred per Capita Demand for Heating}\label{sec:inferred-demand}
+\subsection{Inferred heating demand per capita}\label{sec:inferred-demand}
 
 Most space heating in Texas is either electric or gas \cite{waite_heating:2020} and the majority of power generation in the Texas Interconnection depends on natural gas \cite{everhart_iea:2021}.
-Stress on natural gas production and delivery was therefore just as important as the more visible stress on the electric system.
+Stress on natural gas production and delivery was therefore just as important as the more visible stress on the electric system \cite{ercotpublic_outagesv2:2021}.
 
-The hourly or daily thermal energy requirement for space heating is primarily driven by how much lower the ambient temperature is than an indoor comfort temperature of \SI{68}{\degree F}.
+The hourly or daily thermal energy requirement for space heating is primarily driven by how much lower the ambient temperature is than an indoor comfort temperature of \SI{65}{\degree F}.
 This relationship is often expressed in terms of heating degree days or hours.
-We therefore consider the temperature excursion from \SI{68}{\degree F} as a proxy for thermal heating demand.
-We compute this value each hour for the ERA5 data, defining heating demand at each grid cell as $\text{HD}_t = \max (68 - T_t, 0)$, where $T_t$ is the temperature at hour $t$ in \si{\degree F}.
-The Berkeley Earth and GHCN datasets provide daily minimum and maximum temperatures, so we define heating demand at each grid cell or station as $\text{HD}_d = \max (68 -\frac{T_{\text{min},d} + T_{\text{max},d}}{2}, 0)$, where $T_{\text{min},d}$ is the minimum temperature recorded on day $d$ and $T_{\text{max},d}$ is the maximum temperature recorded on day $d$, both in \si{\degree F}.
+We therefore consider the difference between observed temperatures and a standard indoor temperature of \SI{65}{\degree F} as a proxy for thermal heating demand.
+We compute this value each hour for the ERA5 data, defining heating demand at each grid cell as $\text{HD}_t = \max (65 - T_t, 0)$, where $T_t$ is the temperature at hour $t$ in \si{\degree F}.
+The Berkeley Earth and GHCN datasets provide daily minimum and maximum temperatures, so we define heating demand at each grid cell or station as $\text{HD}_d = \max (65 -\frac{T_{\text{min},d} + T_{\text{max},d}}{2}, 0)$, where $T_{\text{min},d}$ is the minimum temperature recorded on day $d$ and $T_{\text{max},d}$ is the maximum temperature recorded on day $d$, both in \si{\degree F}.
 
 To assess how spatially correlated cold spells might affect the Texas electric grid, we average heating demand in space over the Texas Interconnection domain \cite{useia_regions:2021}, weighting each grid cell by 2020 population density \cite{ciesin_gpwv4:2016}.
-We refer to this spatially aggregated time series, which has the straightforward interpretation as the average heating demand experienced by Texas residents, as ``Inferred per Capita Demand for Heating.''
+We refer to this spatially aggregated time series, which has the straightforward interpretation as the average heating demand experienced by Texas residents, as ``Inferred Heating Demand per Capita.''
 
@@
 Thus, a $T$-year event has a $\frac{1}{T}$ probability of occurring in a given year.
-For each event duration considered, we calculated return periods by fitting a stationary generalized extreme value (GEV) distribution to the time series of annual maxima of inferred per capita demand for heating (see \cref{sec:inferred-demand}).
+For each event duration considered, we calculated return periods by fitting a stationary generalized extreme value (GEV) distribution to the time series of annual maxima of inferred heating demand per capita (see \cref{sec:inferred-demand}).
 Events that occur in December are coded to the following year so that a single winter season is grouped together.
@@
 
-\section{How extreme was inferred per capita heating demand over the Texas Interconnect?}
+\section{How extreme was inferred heating demand per capita over the Texas Interconnection?}
 
@@
 As such, understanding whether there was precedent for a cold snap simultaneously affecting several regions of Texas's grid that today have high population density is critical.
-We therefore use our measure of inferred per capita demand for heating (see \cref{sec:inferred-demand}) to represent the aggregate heating demand induced by cold temperatures.
+We therefore use our measure of inferred heating demand per capita (see \cref{sec:inferred-demand}) to represent the aggregate heating demand induced by cold temperatures.
 Aggregating historic temperature fields in space using the 2020 population, we answer the question ``what would the aggregate demand for heating have been had historic cold snaps occurred today?''
@@
   \caption{
-    The inferred per capita demand for heating induced by the February 2021 cold snap is not unprecedented.
-    For the worst 6 hours, the 1989 event was more severe than the 2021 event, while they are comparable for the longer durations.
-    (a): time series of annual maximum inferred per capita demand for heating (\cref{sec:inferred-demand}) at 6 hour and 2 day durations.
+    The inferred heating demand per capita induced by the February 2021 cold snap is not unprecedented.
+    For the worst 6 hours, the 1989 event was more severe than the 2021 event, while they are comparable for longer durations.
+    (a): time series of annual maximum inferred heating demand per capita (\cref{sec:inferred-demand}) at 6 hour and 2 day durations.
     December extremes, including the December 1989 storm, are coded to the following year so that one maximum per December-February winter season is taken.
@@
 
-Our spatially aggregated metric of inferred per capita demand for heating shows that the February 2021 event was intense but not unforeseeable (\cref{fig:idf_weighted}).
+Our spatially aggregated metric of inferred heating demand per capita shows that the February 2021 event was intense but not unforeseeable (\cref{fig:idf_weighted}).
 Although specific locations experienced very intense ($>100$ year return period) temperatures, we find that for most locations in Texas the temperatures recorded during the February 2021 cold snap had precedent in the historical record.
@@
 A proximate cause of load shedding in the Texas Interconnection during February 2021 was the vulnerability of the electricity generation system to the severe cold temperatures \cite{everhart_iea:2021}.
-\Cref{fig:local_era5,fig:local_ghcnd} show that this happened even though most parts of the state had previously experienced similarly intense cold, notably in 1989.
+As shown in \cref{fig:egova}, generator outages occurred across the state, even though most parts of the state had previously experienced similarly intense cold, notably in 1989 (\cref{fig:local_era5,fig:local_ghcnd}).
 Yet despite temperatures that were, in aggregate, more intense, the electricity system experienced fewer than three hours of rolling blackouts from December 21-23, 1989 \cite{nerc_operation:1989,osborne_twofreezes:2021}.
@@
 Another cause of load shedding was the high demand for electricity that low temperatures induced.
-Currently at 50\% of Texas households use electricity for space heating over the majority of census tracts of the state \cite{waite_heating:2020} and further electrification is a central element of many plans to decarbonize our energy sector \cite{williams_decarbonization:2012,davis:2018,white_txresidential:2019}.
+Currently at least 50\% of Texas households use electricity for space heating over the majority of census tracts of the state \cite{waite_heating:2020} and further electrification is a central element of many plans to decarbonize our energy sector \cite{williams_decarbonization:2012,davis:2018,white_txresidential:2019}.
 While summer peak loads have been a central planning concern on the Texas grid in the past, it is likely that winter peak loads will become a greater concern in the coming decades.
@@
 The February 2021 cold snap was the most intense in 30 years, but was not without precedent in the full historical record.
-In addition to the record cold conditions of 1899 (\cref{fig:historic_bk}), we estimate that the weather of December 1989 would have resulted in higher 6-hour and 2-day aggregate per capita demand for heating over the Texas Interconnection than the February 2021 event had it occurred with today's population.
-Storms in February 1951, January 1962, and December 1983 would have resulted in at least 90\% as much per capita demand for heating had they occurred today.
+In addition to the record cold conditions of 1899 (\cref{fig:historic_bk}), we estimate that the weather of December 1989 would have resulted in higher 6-hour and 2-day values of inferred heating demand per capita over the Texas Interconnection than the February 2021 event.
+Storms in February 1951, January 1962, and December 1983 would have resulted in at least 90\% as much inferred heating demand per capita.
 Given upward trends in the electrification of heating, it is likely that future cold snaps will cause peak annual loads on the Texas Interconnection to occur during the winter season.
@@
 
-\subsection{Inferred per capita demand for heating}
+\subsection{Inferred heating demand per capita}
 
-To complement our analysis of inferred per capita demand for heating, we consider how results change as a function of two modeling decisions.
+To complement our analysis of inferred heating demand per capita, we consider how results change as a function of two modeling decisions.
 First, we consider what happens if the spatial field demand for heating is aggregated using grid cell area rather than population density.
@@
   \caption{
-    A time series of inferred per capita demand for heating over the Texas Interconnection during the February 2021 cold snap.
+    A time series of inferred heating demand per capita over the Texas Interconnection during the February 2021 cold snap.
   }\label{fig:hdd_ts}
@@
 
+\begin{figure}
+  \includegraphics[width=\textwidth]{EGOVA.pdf}
+  \caption{
+    ERCOT generator outages from 12:00AM on 14 February 2021 to 12:00AM on 18 February 2021.
+    Map produced using ERCOT's Generator Outage/Derate Visualization App (EGOVA) available at \url{https://bit.ly/EGOVA}.
+  }\label{fig:egova}
+\end{figure}
+
 \end{document}

### 2021-05-05 doc/submission_02.tex
Ready for last check before resubmission

@@
 \usepackage{siunitx}
+\usepackage{xr-hyper} % to cross-ref from review response
+\usepackage{hyperref}
+\hypersetup{hidelinks}
 \usepackage{cleveref}
@@
   Winter storm Uri brought severe cold to the southern United States in February 2021, causing a cascading failure of interdependent systems in Texas where infrastructure was not adequately prepared for such cold.
-  In particular, the failure of interconnected energy systems reduced electricity supply just as heating demands spiked, leaving millions of Texans without heat or electricity, many for several days.
-  This motivates the question: did historical storms suggest that such temperatures could be anticipated, and if so with what frequency?
+  In particular, the failure of interconnected energy systems restricted electricity supply just as demand for heating spiked, leaving millions of Texans without heat or electricity, many for several days.
+  This motivates the question: did historical storms suggest that such temperatures were known to occur, and if so with what frequency?
   We compute a temperature-based proxy for heating demand and use this metric to answer the question ``what would the aggregate demand for heating have been had historic cold snaps occurred with today's population?''.
-  We find that local temperatures and the inferred demand for heating across the region served by the Texas Interconnection during a storm in December 1989 were more severe than those recorded during February 2021, and that cold snaps in 1951 and 1983 were nearly as severe.
+  We find that local temperatures and the inferred demand for heating per capita across the region served by the Texas Interconnection were more severe during a storm in December 1989 than during February 2021, and that cold snaps in 1951 and 1983 were nearly as severe.
   Given anticipated population growth, future storms may lead to even greater infrastructure failures if adaptive investments are not made.
-  Further, electricity system managers should anticipate that trends in electrification of heating may cause peak annual loads on the Texas Interconnection to occur during winter storms in some years.
+  Further, electricity system managers should prepare for trends in electrification of heating to drive peak annual loads on the Texas Interconnection during severe winter storms.
 \end{abstract}
@@
 This surpassed ERCOT's ``extreme winter forecast'' of \SI{67208}{\mega\watt} in its seasonal assessment of resource adequacy \cite{ercotpublic_sarawinter:2020}.
-On the supply side, ERCOT experienced over \SI{30000}{\mega\watt} of lost output over two days due to outages and derates caused by cold temperatures \cite{ercotpublic_outagesv2:2021}.
-A large fraction of this supply shortfall, which exceeded ERCOT's worst-case scenario for forced outages, came from diverse failure modes in the natural gas supply chain \cite{ercotpublic_outagesv2:2021,smead_eyesoftx:2021,ceser_winterupdate:2021}.
+On the supply side, the Texas Interconnection experienced over \SI{30000}{\mega\watt} of lost output for two consecutive days due to outages and derates caused by cold temperatures \cite{ercotpublic_outagesv2:2021}.
+A large fraction of this supply shortfall, which exceeded ERCOT's worst-case scenario for forced outages, originated in the natural gas supply chain \cite{ercotpublic_outagesv2:2021,smead_eyesoftx:2021,ceser_winterupdate:2021}.
 
-If temperatures experienced in the region served by the Texas Interconnection were unprecedented, then engineering designs derived from  historical data might exhibit substantial vulnerability.
-On the other hand, if the region had experienced similar events previously then it would suggest that these vulnerabilities could have been anticipated.
+If temperatures experienced in the region served by the Texas Interconnection were unprecedented, then this event might prompt discussion about the appropriate use of models to prepare for events that are theoretically possible, but beyond the observational record.
+On the other hand, historical precedent for such temperatures would suggest a broader lack of institutional and infrastructural readiness.
 It is therefore important to assess whether historical data offered a precedent for the temperatures observed during February 2021.
@@
 We then supplement this with a spatially distributed analysis of how unexpected the cold experienced by local roads, water mains, gas pipelines, energy generation facilities, and critical infrastructure installations was across Texas.
-We conclude by discussing the implications of these findings for long-term electricity systems planning given anticipated growth of population and electrification.
+We conclude by discussing the implications of these findings for long-term electricity systems planning given anticipated population growth and electrification.
 
@@
 The structure of the storm also drives the aggregated hazard to energy infrastructure, which has implications for the costs and benefits of infrastructure hardening.
-The spatiotemporal patterns of historical cold snaps in Texas are illustrated in \cref{fig:historic_era5} and supplemental \cref{fig:historic_bk,fig:historic_tx}.
+The spatiotemporal patterns of historical cold snaps in Texas are illustrated in \cref{fig:historic_era5,fig:historic_bk,fig:historic_tx}.
 Although the spatiotemporal structure of each event is distinct, it is apparent that cold extremes in Texas tend to co-occur with cold temperatures across much of the United States, particularly the Great Plains.
@@
 To assess how spatially correlated cold spells might affect the Texas electric grid, we average heating demand in space over the Texas Interconnection domain \cite{useia_regions:2021}, weighting each grid cell by 2020 population density \cite{ciesin_gpwv4:2016}.
-We refer to this spatially aggregated time series, which has the straightforward interpretation as the average heating demand experienced by Texas residents, as ``Inferred Heating Demand per Capita.''
+We refer to this spatially aggregated time series, which has the straightforward interpretation as the average heating demand experienced by a Texas resident, as ``inferred heating demand per capita.''
 
@@
 Return periods define the probability with which a particular event can be expected to occur.
-Thus, a $T$-year event has a $\frac{1}{T}$ probability of occurring in a given year.
-For each event duration considered, we calculated return periods by fitting a stationary generalized extreme value (GEV) distribution to the time series of annual maxima of inferred heating demand per capita (see \cref{sec:inferred-demand}).
-Events that occur in December are coded to the following year so that a single winter season is grouped together.
+By definition, an event with return period $T$ years has a $\frac{1}{T}$ probability of occurring in a given year.
+
+For each event duration considered, we calculated return periods by fitting a stationary generalized extreme value (GEV) distribution to the time series of annual maxima of inferred heating demand per capita (in \cref{sec:res-agg}) or to the time series of $-T$, where $T$ is temperature (\cref{sec:res-spatial}).
+This negative value is analyzed because the GEV distribution is justified for block maxima, but we analyze annual minimum temperatures in \cref{sec:res-spatial}.
+Events that occur in December are coded to the following year so that a single December-February winter season is grouped together.
 The 2021 winter season was excluded from return period estimates, allowing us to interpret return periods for the February 2021 event as \emph{a priori} estimates.
@@
 Our open source code is freely available in a live repository at \url{https://github.com/jdossgollin/2021-TXtreme} and in an archived repository at \url{https://dx.doi.org/10.5281/zenodo.4568923}.
-Reproducibility is facilitated through the Snakemake library \cite{koster_snakemake:2012}.
 
-\section{How extreme was inferred heating demand per capita over the Texas Interconnection?}
+\section{How extreme was inferred heating demand per capita over the Texas Interconnection?}\label{sec:res-agg}
 
@@
 
-\section{Spatially distributed temperature extremes}
+\section{Spatially distributed temperature extremes}\label{sec:res-spatial}
 
@@
 
-\Cref{fig:local_era5} shows local return periods for the February 2021 cold snap at 6 hour, 1 day, 2 day, and 4 day durations.
+\Cref{fig:local_era5} shows local return periods for February 2021 temperature at 6 hour, 1 day, 2 day, and 4 day durations.
 Other than a band from south-central to south-east Texas, nearly all regions of the Texas Interconnection (gray outline in \cref{fig:local_era5}a,d) experienced cold with a return period below 50 years.
@@
 Importantly for the energy system, the band experiencing cold with return period greater than 50 years includes a substantial fraction of Texas’s population (\cref{fig:local_era5}a) and natural gas generation (\cref{fig:local_era5}d).
-Outside the Texas Interconnection region, the Midcontinent Independent System Operator and Southwest Power Pool instructed utilities to shed firm load, but of the approximately 4.89 million customers without power in Texas, Louisiana, and Oklahoma, an estimated 4.5 million were in Texas \cite{ceser_winterupdate:2021}.
+Outside the Texas Interconnection region, the Midcontinent Independent System Operator and Southwest Power Pool instructed utilities to shed firm load.
+Yet despite local return periods for temperature in central Oklahoma equal to or greater than local return periods for temperature in Texas, 92\% of the customers in Texas, Louisiana, and Oklahoma without power were in Texas \cite{ceser_winterupdate:2021}.
 
@@
 
-Our spatially aggregated metric of inferred heating demand per capita shows that the February 2021 event was intense but not unforeseeable (\cref{fig:idf_weighted}).
+Our spatially aggregated metric of inferred heating demand per capita shows that the February 2021 event was intense but not without precedent in the histoircal record (\cref{fig:idf_weighted}).
 Although specific locations experienced very intense ($>100$ year return period) temperatures, we find that for most locations in Texas the temperatures recorded during the February 2021 cold snap had precedent in the historical record.
 
-A proximate cause of load shedding in the Texas Interconnection during February 2021 was the vulnerability of the electricity generation system to the severe cold temperatures \cite{everhart_iea:2021}.
+A proximate cause of load shedding in the Texas Interconnection during February 2021 was the vulnerability of the electricity generation system to cold \cite{everhart_iea:2021}.
 As shown in \cref{fig:egova}, generator outages occurred across the state, even though most parts of the state had previously experienced similarly intense cold, notably in 1989 (\cref{fig:local_era5,fig:local_ghcnd}).
-Yet despite temperatures that were, in aggregate, more intense, the electricity system experienced fewer than three hours of rolling blackouts from December 21-23, 1989 \cite{nerc_operation:1989,osborne_twofreezes:2021}.
-Following the 1983, 1989, and 2011 cold snaps, the North American Electric Reliability Corporation (NERC) warned that ``constraints on natural gas fuel supplies to generating plants'' and ``generating unit trips, derates, or failure to start due to weather related causes'' were common themes across severe and moderate cold extremes \cite{nerc_previous:2013}, foreshadowing many of the causes of February 2021 energy system failures identified by ERCOT \cite{ercotpublic_outagesv2:2021,magness_review:2021}.
+Yet despite temperatures that were, in aggregate, more intense, the Texas Interconnection experienced fewer than three hours of rolling blackouts from December 21-23, 1989 \cite{nerc_operation:1989,osborne_twofreezes:2021}.
+Following the 1983, 1989, and 2011 cold snaps, the North American Electric Reliability Corporation (NERC) identified ``constraints on natural gas fuel supplies to generating plants'' and ``generating unit trips, derates, or failure to start due to weather related causes'' as key vulnerabilities \cite{nerc_previous:2013}, foreshadowing many of the causes of February 2021 energy system failures identified by ERCOT \cite{ercotpublic_outagesv2:2021,magness_review:2021}.
 While our analysis neglects other meteorological factors, like freezing rain, that may have impeded operations at specific facilities, we find that the February 2021 failures of energy and electricity systems in the Texas Interconnection took place during temperatures with precedent in the historical record.
@@
 Another cause of load shedding was the high demand for electricity that low temperatures induced.
-Currently at least 50\% of Texas households use electricity for space heating over the majority of census tracts of the state \cite{waite_heating:2020} and further electrification is a central element of many plans to decarbonize our energy sector \cite{williams_decarbonization:2012,davis:2018,white_txresidential:2019}.
+In fact, around 55\% of both residential and commercial spaces in Texas are currently heated using electricity \cite{waite_heating:2020} and further electrification is a central element of many plans to decarbonize the energy sector \cite{williams_decarbonization:2012,davis:2018,white_txresidential:2019}.
 While summer peak loads have been a central planning concern on the Texas grid in the past, it is likely that winter peak loads will become a greater concern in the coming decades.
 In fact, the estimated \SI{76819}{\mega\watt} of peak demand without load shedding during this event \cite{magness_review:2021} exceeded not only the previous winter demand record of \SI{65900}{\mega\watt} recorded on January 17, 2018 but also the all-season record actual demand of \SI{74800}{\mega\watt} recorded on August 19, 2019 \cite{everhart_iea:2021}.
-As electrification of heating continues, severe cold snaps may consistently represent the peak demand on the Texas Interconnect.
+As electrification of heating continues, severe cold snaps may drive peak demands on the Texas Interconnection.
 
@@
 Similarly, decisions at multiple time scales, including disaster preparedness and risk communication, contribute to the human consequences of physical infrastructure failure.
-Thus, the exact chain of events that led to the blackouts and water system disruptions during February 2021 should be sorted out only after thorough investigations by parties on the ground in Texas.
+Thus, the exact chain of events that led to the blackouts and water system disruptions during February 2021 should be sorted out only after further investigations by parties on the ground in Texas.
 
@@
 In addition to the record cold conditions of 1899 (\cref{fig:historic_bk}), we estimate that the weather of December 1989 would have resulted in higher 6-hour and 2-day values of inferred heating demand per capita over the Texas Interconnection than the February 2021 event.
-Storms in February 1951, January 1962, and December 1983 would have resulted in at least 90\% as much inferred heating demand per capita.
+Storms in February 1951, January 1962, and December 1983 would have resulted in at least 90\% as much inferred heating demand per capita at 24 and 48 hour durations.
 Given upward trends in the electrification of heating, it is likely that future cold snaps will cause peak annual loads on the Texas Interconnection to occur during the winter season.
@@
     Map produced using ERCOT's Generator Outage/Derate Visualization App (EGOVA) available at \url{https://bit.ly/EGOVA}.
+    Dots show the capacity reduction compared to the full rated capacity.
+    However, as discussed in ref.~\cite{ercotpublic_outagesv2:2021}, ERCOT's seasonal assessment of resource adequacy does not assume that all generators will provide 100\% of their rated capacity.
+    This is particularly true of intermittent resources, so this metric will tend to over-state the generator outages associated with wind and solar resources.
+    We retain this figure only to show that outages occurred across the state.
   }\label{fig:egova}

### 2021-05-06 doc/submission_02.tex
update caption of supplemental fig s8 following discussion w/ Edgar Virguez

@@
 Alison Silverstein,
+Vivek Srikrishnan,
 and
-Vivek Srikrishnan
+Edgar Virgüez
 for discussion of this work.
@@
     Map produced using ERCOT's Generator Outage/Derate Visualization App (EGOVA) available at \url{https://bit.ly/EGOVA}.
-    Dots show the capacity reduction compared to the full rated capacity.
-    However, as discussed in ref.~\cite{ercotpublic_outagesv2:2021}, ERCOT's seasonal assessment of resource adequacy does not assume that all generators will provide 100\% of their rated capacity.
-    This is particularly true of intermittent resources, so this metric will tend to over-state the generator outages associated with wind and solar resources.
-    We retain this figure only to show that outages occurred across the state.
+    Dot sizes describe the capacity reduction for each outage or derate event, which may not be the same as the difference between the power produced and what ERCOT anticipated, particularly for intermittent resources \cite{ercotpublic_outagesv2:2021}.
   }\label{fig:egova}

### 2021-05-06 doc/submission_02.tex
after creating new release, update link to zenodo DOI in paper

@@
 We are committed to open science.
-Our open source code is freely available in a live repository at \url{https://github.com/jdossgollin/2021-TXtreme} and in an archived repository at \url{https://dx.doi.org/10.5281/zenodo.4568923}.
+Our open source code is freely available in a live repository at \url{https://github.com/jdossgollin/2021-TXtreme} and in an archived repository at \url{https://dx.doi.org/10.5281/zenodo.4739881}.
 

### 2021-05-06 doc/supplemental.tex
for submission, pull out the supplemental info

--- 2021-05-06 doc/submission_02.tex
@@
 \usepackage{siunitx}
-\usepackage{xr-hyper} % to cross-ref from review response
+\usepackage{xr} % to cross-ref from review response
+\externaldocument{./submission_02}
 \usepackage{hyperref}
@@
 
-\title{How unprecedented was the February 2021 Texas cold snap?}
+\title{Supplemental Information: How unprecedented was the February 2021 Texas cold snap?}
 
@@
 
-\begin{abstract}
-  Winter storm Uri brought severe cold to the southern United States in February 2021, causing a cascading failure of interdependent systems in Texas where infrastructure was not adequately prepared for such cold.
-  In particular, the failure of interconnected energy systems restricted electricity supply just as demand for heating spiked, leaving millions of Texans without heat or electricity, many for several days.
-  This motivates the question: did historical storms suggest that such temperatures were known to occur, and if so with what frequency?
-  We compute a temperature-based proxy for heating demand and use this metric to answer the question ``what would the aggregate demand for heating have been had historic cold snaps occurred with today's population?''.
-  We find that local temperatures and the inferred demand for heating per capita across the region served by the Texas Interconnection were more severe during a storm in December 1989 than during February 2021, and that cold snaps in 1951 and 1983 were nearly as severe.
-  Given anticipated population growth, future storms may lead to even greater infrastructure failures if adaptive investments are not made.
-  Further, electricity system managers should prepare for trends in electrification of heating to drive peak annual loads on the Texas Interconnection during severe winter storms.
-\end{abstract}
-\noindent{\it Keywords: Energy, Electricity, Texas, Natural Hazards, Climate Resilience}
-
 \submitto{\ERL}
@@
 
-\section{Introduction}
-
-Between February 14th and 17th, 2021, a  northern air mass blanketed much of the continental United States, causing anomalously low surface temperatures across the Great Plains.
-The state of Texas was particularly hard hit, with coincident and cascading failures of natural gas production, power generation, transportation, and water systems leaving millions of Texans without electricity, heat, and water, many for several days \cite{ceser_winterupdate:2021,clack_uri:2021,smead_eyesoftx:2021}.
-These failures disproportionately affected vulnerable populations \cite{dobbins_blackoutdisparity:2021}, left at least 111 Texans dead \cite{mulcahy_urideath:2021}, and brought the Texas electricity grid within minutes of collapse \cite{magness_review:2021}.
-
-Since production and distribution of electricity is possible under conditions far colder than any Texas experienced in February 2021, energy system failures reflect inadequate preparedness for cold.
-These failures occurred both because electricity demand exceeded projections, and because electricity supply failed to meet them.
-On the demand side, the Electric Reliability Council of Texas (ERCOT), which operates the Texas Interconnection bulk electric power system (hence ``Texas Interconnection''), estimated that the peak demand would have been \SI{76819}{\mega\watt} without load shedding \cite{magness_review:2021}.
-This surpassed ERCOT's ``extreme winter forecast'' of \SI{67208}{\mega\watt} in its seasonal assessment of resource adequacy \cite{ercotpublic_sarawinter:2020}.
-On the supply side, the Texas Interconnection experienced over \SI{30000}{\mega\watt} of lost output for two consecutive days due to outages and derates caused by cold temperatures \cite{ercotpublic_outagesv2:2021}.
-A large fraction of this supply shortfall, which exceeded ERCOT's worst-case scenario for forced outages, originated in the natural gas supply chain \cite{ercotpublic_outagesv2:2021,smead_eyesoftx:2021,ceser_winterupdate:2021}.
-
-If temperatures experienced in the region served by the Texas Interconnection were unprecedented, then this event might prompt discussion about the appropriate use of models to prepare for events that are theoretically possible, but beyond the observational record.
-On the other hand, historical precedent for such temperatures would suggest a broader lack of institutional and infrastructural readiness.
-It is therefore important to assess whether historical data offered a precedent for the temperatures observed during February 2021.
-
-To answer this question, we first compute the population weighted difference between observed temperatures and a standard indoor temperature of \SI{65}{\degree F} as a proxy for the unknown heating demand, then use standard statistical procedures to assess the probability with which the temperatures observed during February 2021 might have been expected to occur \emph{a priori}.
-We then supplement this with a spatially distributed analysis of how unexpected the cold experienced by local roads, water mains, gas pipelines, energy generation facilities, and critical infrastructure installations was across Texas.
-We conclude by discussing the implications of these findings for long-term electricity systems planning given anticipated population growth and electrification.
-
-\subsection{Previous Cold Snaps in Texas}
-
-Texas state climatologist John Nielsen-Gammon wrote in 2011 that ``winter weather is a danger to TX in part because it is so rare,'' \cite{nielsen-gammon_txacc:2011}.
-Previous cold snaps in Texas, notably in 1899, 1951, 1983, 1989, and 2011 (see \cref{fig:historic_era5,fig:historic_bk}), have affected both human and ecological systems.
-For example, the 1951 cold event caused a  significant die-off of fish life in the shallow Gulf Coast \cite{gunter_fishes:1951}.
-
-The specific spatiotemporal structure of a cold event, and its correspondence with population centers, determines the grid-wide demand for heating (see \cref{sec:inferred-demand}).
-The structure of the storm also drives the aggregated hazard to energy infrastructure, which has implications for the costs and benefits of infrastructure hardening.
-The spatiotemporal patterns of historical cold snaps in Texas are illustrated in \cref{fig:historic_era5,fig:historic_bk,fig:historic_tx}.
-Although the spatiotemporal structure of each event is distinct, it is apparent that cold extremes in Texas tend to co-occur with cold temperatures across much of the United States, particularly the Great Plains.
-While the 2021 event was severe, daily temperature extrema in Texas appear qualitatively comparable to historical events.
-The ``Great Blizzard'' of February 1899, shown in \cref{fig:historic_bk}, caused even more intense cold.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{historic_events_era5.pdf}
-  \caption{
-    Severe cold snaps that affect Texas and extend into the central United States have several precedents in the historical record.
-    Plot shows anomalies of daily mean temperatures from the ERA-5 reanalysis \cite{hersbach_era5:2020} for historic major cold events affecting Texas, defined as the departure from the seasonal (December-February) mean of the observational record.
-    Anomalies facilitate identification of large-scale weather patterns superimposed on long-term climatological averages.
-    Hourly temperatures are averaged to 1-day (a,d,g,j,m), 3-day (b,e,h,k,n), and 5-day (c,f,i,l,o) average temperature anomalies.
-  }\label{fig:historic_era5}
-\end{figure}
-
-\section{Data and Methods}
-
-We use three distinct datasets to analyze temperature minima in the region covered by the Texas Interconnection through the lens of distributed (each grid cell analyzed separately) and aggregated (weighted averages taken across space) extreme values analysis.
-
-\subsection{Datasets}
-
-We use three temperature datasets to ensure robust findings:
-\begin{enumerate}
-  \item Hourly \SI{2}{\meter} air temperature reanalysis on a \SI{0.25}{\degree} grid from the ERA-5 reanalysis project produced by the European Centre for Medium Range Weather Forecasting \cite{hersbach_era5:2020} and available  from the Copernicus Data Store (\url{https://cds.climate.copernicus.eu}) from 1950 to the present.
-        The period from 1950 to 1979 is released as a preliminary back extension.
-        All plots shown in the main text use the ERA-5 data, but supplemental figures use other data sets.
-  \item Daily mean, minimum and maximum temperatures, gridded to \SI{1}{\degree}, produced by Berkeley Earth (\url{http://berkeleyearth.org/data/}).
-        This gridded product is based on statistical analysis of station data and is available from 1880 to 2019.
-        This dataset is considered an experimental product, so we use it only for comparative purposes.
-  \item To complement blended gridded data products, we use station temperature data from the Global Historical Climatology Network (GHCN) dataset compiled by the National Ocean and Atmospheric Administration \cite{Menne:2012hk} and available at \url{https://www1.ncdc.noaa.gov/pub/data/ghcn/daily/}.
-        This dataset provides daily mean, maximum, and minimum temperature observations.
-        These measurements represent point measurements, which can differ in important ways from gridded products describing spatial averages due to the spatial heterogeneity of temperature fields.
-        We retain stations within the state of Texas if they provide at least 60 years of data and if they contain observations for the set of historical cold extremes shown in \cref{fig:historic_era5}.
-\end{enumerate}
-We also use population density data from the GPWv4 dataset \cite{ciesin_gpwv4:2016}, a list of power generation facilities from the US Energy Information Administration \cite{useia_generators:2021}, and a map of the Texas Interconnection \cite{useia_regions:2021}.
-
-\subsection{Inferred heating demand per capita}\label{sec:inferred-demand}
-
-Most space heating in Texas is either electric or gas \cite{waite_heating:2020} and the majority of power generation in the Texas Interconnection depends on natural gas \cite{everhart_iea:2021}.
-Stress on natural gas production and delivery was therefore just as important as the more visible stress on the electric system \cite{ercotpublic_outagesv2:2021}.
-
-The hourly or daily thermal energy requirement for space heating is primarily driven by how much lower the ambient temperature is than an indoor comfort temperature of \SI{65}{\degree F}.
-This relationship is often expressed in terms of heating degree days or hours.
-We therefore consider the difference between observed temperatures and a standard indoor temperature of \SI{65}{\degree F} as a proxy for thermal heating demand.
-We compute this value each hour for the ERA5 data, defining heating demand at each grid cell as $\text{HD}_t = \max (65 - T_t, 0)$, where $T_t$ is the temperature at hour $t$ in \si{\degree F}.
-The Berkeley Earth and GHCN datasets provide daily minimum and maximum temperatures, so we define heating demand at each grid cell or station as $\text{HD}_d = \max (65 -\frac{T_{\text{min},d} + T_{\text{max},d}}{2}, 0)$, where $T_{\text{min},d}$ is the minimum temperature recorded on day $d$ and $T_{\text{max},d}$ is the maximum temperature recorded on day $d$, both in \si{\degree F}.
-
-To assess how spatially correlated cold spells might affect the Texas electric grid, we average heating demand in space over the Texas Interconnection domain \cite{useia_regions:2021}, weighting each grid cell by 2020 population density \cite{ciesin_gpwv4:2016}.
-We refer to this spatially aggregated time series, which has the straightforward interpretation as the average heating demand experienced by a Texas resident, as ``inferred heating demand per capita.''
-
-\subsection{Return Period}
-
-Return periods define the probability with which a particular event can be expected to occur.
-By definition, an event with return period $T$ years has a $\frac{1}{T}$ probability of occurring in a given year.
-
-For each event duration considered, we calculated return periods by fitting a stationary generalized extreme value (GEV) distribution to the time series of annual maxima of inferred heating demand per capita (in \cref{sec:res-agg}) or to the time series of $-T$, where $T$ is temperature (\cref{sec:res-spatial}).
-This negative value is analyzed because the GEV distribution is justified for block maxima, but we analyze annual minimum temperatures in \cref{sec:res-spatial}.
-Events that occur in December are coded to the following year so that a single December-February winter season is grouped together.
-The 2021 winter season was excluded from return period estimates, allowing us to interpret return periods for the February 2021 event as \emph{a priori} estimates.
-
-\subsection{Cold Duration}
-
-The effect of cold temperatures on energy demand and critical infrastructure depends on how long the cold persists.
-Short duration cold snaps can kill plants, freeze exposed pipes, freeze wind turbines, and contribute to dangerous roadway conditions.
-Longer duration cold spells contribute to demand for heating and energy and cause pipes to burst even if they have some insulation.
-We calculate demand for heating by taking temporal averages over a range of durations from 1 hour to 4 days.
-
-\subsection{Code and Data}
-
-We are committed to open science.
-Our open source code is freely available in a live repository at \url{https://github.com/jdossgollin/2021-TXtreme} and in an archived repository at \url{https://dx.doi.org/10.5281/zenodo.4739881}.
-
-\section{How extreme was inferred heating demand per capita over the Texas Interconnection?}\label{sec:res-agg}
-
-The total shock to Texas heating demand is partially determined by the extent to which cold snaps impact multiple population centers simultaneously.
-As such, understanding whether there was precedent for a cold snap simultaneously affecting several regions of Texas's grid that today have high population density is critical.
-We therefore use our measure of inferred heating demand per capita (see \cref{sec:inferred-demand}) to represent the aggregate heating demand induced by cold temperatures.
-Aggregating historic temperature fields in space using the 2020 population, we answer the question ``what would the aggregate demand for heating have been had historic cold snaps occurred today?''
-
-\Cref{fig:idf_weighted} shows that the intensity, duration, and recurrence intervals of the February 2021 storm are severe but not unprecedented in the historical record.
-For example, at the 6 hour duration the December 1989 storm was substantially more intense and other storms including February 1951 were nearly as intense.
-At the two day duration, the 2021 and 1989 events were approximately equally intense and other storms including December 1983 were nearly as intense.
-The 2011 storm, which caused rolling blackouts and motivated research into the energy system's vulnerability to cold \cite{ferc_outages:2011}, was quite modest by comparison.
-The right panel shows statistical return periods for these extreme events.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{ERCOT_HDD_IDF_MLE_popweighted.pdf}
-  \caption{
-    The inferred heating demand per capita induced by the February 2021 cold snap is not unprecedented.
-    For the worst 6 hours, the 1989 event was more severe than the 2021 event, while they are comparable for longer durations.
-    (a): time series of annual maximum inferred heating demand per capita (\cref{sec:inferred-demand}) at 6 hour and 2 day durations.
-    December extremes, including the December 1989 storm, are coded to the following year so that one maximum per December-February winter season is taken.
-    (b): the intensity-duration-frequency intervals estimated using 1950-2020 data (i.e., not using the 2021 event), overlaid by the annual maxima from the 1989, 2011, and 2021 events.
-    Gray dashed lines indicate 2, 5, 10, 25, 50, and 100 year return levels.
-  }\label{fig:idf_weighted}
-\end{figure}
-
-\section{Spatially distributed temperature extremes}\label{sec:res-spatial}
-
-It is difficult to establish a spatially aggregated proxy for supply-side risk given complex interlinkages between natural gas, electric, and other systems which create the possibility for cascading failures as observed in February 2021.
-Water treatment and distribution systems, as well as other essential services, also rely on electricity, further increasing vulnerabilities.
-Instead of aggregating this risk in space, we estimate the exceedance probability of the February 2021 temperatures at each grid cell separately to shed light on the severity of cold experienced by installations across the region.
-
-\Cref{fig:local_era5} shows local return periods for February 2021 temperature at 6 hour, 1 day, 2 day, and 4 day durations.
-Other than a band from south-central to south-east Texas, nearly all regions of the Texas Interconnection (gray outline in \cref{fig:local_era5}a,d) experienced cold with a return period below 50 years.
-Results are similar using station data (\cref{fig:local_ghcnd}).
-Importantly for the energy system, the band experiencing cold with return period greater than 50 years includes a substantial fraction of Texas’s population (\cref{fig:local_era5}a) and natural gas generation (\cref{fig:local_era5}d).
-Outside the Texas Interconnection region, the Midcontinent Independent System Operator and Southwest Power Pool instructed utilities to shed firm load.
-Yet despite local return periods for temperature in central Oklahoma equal to or greater than local return periods for temperature in Texas, 92\% of the customers in Texas, Louisiana, and Oklahoma without power were in Texas \cite{ceser_winterupdate:2021}.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{local_rt_era5.pdf}
-  \caption{
-    Return periods for the February 2021 event, calculated using stationary estimates of annual extremes over the period 1950-2020.
-    Return periods are calculated separately for each cell.
-    (a): estimates of 2020 population density \cite{ciesin_gpwv4:2016}.
-    (d): energy generation facilities in Texas \cite{useia_generators:2021}.
-    (b,c,e,f): local return periods for 6 hour, 1 day, 2 day, and 4 day durations, respectively.
-    Contours enclose regions that recorded 50 and 100 year return levels.
-    The gray region in panels (a) and (d) shows boundaries of the Texas Interconnection \cite{useia_regions:2021}.
-  }\label{fig:local_era5}
-\end{figure}
-
-\section{Discussion}
-
-Our spatially aggregated metric of inferred heating demand per capita shows that the February 2021 event was intense but not without precedent in the histoircal record (\cref{fig:idf_weighted}).
-Although specific locations experienced very intense ($>100$ year return period) temperatures, we find that for most locations in Texas the temperatures recorded during the February 2021 cold snap had precedent in the historical record.
-
-A proximate cause of load shedding in the Texas Interconnection during February 2021 was the vulnerability of the electricity generation system to cold \cite{everhart_iea:2021}.
-As shown in \cref{fig:egova}, generator outages occurred across the state, even though most parts of the state had previously experienced similarly intense cold, notably in 1989 (\cref{fig:local_era5,fig:local_ghcnd}).
-Yet despite temperatures that were, in aggregate, more intense, the Texas Interconnection experienced fewer than three hours of rolling blackouts from December 21-23, 1989 \cite{nerc_operation:1989,osborne_twofreezes:2021}.
-Following the 1983, 1989, and 2011 cold snaps, the North American Electric Reliability Corporation (NERC) identified ``constraints on natural gas fuel supplies to generating plants'' and ``generating unit trips, derates, or failure to start due to weather related causes'' as key vulnerabilities \cite{nerc_previous:2013}, foreshadowing many of the causes of February 2021 energy system failures identified by ERCOT \cite{ercotpublic_outagesv2:2021,magness_review:2021}.
-While our analysis neglects other meteorological factors, like freezing rain, that may have impeded operations at specific facilities, we find that the February 2021 failures of energy and electricity systems in the Texas Interconnection took place during temperatures with precedent in the historical record.
-
-Another cause of load shedding was the high demand for electricity that low temperatures induced.
-In fact, around 55\% of both residential and commercial spaces in Texas are currently heated using electricity \cite{waite_heating:2020} and further electrification is a central element of many plans to decarbonize the energy sector \cite{williams_decarbonization:2012,davis:2018,white_txresidential:2019}.
-While summer peak loads have been a central planning concern on the Texas grid in the past, it is likely that winter peak loads will become a greater concern in the coming decades.
-In fact, the estimated \SI{76819}{\mega\watt} of peak demand without load shedding during this event \cite{magness_review:2021} exceeded not only the previous winter demand record of \SI{65900}{\mega\watt} recorded on January 17, 2018 but also the all-season record actual demand of \SI{74800}{\mega\watt} recorded on August 19, 2019 \cite{everhart_iea:2021}.
-As electrification of heating continues, severe cold snaps may drive peak demands on the Texas Interconnection.
-
-Our primary findings hold for an alternative gridded dataset and station data (see supplemental material).
-However, calculated return periods are sensitive to the method of estimation (\cref{fig:idf_lmoments_unweighted,fig:idf_lmoments_weighted}).
-Future analysis could address parametric uncertainty, model structure uncertainty \cite{wong_floodrisk:2018}, non-stationarity \cite{Milly:2008dg}, or regime-like modes of climate variability \cite{DossGollin:2019}.
-More fundamentally, an assessment of exposure to cold extremes over the next decades should consider the deeply uncertain distribution of future climate change, and the induced effect on cold extremes in Texas.
-Although a broad scientific consensus suggests the frequency of cold extremes should decrease under warming in most places \cite{ipcc_ar5:2014}, possible links between North-South temperature gradients and mid-latitude temperature extremes remains an area of active research \cite{Barnes:2013fp,Cohen:2014gx,Screen:2013ho,romanowsky_stratosphere:2019}.
-Regardless, the effect of climate change on peak demand for heating is likely to be small compared to the effect of rapid population growth which the Texas Water Development Board, for example, anticipates to be 40\% from 2020 to 2050 \cite{texaswaterplan:2012}.
-
-Our analysis quantifies the frequency with which the temperatures observed during February 2021 could have been expected to occur \emph{a priori}.
-Other factors also govern infrastructure performance and failure, including precipitation, the demand for natural gas in adjacent regions, and complex connections within and between regional systems.
-Similarly, decisions at multiple time scales, including disaster preparedness and risk communication, contribute to the human consequences of physical infrastructure failure.
-Thus, the exact chain of events that led to the blackouts and water system disruptions during February 2021 should be sorted out only after further investigations by parties on the ground in Texas.
-
-\section{Conclusions}
-
-The February 2021 cold snap was the most intense in 30 years, but was not without precedent in the full historical record.
-In addition to the record cold conditions of 1899 (\cref{fig:historic_bk}), we estimate that the weather of December 1989 would have resulted in higher 6-hour and 2-day values of inferred heating demand per capita over the Texas Interconnection than the February 2021 event.
-Storms in February 1951, January 1962, and December 1983 would have resulted in at least 90\% as much inferred heating demand per capita at 24 and 48 hour durations.
-Given upward trends in the electrification of heating, it is likely that future cold snaps will cause peak annual loads on the Texas Interconnection to occur during the winter season.
-Infrastructure expansion necessitated by a rapidly growing population offers Texas the opportunity to invest in a more resilient energy system.
-
-\section*{Acknowledgements}
-
-DJF was supported by a gift from Gates Ventures LLC to the Carnegie Institution for Science.
-The authors thank
-Daniel Cohan,
-Sylvia Dee,
-Elisabeth Gawthrop,
-Pedram Hassanzadeh,
-Fred Heutte,
-Adam Massmann,
-Alison Silverstein,
-Vivek Srikrishnan,
-and
-Edgar Virgüez
-for discussion of this work.
-The authors also thank the developers and maintainers of the open source packages we used, in particular the pydata and pangeo communities, and the many journalists and academics whose timely reporting and commentary has informed our thinking.
-
-\section*{Bibliography}
-\bibliographystyle{unsrt}
-\bibliography{references}
-
-\clearpage
-\appendix
 \section{Supplemental Results}

### 2021-12-06 doc/agu21/aggregate_rt.tex
Create first draft of AGU poster; also moved documents for ERL paper into their own folder

\begin{block}{Spatially Aggregated Temperature Extremes}
    Analyzing our ``inferred heating demand per capita'' (see Methods and Data) informs the question ``what would the aggregate demand for heating have been had historic cold snaps occurred with 2020's population?''
    \begin{framed}
        \begin{figure}
            \centering
            \includegraphics[width=\textwidth]{ERCOT_HDD_IDF_MLE_popweighted.pdf}\\
            \caption{
                \textbf{The inferred heating demand per capita induced by the February 2021 cold snap is not unprecedented.}
                (a): time series of annual maximum inferred heating demand per capita.
                (b): the intensity-duration-frequency intervals estimated using 1950-2020 data (i.e., not using the 2021 event), overlaid by the annual maxima from the 1989, 2011, and 2021 events.
                Gray dashed lines indicate 2, 5, 10, 25, 50, and 100 year return levels.
            }\label{fig:idf_weighted}
        \end{figure}
    \end{framed}
\end{block}

### 2021-12-06 doc/agu21/at-a-glance.tex
Create first draft of AGU poster; also moved documents for ERL paper into their own folder

\begin{block}{Human Impacts}
    \begin{itemize}
        \item Electricity demand would have surpassed the all-season record had generation kept up \cite{ercotpublic_outagesv1:2021}
        \item Over \SI{30000}{\mega\watt} of lost output (mostly natural gas; \cite{busby_cascadingrisks:2021}) in TX for $> \SI{2}{days}$
        \item Grid within minutes of catastrophic failure (\cref{fig:ercot-frequency})
        \item Over 100 people died \cite{mulcahy_urideath:2021}
        \item Estimated \$130 billion damages \cite{busby_cascadingrisks:2021}
        \item Electricity outages caused cascading failures of water supply and other critical infrastructure
        \item Marginalized communities disproportionately affected \cite{dobbins_blackoutdisparity:2021}
    \end{itemize}
    \begin{framed}
        \begin{figure}
            \centering
            \includegraphics[width=\textwidth]{magness_frequency.pdf}
            \caption{
                As demand spiked and generation failed, the Texas grid came within minutes of catastrophic failure.
                Figure from ERCOT~\cite{magness_review:2021}.
            }
            \label{fig:ercot-frequency}
        \end{figure}
    \end{framed}
\end{block}

### 2021-12-06 doc/agu21/how-cold.tex
Create first draft of AGU poster; also moved documents for ERL paper into their own folder

\begin{block}{Historic Cold Snaps}
    Visualizing temperature anomalies facilitates identification of large-scale weather patterns superimposed on long-term climatological averages.
    We compare the February 2021 cold snap (bottom row) to four historic cold snaps.
    \begin{framed}
        \begin{figure}
            \centering
            \includegraphics[width=\textwidth]{historic_events_era5.pdf}
            \caption{
                Anomalies of daily mean temperatures from the ERA-5 reanalysis \cite{hersbach_era5:2020} during previous cold snaps demonstrate a qualitative precedent for severe cold.
            }\label{fig:historic_era5}
        \end{figure}
    \end{framed}
\end{block}

### 2021-12-06 doc/agu21/inferred-heating.tex
Create first draft of AGU poster; also moved documents for ERL paper into their own folder

\begin{block}{Methods and Data}
    Engineers often use ``heating degree days'' to quantify the effect of cold temperature on people and buildings.
    We develop a spatially aggregated time series, which has the straightforward interpretation as the average heating demand experienced by a Texas resident, called ``\textbf{inferred heating demand per capita}.''
    \begin{enumerate}
        \item Take gridded (\SI{0.25}{\degree}) hourly temperature data from ERA-5 reanalysis \cite{hersbach_era5:2020} (validated using other datasets -- see online supporting information)
        \item Gridded 2020 population density from GPWv4 dataset \cite{ciesin_gpwv4:2016}
        \item For each hour $t$ and grid cell $s$: calculate
              $\text{HD}_{s,t} = \max (\SI{65}{\degree F} - T_{s,t}, \SI{0}{\degree F})$
        \item Average over the region served by the Texas Interconnection (\cref{fig:local_era5}a), weighting each grid cell by population density
    \end{enumerate}
    Finally, we compute return periods using maximum likelihood GEV models (validated using other methods -- see online supporting information).
\end{block}

### 2021-12-06 doc/agu21/key-findings.tex
Create first draft of AGU poster; also moved documents for ERL paper into their own folder

\begin{block}{Key Insights}
    \begin{enumerate}
        \item Had they occurred today, the December 1989 cold snap would have caused more demand for heating than the February 2021 storm, and several other storms would have caused at least 90\% as much
        \item The Texas grid is designed for summer peak demands, but is highly vulnerable to rare winter freezes
        \item Population growth and electrification necessitate investment in Texas's energy supply; these must be made with system resilience in mind
    \end{enumerate}
\end{block}

### 2021-12-06 doc/agu21/poster.tex
Create first draft of AGU poster; also moved documents for ERL paper into their own folder

\documentclass[serif,mathserif,final]{beamer}
\mode<presentation>{\usetheme{Lankton}}

\usepackage[
  orientation=landscape,
  width=48in,height=36in,%size=a0, % paper size
  scale=1.25 % font scale factor
  ]{beamerposter}
 \setbeamertemplate{caption}[numbered] % number figs

% Graphics Path
\usepackage{graphicx}
\graphicspath{{../../fig/}{./}}

\usepackage[ttscale=0.8]{libertine} % better font

% OTHER PACKAGE CALLS
\usepackage[utf8]{inputenc}
\usepackage{
  amsmath,
  amsfonts,
  amssymb,
  xspace,
  framed,
  siunitx,
  nth,
  physics,
  nicefrac,
}
\sisetup{round-mode = figures, round-precision = 3}
\usepackage{cleveref}

%------ USE BIBLATEX WITH EXTREMELY MINAMALIST STYLE -----
% Use BibLaTeX
\usepackage[
	backend=biber,
	url=false,
	isbn=false,
	doi=false,
	maxnames = 2,
	style=numeric-comp,
	uniquename=false,
	uniquelist=false,
	firstinits=true,
	sorting=none,
	natbib=false,
	isbn=false,
]{biblatex}
\addbibresource{library.bib}
\renewbibmacro{in:}{}

% One-paragraph bibliography environment
\defbibenvironment{bibliography}
{\list%
  {\printtext[labelnumberwidth]{%
    \printfield{prefixnumber}%
    \printfield{labelnumber}}%
  \ifentrytype{article}{}{}}%
    {\setlength{\leftmargin}{0pt}%
    \setlength{\topsep}{0pt}}%
    \renewcommand*{\makelabel}[1]{##1}}%
  {\endlist}%
{\mkbibitem}

% \mkbibitem just prints item label and non-breakable space
\makeatletter
\newcommand{\mkbibitem}{\@itemlabel\addnbspace}
\makeatother

% Add breakable space between bibliography items
\renewcommand*{\finentrypunct}{\addperiod\space}

% Set fields to clear
\AtEveryBibitem{%
	\clearfield{month}%
	\clearfield{day}%
  \clearlist {language}%why this has to be clear list.....
  \clearfield{pages}%
	\clearfield{pagetotal}%
	\clearfield{eprinttype}%
	\clearfield{eprint}%
	\clearfield{number}%
	\clearfield{volume}%
	\clearfield{issue}%
	\clearfield{name:given}
	\clearfield{name:first}
	\ifentrytype{article}{%
		\clearfield{title}%
	}%
}

%------- GLOSSARIES PACKAGE -------
%\usepackage[acronym]{glossaries}
%\makeglossaries

%-- Header and footer information ----------------------------------
\newcommand{\footleft}{\textbf{Paper}: How unprecedented was the February 2021 Texas cold snap? \textit{Environmental Research Letters} (2021). \href{https://doi.org/10.1088/1748-9326/ac0278}{doi.org/10.1088/1748-9326/ac0278}.}
\newcommand{\footright}{\textbf{Codes}: \href{https://github.com/jdossgollin/2021-TXtreme/}{github.com/jdossgollin/2021-TXtreme/}}
\title{How Unprecedented Was the February 2021 Texas Cold Snap?}
\author{James Doss-Gollin\inst{*}\inst{1} \quad David J. Farnham\inst{2}\inst{3} \quad Vijay Modi\inst{4} \quad Upmanu Lall\inst{5}}
\institute{\inst{*}\href{mailto:jdossgollin@rice.edu}{jdossgollin@rice.edu} \quad \inst{1} Department of Civil and Environmental Engineering, Rice University \quad  \inst{2} ClimateAI \quad \inst{3} Department of Global Ecology, Carnegie Institution for Science \\ \inst{4} Department of Mechanical Engineering, Columbia University \quad \inst{5} Department of Earth and Environmental Engineering, Columbia University}

%-------------------------------------------------------------------

%-- Main Document --------------------------------------------------
\begin{document}
\begin{frame}{}
  \begin{columns}[t]

    \nocite{doss-gollin_txtreme:2021,doss-gollin_codes_txtreme:2021}

    %-- Column 1 ---------------------------------------------------
    \begin{column}{0.275\linewidth}

      \input{at-a-glance.tex}
      \input{key-findings.tex}
      \begin{block}{References}
        \renewcommand*{\bibfont}{\footnotesize}
        \printbibliography[heading=none]
      \end{block}

    \end{column}%1

    %-- Column 2 ---------------------------------------------------
    \begin{column}{0.325\linewidth}

      \input{how-cold.tex}
      \input{inferred-heating.tex}

    \end{column}

    %-- Column 3 ---------------------------------------------------
    \begin{column}{0.325\linewidth}

      \input{aggregate_rt.tex}
      \input{spatial_rt.tex}

    \end{column}

  \end{columns}
\end{frame}
\end{document}

### 2021-12-06 doc/agu21/spatial_rt.tex
Create first draft of AGU poster; also moved documents for ERL paper into their own folder

--- 2021-05-06 doc/submission_02.tex
@@
-\documentclass[12pt]{iopart}
-
-\usepackage{graphicx}
-\graphicspath{{../fig/}}
-
-\usepackage{url}
-\usepackage{siunitx}
-\usepackage{xr-hyper} % to cross-ref from review response
-\usepackage{hyperref}
-\hypersetup{hidelinks}
-\usepackage{cleveref}
-
-\begin{document}
-
-\title{How unprecedented was the February 2021 Texas cold snap?}
-
-\author{James Doss-Gollin$^1$, David J. Farnham$^2$, Upmanu Lall,$^{3,4}$ and Vijay Modi$^5$}
-\address{$^1$ Department of Civil and Environmental Engineering, Rice University, Houston, TX, USA (ORCID 0000-0002-3428-2224)}
-\address{$^2$ Department of Global Ecology, Carnegie Institution for Science, Stanford, CA, USA (ORCID 0000-0002-6690-4251)}
-\address{$^3$ Columbia Water Center, Columbia University, New York, NY, USA (ORCID 0000-0003-0529-8128)}
-\address{$^4$ Department of Earth and Environmental Engineering, Columbia University, New York, NY, USA}
-\address{$^4$ Department of Mechanical Engineering, Columbia University, New York, NY, USA (ORCID 0000-0003-2513-0437)}
-\ead{jdossgollin@rice.edu}
-\vspace{10pt}
-
-\begin{abstract}
-  Winter storm Uri brought severe cold to the southern United States in February 2021, causing a cascading failure of interdependent systems in Texas where infrastructure was not adequately prepared for such cold.
-  In particular, the failure of interconnected energy systems restricted electricity supply just as demand for heating spiked, leaving millions of Texans without heat or electricity, many for several days.
-  This motivates the question: did historical storms suggest that such temperatures were known to occur, and if so with what frequency?
-  We compute a temperature-based proxy for heating demand and use this metric to answer the question ``what would the aggregate demand for heating have been had historic cold snaps occurred with today's population?''.
-  We find that local temperatures and the inferred demand for heating per capita across the region served by the Texas Interconnection were more severe during a storm in December 1989 than during February 2021, and that cold snaps in 1951 and 1983 were nearly as severe.
-  Given anticipated population growth, future storms may lead to even greater infrastructure failures if adaptive investments are not made.
-  Further, electricity system managers should prepare for trends in electrification of heating to drive peak annual loads on the Texas Interconnection during severe winter storms.
-\end{abstract}
-\noindent{\it Keywords: Energy, Electricity, Texas, Natural Hazards, Climate Resilience}
-
-\submitto{\ERL}
-\maketitle
-
-\section{Introduction}
-
-Between February 14th and 17th, 2021, a  northern air mass blanketed much of the continental United States, causing anomalously low surface temperatures across the Great Plains.
-The state of Texas was particularly hard hit, with coincident and cascading failures of natural gas production, power generation, transportation, and water systems leaving millions of Texans without electricity, heat, and water, many for several days \cite{ceser_winterupdate:2021,clack_uri:2021,smead_eyesoftx:2021}.
-These failures disproportionately affected vulnerable populations \cite{dobbins_blackoutdisparity:2021}, left at least 111 Texans dead \cite{mulcahy_urideath:2021}, and brought the Texas electricity grid within minutes of collapse \cite{magness_review:2021}.
-
-Since production and distribution of electricity is possible under conditions far colder than any Texas experienced in February 2021, energy system failures reflect inadequate preparedness for cold.
-These failures occurred both because electricity demand exceeded projections, and because electricity supply failed to meet them.
-On the demand side, the Electric Reliability Council of Texas (ERCOT), which operates the Texas Interconnection bulk electric power system (hence ``Texas Interconnection''), estimated that the peak demand would have been \SI{76819}{\mega\watt} without load shedding \cite{magness_review:2021}.
-This surpassed ERCOT's ``extreme winter forecast'' of \SI{67208}{\mega\watt} in its seasonal assessment of resource adequacy \cite{ercotpublic_sarawinter:2020}.
-On the supply side, the Texas Interconnection experienced over \SI{30000}{\mega\watt} of lost output for two consecutive days due to outages and derates caused by cold temperatures \cite{ercotpublic_outagesv2:2021}.
-A large fraction of this supply shortfall, which exceeded ERCOT's worst-case scenario for forced outages, originated in the natural gas supply chain \cite{ercotpublic_outagesv2:2021,smead_eyesoftx:2021,ceser_winterupdate:2021}.
-
-If temperatures experienced in the region served by the Texas Interconnection were unprecedented, then this event might prompt discussion about the appropriate use of models to prepare for events that are theoretically possible, but beyond the observational record.
-On the other hand, historical precedent for such temperatures would suggest a broader lack of institutional and infrastructural readiness.
-It is therefore important to assess whether historical data offered a precedent for the temperatures observed during February 2021.
-
-To answer this question, we first compute the population weighted difference between observed temperatures and a standard indoor temperature of \SI{65}{\degree F} as a proxy for the unknown heating demand, then use standard statistical procedures to assess the probability with which the temperatures observed during February 2021 might have been expected to occur \emph{a priori}.
-We then supplement this with a spatially distributed analysis of how unexpected the cold experienced by local roads, water mains, gas pipelines, energy generation facilities, and critical infrastructure installations was across Texas.
-We conclude by discussing the implications of these findings for long-term electricity systems planning given anticipated population growth and electrification.
-
-\subsection{Previous Cold Snaps in Texas}
-
-Texas state climatologist John Nielsen-Gammon wrote in 2011 that ``winter weather is a danger to TX in part because it is so rare,'' \cite{nielsen-gammon_txacc:2011}.
-Previous cold snaps in Texas, notably in 1899, 1951, 1983, 1989, and 2011 (see \cref{fig:historic_era5,fig:historic_bk}), have affected both human and ecological systems.
-For example, the 1951 cold event caused a  significant die-off of fish life in the shallow Gulf Coast \cite{gunter_fishes:1951}.
-
-The specific spatiotemporal structure of a cold event, and its correspondence with population centers, determines the grid-wide demand for heating (see \cref{sec:inferred-demand}).
-The structure of the storm also drives the aggregated hazard to energy infrastructure, which has implications for the costs and benefits of infrastructure hardening.
-The spatiotemporal patterns of historical cold snaps in Texas are illustrated in \cref{fig:historic_era5,fig:historic_bk,fig:historic_tx}.
-Although the spatiotemporal structure of each event is distinct, it is apparent that cold extremes in Texas tend to co-occur with cold temperatures across much of the United States, particularly the Great Plains.
-While the 2021 event was severe, daily temperature extrema in Texas appear qualitatively comparable to historical events.
-The ``Great Blizzard'' of February 1899, shown in \cref{fig:historic_bk}, caused even more intense cold.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{historic_events_era5.pdf}
-  \caption{
-    Severe cold snaps that affect Texas and extend into the central United States have several precedents in the historical record.
-    Plot shows anomalies of daily mean temperatures from the ERA-5 reanalysis \cite{hersbach_era5:2020} for historic major cold events affecting Texas, defined as the departure from the seasonal (December-February) mean of the observational record.
-    Anomalies facilitate identification of large-scale weather patterns superimposed on long-term climatological averages.
-    Hourly temperatures are averaged to 1-day (a,d,g,j,m), 3-day (b,e,h,k,n), and 5-day (c,f,i,l,o) average temperature anomalies.
-  }\label{fig:historic_era5}
-\end{figure}
-
-\section{Data and Methods}
-
-We use three distinct datasets to analyze temperature minima in the region covered by the Texas Interconnection through the lens of distributed (each grid cell analyzed separately) and aggregated (weighted averages taken across space) extreme values analysis.
-
-\subsection{Datasets}
-
-We use three temperature datasets to ensure robust findings:
-\begin{enumerate}
-  \item Hourly \SI{2}{\meter} air temperature reanalysis on a \SI{0.25}{\degree} grid from the ERA-5 reanalysis project produced by the European Centre for Medium Range Weather Forecasting \cite{hersbach_era5:2020} and available  from the Copernicus Data Store (\url{https://cds.climate.copernicus.eu}) from 1950 to the present.
-        The period from 1950 to 1979 is released as a preliminary back extension.
-        All plots shown in the main text use the ERA-5 data, but supplemental figures use other data sets.
-  \item Daily mean, minimum and maximum temperatures, gridded to \SI{1}{\degree}, produced by Berkeley Earth (\url{http://berkeleyearth.org/data/}).
-        This gridded product is based on statistical analysis of station data and is available from 1880 to 2019.
-        This dataset is considered an experimental product, so we use it only for comparative purposes.
-  \item To complement blended gridded data products, we use station temperature data from the Global Historical Climatology Network (GHCN) dataset compiled by the National Ocean and Atmospheric Administration \cite{Menne:2012hk} and available at \url{https://www1.ncdc.noaa.gov/pub/data/ghcn/daily/}.
-        This dataset provides daily mean, maximum, and minimum temperature observations.
-        These measurements represent point measurements, which can differ in important ways from gridded products describing spatial averages due to the spatial heterogeneity of temperature fields.
-        We retain stations within the state of Texas if they provide at least 60 years of data and if they contain observations for the set of historical cold extremes shown in \cref{fig:historic_era5}.
-\end{enumerate}
-We also use population density data from the GPWv4 dataset \cite{ciesin_gpwv4:2016}, a list of power generation facilities from the US Energy Information Administration \cite{useia_generators:2021}, and a map of the Texas Interconnection \cite{useia_regions:2021}.
-
-\subsection{Inferred heating demand per capita}\label{sec:inferred-demand}
-
-Most space heating in Texas is either electric or gas \cite{waite_heating:2020} and the majority of power generation in the Texas Interconnection depends on natural gas \cite{everhart_iea:2021}.
-Stress on natural gas production and delivery was therefore just as important as the more visible stress on the electric system \cite{ercotpublic_outagesv2:2021}.
-
-The hourly or daily thermal energy requirement for space heating is primarily driven by how much lower the ambient temperature is than an indoor comfort temperature of \SI{65}{\degree F}.
-This relationship is often expressed in terms of heating degree days or hours.
-We therefore consider the difference between observed temperatures and a standard indoor temperature of \SI{65}{\degree F} as a proxy for thermal heating demand.
-We compute this value each hour for the ERA5 data, defining heating demand at each grid cell as $\text{HD}_t = \max (65 - T_t, 0)$, where $T_t$ is the temperature at hour $t$ in \si{\degree F}.
-The Berkeley Earth and GHCN datasets provide daily minimum and maximum temperatures, so we define heating demand at each grid cell or station as $\text{HD}_d = \max (65 -\frac{T_{\text{min},d} + T_{\text{max},d}}{2}, 0)$, where $T_{\text{min},d}$ is the minimum temperature recorded on day $d$ and $T_{\text{max},d}$ is the maximum temperature recorded on day $d$, both in \si{\degree F}.
-
-To assess how spatially correlated cold spells might affect the Texas electric grid, we average heating demand in space over the Texas Interconnection domain \cite{useia_regions:2021}, weighting each grid cell by 2020 population density \cite{ciesin_gpwv4:2016}.
-We refer to this spatially aggregated time series, which has the straightforward interpretation as the average heating demand experienced by a Texas resident, as ``inferred heating demand per capita.''
-
-\subsection{Return Period}
-
-Return periods define the probability with which a particular event can be expected to occur.
-By definition, an event with return period $T$ years has a $\frac{1}{T}$ probability of occurring in a given year.
-
-For each event duration considered, we calculated return periods by fitting a stationary generalized extreme value (GEV) distribution to the time series of annual maxima of inferred heating demand per capita (in \cref{sec:res-agg}) or to the time series of $-T$, where $T$ is temperature (\cref{sec:res-spatial}).
-This negative value is analyzed because the GEV distribution is justified for block maxima, but we analyze annual minimum temperatures in \cref{sec:res-spatial}.
-Events that occur in December are coded to the following year so that a single December-February winter season is grouped together.
-The 2021 winter season was excluded from return period estimates, allowing us to interpret return periods for the February 2021 event as \emph{a priori} estimates.
-
-\subsection{Cold Duration}
-
-The effect of cold temperatures on energy demand and critical infrastructure depends on how long the cold persists.
-Short duration cold snaps can kill plants, freeze exposed pipes, freeze wind turbines, and contribute to dangerous roadway conditions.
-Longer duration cold spells contribute to demand for heating and energy and cause pipes to burst even if they have some insulation.
-We calculate demand for heating by taking temporal averages over a range of durations from 1 hour to 4 days.
-
-\subsection{Code and Data}
-
-We are committed to open science.
-Our open source code is freely available in a live repository at \url{https://github.com/jdossgollin/2021-TXtreme} and in an archived repository at \url{https://dx.doi.org/10.5281/zenodo.4739881}.
-
-\section{How extreme was inferred heating demand per capita over the Texas Interconnection?}\label{sec:res-agg}
-
-The total shock to Texas heating demand is partially determined by the extent to which cold snaps impact multiple population centers simultaneously.
-As such, understanding whether there was precedent for a cold snap simultaneously affecting several regions of Texas's grid that today have high population density is critical.
-We therefore use our measure of inferred heating demand per capita (see \cref{sec:inferred-demand}) to represent the aggregate heating demand induced by cold temperatures.
-Aggregating historic temperature fields in space using the 2020 population, we answer the question ``what would the aggregate demand for heating have been had historic cold snaps occurred today?''
-
-\Cref{fig:idf_weighted} shows that the intensity, duration, and recurrence intervals of the February 2021 storm are severe but not unprecedented in the historical record.
-For example, at the 6 hour duration the December 1989 storm was substantially more intense and other storms including February 1951 were nearly as intense.
-At the two day duration, the 2021 and 1989 events were approximately equally intense and other storms including December 1983 were nearly as intense.
-The 2011 storm, which caused rolling blackouts and motivated research into the energy system's vulnerability to cold \cite{ferc_outages:2011}, was quite modest by comparison.
-The right panel shows statistical return periods for these extreme events.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{ERCOT_HDD_IDF_MLE_popweighted.pdf}
-  \caption{
-    The inferred heating demand per capita induced by the February 2021 cold snap is not unprecedented.
-    For the worst 6 hours, the 1989 event was more severe than the 2021 event, while they are comparable for longer durations.
-    (a): time series of annual maximum inferred heating demand per capita (\cref{sec:inferred-demand}) at 6 hour and 2 day durations.
-    December extremes, including the December 1989 storm, are coded to the following year so that one maximum per December-February winter season is taken.
-    (b): the intensity-duration-frequency intervals estimated using 1950-2020 data (i.e., not using the 2021 event), overlaid by the annual maxima from the 1989, 2011, and 2021 events.
-    Gray dashed lines indicate 2, 5, 10, 25, 50, and 100 year return levels.
-  }\label{fig:idf_weighted}
-\end{figure}
-
-\section{Spatially distributed temperature extremes}\label{sec:res-spatial}
-
-It is difficult to establish a spatially aggregated proxy for supply-side risk given complex interlinkages between natural gas, electric, and other systems which create the possibility for cascading failures as observed in February 2021.
-Water treatment and distribution systems, as well as other essential services, also rely on electricity, further increasing vulnerabilities.
-Instead of aggregating this risk in space, we estimate the exceedance probability of the February 2021 temperatures at each grid cell separately to shed light on the severity of cold experienced by installations across the region.
-
-\Cref{fig:local_era5} shows local return periods for February 2021 temperature at 6 hour, 1 day, 2 day, and 4 day durations.
-Other than a band from south-central to south-east Texas, nearly all regions of the Texas Interconnection (gray outline in \cref{fig:local_era5}a,d) experienced cold with a return period below 50 years.
-Results are similar using station data (\cref{fig:local_ghcnd}).
-Importantly for the energy system, the band experiencing cold with return period greater than 50 years includes a substantial fraction of Texas’s population (\cref{fig:local_era5}a) and natural gas generation (\cref{fig:local_era5}d).
-Outside the Texas Interconnection region, the Midcontinent Independent System Operator and Southwest Power Pool instructed utilities to shed firm load.
-Yet despite local return periods for temperature in central Oklahoma equal to or greater than local return periods for temperature in Texas, 92\% of the customers in Texas, Louisiana, and Oklahoma without power were in Texas \cite{ceser_winterupdate:2021}.
-
+\begin{block}{Spatially Distributed Temperature Extremes}
+    It is difficult for a single index to capture supply-side risk given complex interlinkages between natural gas, electric, and other systems.
+    We thus estimate the exceedance probability of the February 2021 temperatures at each grid cell separately to shed light on the degree to which cold experienced by installations across the region was unprecedented.
+    \begin{framed}
         \begin{figure}
@@
             \caption{
-    Return periods for the February 2021 event, calculated using stationary estimates of annual extremes over the period 1950-2020.
+                \textbf{Although the exceedance probability of February 2021's cold was less than 1/100 for some locations, for most it was between 1/25 and 1/50.}
                 Return periods are calculated separately for each cell.
@@
         \end{figure}
-
-\section{Discussion}
-
-Our spatially aggregated metric of inferred heating demand per capita shows that the February 2021 event was intense but not without precedent in the histoircal record (\cref{fig:idf_weighted}).
-Although specific locations experienced very intense ($>100$ year return period) temperatures, we find that for most locations in Texas the temperatures recorded during the February 2021 cold snap had precedent in the historical record.
-
-A proximate cause of load shedding in the Texas Interconnection during February 2021 was the vulnerability of the electricity generation system to cold \cite{everhart_iea:2021}.
-As shown in \cref{fig:egova}, generator outages occurred across the state, even though most parts of the state had previously experienced similarly intense cold, notably in 1989 (\cref{fig:local_era5,fig:local_ghcnd}).
-Yet despite temperatures that were, in aggregate, more intense, the Texas Interconnection experienced fewer than three hours of rolling blackouts from December 21-23, 1989 \cite{nerc_operation:1989,osborne_twofreezes:2021}.
-Following the 1983, 1989, and 2011 cold snaps, the North American Electric Reliability Corporation (NERC) identified ``constraints on natural gas fuel supplies to generating plants'' and ``generating unit trips, derates, or failure to start due to weather related causes'' as key vulnerabilities \cite{nerc_previous:2013}, foreshadowing many of the causes of February 2021 energy system failures identified by ERCOT \cite{ercotpublic_outagesv2:2021,magness_review:2021}.
-While our analysis neglects other meteorological factors, like freezing rain, that may have impeded operations at specific facilities, we find that the February 2021 failures of energy and electricity systems in the Texas Interconnection took place during temperatures with precedent in the historical record.
-
-Another cause of load shedding was the high demand for electricity that low temperatures induced.
-In fact, around 55\% of both residential and commercial spaces in Texas are currently heated using electricity \cite{waite_heating:2020} and further electrification is a central element of many plans to decarbonize the energy sector \cite{williams_decarbonization:2012,davis:2018,white_txresidential:2019}.
-While summer peak loads have been a central planning concern on the Texas grid in the past, it is likely that winter peak loads will become a greater concern in the coming decades.
-In fact, the estimated \SI{76819}{\mega\watt} of peak demand without load shedding during this event \cite{magness_review:2021} exceeded not only the previous winter demand record of \SI{65900}{\mega\watt} recorded on January 17, 2018 but also the all-season record actual demand of \SI{74800}{\mega\watt} recorded on August 19, 2019 \cite{everhart_iea:2021}.
-As electrification of heating continues, severe cold snaps may drive peak demands on the Texas Interconnection.
-
-Our primary findings hold for an alternative gridded dataset and station data (see supplemental material).
-However, calculated return periods are sensitive to the method of estimation (\cref{fig:idf_lmoments_unweighted,fig:idf_lmoments_weighted}).
-Future analysis could address parametric uncertainty, model structure uncertainty \cite{wong_floodrisk:2018}, non-stationarity \cite{Milly:2008dg}, or regime-like modes of climate variability \cite{DossGollin:2019}.
-More fundamentally, an assessment of exposure to cold extremes over the next decades should consider the deeply uncertain distribution of future climate change, and the induced effect on cold extremes in Texas.
-Although a broad scientific consensus suggests the frequency of cold extremes should decrease under warming in most places \cite{ipcc_ar5:2014}, possible links between North-South temperature gradients and mid-latitude temperature extremes remains an area of active research \cite{Barnes:2013fp,Cohen:2014gx,Screen:2013ho,romanowsky_stratosphere:2019}.
-Regardless, the effect of climate change on peak demand for heating is likely to be small compared to the effect of rapid population growth which the Texas Water Development Board, for example, anticipates to be 40\% from 2020 to 2050 \cite{texaswaterplan:2012}.
-
-Our analysis quantifies the frequency with which the temperatures observed during February 2021 could have been expected to occur \emph{a priori}.
-Other factors also govern infrastructure performance and failure, including precipitation, the demand for natural gas in adjacent regions, and complex connections within and between regional systems.
-Similarly, decisions at multiple time scales, including disaster preparedness and risk communication, contribute to the human consequences of physical infrastructure failure.
-Thus, the exact chain of events that led to the blackouts and water system disruptions during February 2021 should be sorted out only after further investigations by parties on the ground in Texas.
-
-\section{Conclusions}
-
-The February 2021 cold snap was the most intense in 30 years, but was not without precedent in the full historical record.
-In addition to the record cold conditions of 1899 (\cref{fig:historic_bk}), we estimate that the weather of December 1989 would have resulted in higher 6-hour and 2-day values of inferred heating demand per capita over the Texas Interconnection than the February 2021 event.
-Storms in February 1951, January 1962, and December 1983 would have resulted in at least 90\% as much inferred heating demand per capita at 24 and 48 hour durations.
-Given upward trends in the electrification of heating, it is likely that future cold snaps will cause peak annual loads on the Texas Interconnection to occur during the winter season.
-Infrastructure expansion necessitated by a rapidly growing population offers Texas the opportunity to invest in a more resilient energy system.
-
-\section*{Acknowledgements}
-
-DJF was supported by a gift from Gates Ventures LLC to the Carnegie Institution for Science.
-The authors thank
-Daniel Cohan,
-Sylvia Dee,
-Elisabeth Gawthrop,
-Pedram Hassanzadeh,
-Fred Heutte,
-Adam Massmann,
-Alison Silverstein,
-Vivek Srikrishnan,
-and
-Edgar Virgüez
-for discussion of this work.
-The authors also thank the developers and maintainers of the open source packages we used, in particular the pydata and pangeo communities, and the many journalists and academics whose timely reporting and commentary has informed our thinking.
-
-\section*{Bibliography}
-\bibliographystyle{unsrt}
-\bibliography{references}
-
-\clearpage
-\appendix
-\section{Supplemental Results}
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{0}
-
-\subsection{Historic Extreme Temperatures}
-
-To complement \cref{fig:historic_era5}, we plot extreme cold temperatures using alternate data.
-First, \cref{fig:historic_bk} shows historic events over the Continental United States.
-The set of events is slightly different than that of \cref{fig:historic_era5}: the data set does not include the 2021 event, but does include the 1899 ``Great Blizzard.''
-The 1899 event shows more intense and persistent cold than the other events in the dataset.
-Next, \cref{fig:historic_tx} shows the same data as \cref{fig:historic_era5} but zooms in on Texas.
-The 1989 (g-i) and 2021 (m-o) appear to be the most severe events in this data set, and the 1-day cold extremes in the 1989 event are more intense than in the February 2021 event, consistent with results in the main text.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{historic_events_bk.pdf}
-  \caption{
-    As \cref{fig:historic_bk} but the Berkely Earth temperature data is used.
-    The dataset does not contain the 2021 event, but the ``Great Blizzard'' of February 1899 is included.
-    Spatial patterns of cold from this dataset are qualitatively similar to Figure 1.
-    The 1899 event emphasizes that the modern historical record does not yield a full sample from the full distribution of possible hazards.
-  }\label{fig:historic_bk}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{historic_events_era5_TX.pdf}
-  \caption{
-    As \cref{fig:historic_bk} but only Texas is shown.
-  }\label{fig:historic_tx}
-\end{figure}
-
-\subsection{Spatially distributed temperature extremes}
-
-To complement \cref{fig:local_era5}, we compute local return periods using station data from the GHCN \cite{Menne:2012hk}.
-\Cref{fig:local_ghcnd} shows the return periods of the February 2021 cold snap for 1, 2, 3, and 4 day durations.
-Only stations with at least 60 years of data are considered, and since the locations of these stations are not chosen at random, this does not constitute a representative sample of all points across Texas.
-However, the spatial pattern matches that of \cref{fig:local_era5}, with a band of severe cold stretching from south-central to eastern Texas and in the Texas Panhandle.
-
-\subsection{Inferred heating demand per capita}
-
-To complement our analysis of inferred heating demand per capita, we consider how results change as a function of two modeling decisions.
-First, we consider what happens if the spatial field demand for heating is aggregated using grid cell area rather than population density.
-Next, we compute return periods using an estimator based on the method of $L$-moments.
-Although $L$-moment estimators for the generalized extreme value distribution are not unbiased, they are popular in the statistical hydrology literature for their stability \cite{hosking_gev:1985,martins_gev:2001,morrison_gev:2002}.
-
-We draw two conclusions from these plots.
-First, the 2021 event appears more severe if grid cells are weighted by population density (\cref{fig:idf_weighted,fig:idf_lmoments_weighted}) than if they are weighted only by area \cref{fig:idf_unweighted,fig:idf_lmoments_unweighted}).
-This is consistent with our observation of a correspondence between the most extreme temperatures in February 2021 and population density (\cref{fig:local_era5}).
-By contrast, the 2011 event appears more extreme when grid cells are weighted by area, which is consistent with \cref{fig:historic_tx,fig:historic_era5,fig:historic_bk} showing the coldest temperatures in relatively less populated West Texas.
-Second, the $L$-moment estimators (\cref{fig:idf_lmoments_weighted,fig:idf_lmoments_unweighted}) assign a lower return period to the 1989 and 2021 events than the maximum likelihood estimators (\cref{fig:idf_unweighted,fig:idf_weighted}).
-
-To provide some context for our inferred demand for heating metric, \cref{fig:hdd_ts} plots its time series during the peak of the February 2021 cold snap.
-This reveals a rise from approximately \SI{10}{\degree F} to nearly \SI{60}{\degree F} during the peak of the February 2021 cold snap.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{local_rt_ghcnd.pdf}
-  \caption{
-    As \cref{fig:local_era5} but return periods are calculated using station data from the GHCN data set \cite{Menne:2012hk}.
-    Black circles indicate that a station exceeded its own record for a particular duration.
-  }\label{fig:local_ghcnd}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{ERCOT_HDD_IDF_MLE_unweighted.pdf}
-  \caption{
-    As \cref{fig:idf_unweighted} but grid cells are weighted by area $A=\cos(\phi)$ where $\phi$ is latitude.
-  }\label{fig:idf_unweighted}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{ERCOT_HDD_IDF_plotpos_popweighted.pdf}
-  \caption{
-    As \cref{fig:idf_unweighted} but return periods are calculated using the $L$-moments estimator.
-  }\label{fig:idf_lmoments_weighted}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{ERCOT_HDD_IDF_plotpos_unweighted.pdf}
-  \caption{
-    As \cref{fig:idf_unweighted} but return periods are calculated using the $L$-moments estimator.
-  }\label{fig:idf_lmoments_unweighted}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{HDD_pop_weighted_ts.pdf}
-  \caption{
-    A time series of inferred heating demand per capita over the Texas Interconnection during the February 2021 cold snap.
-  }\label{fig:hdd_ts}
-\end{figure}
-
-\begin{figure}
-  \includegraphics[width=\textwidth]{EGOVA.pdf}
-  \caption{
-    ERCOT generator outages from 12:00AM on 14 February 2021 to 12:00AM on 18 February 2021.
-    Map produced using ERCOT's Generator Outage/Derate Visualization App (EGOVA) available at \url{https://bit.ly/EGOVA}.
-    Dot sizes describe the capacity reduction for each outage or derate event, which may not be the same as the difference between the power produced and what ERCOT anticipated, particularly for intermittent resources \cite{ercotpublic_outagesv2:2021}.
-  }\label{fig:egova}
-\end{figure}
-
-\end{document}
+    \end{framed}
+\end{block}

### 2021-12-06 doc/erl/submission_02.tex
Create first draft of AGU poster; also moved documents for ERL paper into their own folder

--- 2021-05-06 doc/submission_02.tex
@@
 \usepackage{graphicx}
-\graphicspath{{../fig/}}
+\graphicspath{{../fig/}{./}}
 
@@
 \usepackage{siunitx}
-\usepackage{xr-hyper} % to cross-ref from review response
+\usepackage{xr} % to cross-ref from review response
 \usepackage{hyperref}
@@
 Texas state climatologist John Nielsen-Gammon wrote in 2011 that ``winter weather is a danger to TX in part because it is so rare,'' \cite{nielsen-gammon_txacc:2011}.
-Previous cold snaps in Texas, notably in 1899, 1951, 1983, 1989, and 2011 (see \cref{fig:historic_era5,fig:historic_bk}), have affected both human and ecological systems.
+Previous cold snaps in Texas, notably in 1899, 1951, 1983, 1989, and 2011 (see \cref{fig:historic_era5} and supplemental fig.~S1), have affected both human and ecological systems.
 For example, the 1951 cold event caused a  significant die-off of fish life in the shallow Gulf Coast \cite{gunter_fishes:1951}.
@@
 The structure of the storm also drives the aggregated hazard to energy infrastructure, which has implications for the costs and benefits of infrastructure hardening.
-The spatiotemporal patterns of historical cold snaps in Texas are illustrated in \cref{fig:historic_era5,fig:historic_bk,fig:historic_tx}.
+The spatiotemporal patterns of historical cold snaps in Texas are illustrated in \cref{fig:historic_era5} and supplemental figs.~S1 and S2.
 Although the spatiotemporal structure of each event is distinct, it is apparent that cold extremes in Texas tend to co-occur with cold temperatures across much of the United States, particularly the Great Plains.
 While the 2021 event was severe, daily temperature extrema in Texas appear qualitatively comparable to historical events.
-The ``Great Blizzard'' of February 1899, shown in \cref{fig:historic_bk}, caused even more intense cold.
+The ``Great Blizzard'' of February 1899, shown in supplemental fig.~S1, caused even more intense cold.
 
@@
 Other than a band from south-central to south-east Texas, nearly all regions of the Texas Interconnection (gray outline in \cref{fig:local_era5}a,d) experienced cold with a return period below 50 years.
-Results are similar using station data (\cref{fig:local_ghcnd}).
+Results are similar using station data (supplemental fig.~S3).
 Importantly for the energy system, the band experiencing cold with return period greater than 50 years includes a substantial fraction of Texas’s population (\cref{fig:local_era5}a) and natural gas generation (\cref{fig:local_era5}d).
@@
 A proximate cause of load shedding in the Texas Interconnection during February 2021 was the vulnerability of the electricity generation system to cold \cite{everhart_iea:2021}.
-As shown in \cref{fig:egova}, generator outages occurred across the state, even though most parts of the state had previously experienced similarly intense cold, notably in 1989 (\cref{fig:local_era5,fig:local_ghcnd}).
+As shown in supplemental fig.~S8, generator outages occurred across the state, even though most parts of the state had previously experienced similarly intense cold, notably in 1989 (\cref{fig:local_era5} and supplemental fig.~S3).
 Yet despite temperatures that were, in aggregate, more intense, the Texas Interconnection experienced fewer than three hours of rolling blackouts from December 21-23, 1989 \cite{nerc_operation:1989,osborne_twofreezes:2021}.
@@
 Our primary findings hold for an alternative gridded dataset and station data (see supplemental material).
-However, calculated return periods are sensitive to the method of estimation (\cref{fig:idf_lmoments_unweighted,fig:idf_lmoments_weighted}).
+However, calculated return periods are sensitive to the method of estimation (supplemental figs.~S5 and S6).
 Future analysis could address parametric uncertainty, model structure uncertainty \cite{wong_floodrisk:2018}, non-stationarity \cite{Milly:2008dg}, or regime-like modes of climate variability \cite{DossGollin:2019}.
@@
 The February 2021 cold snap was the most intense in 30 years, but was not without precedent in the full historical record.
-In addition to the record cold conditions of 1899 (\cref{fig:historic_bk}), we estimate that the weather of December 1989 would have resulted in higher 6-hour and 2-day values of inferred heating demand per capita over the Texas Interconnection than the February 2021 event.
+In addition to the record cold conditions of 1899 (supplemental fig.~S1), we estimate that the weather of December 1989 would have resulted in higher 6-hour and 2-day values of inferred heating demand per capita over the Texas Interconnection than the February 2021 event.
 Storms in February 1951, January 1962, and December 1983 would have resulted in at least 90\% as much inferred heating demand per capita at 24 and 48 hour durations.
@@
 
-\clearpage
-\appendix
-\section{Supplemental Results}
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{0}
-
-\subsection{Historic Extreme Temperatures}
-
-To complement \cref{fig:historic_era5}, we plot extreme cold temperatures using alternate data.
-First, \cref{fig:historic_bk} shows historic events over the Continental United States.
-The set of events is slightly different than that of \cref{fig:historic_era5}: the data set does not include the 2021 event, but does include the 1899 ``Great Blizzard.''
-The 1899 event shows more intense and persistent cold than the other events in the dataset.
-Next, \cref{fig:historic_tx} shows the same data as \cref{fig:historic_era5} but zooms in on Texas.
-The 1989 (g-i) and 2021 (m-o) appear to be the most severe events in this data set, and the 1-day cold extremes in the 1989 event are more intense than in the February 2021 event, consistent with results in the main text.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{historic_events_bk.pdf}
-  \caption{
-    As \cref{fig:historic_bk} but the Berkely Earth temperature data is used.
-    The dataset does not contain the 2021 event, but the ``Great Blizzard'' of February 1899 is included.
-    Spatial patterns of cold from this dataset are qualitatively similar to Figure 1.
-    The 1899 event emphasizes that the modern historical record does not yield a full sample from the full distribution of possible hazards.
-  }\label{fig:historic_bk}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{historic_events_era5_TX.pdf}
-  \caption{
-    As \cref{fig:historic_bk} but only Texas is shown.
-  }\label{fig:historic_tx}
-\end{figure}
-
-\subsection{Spatially distributed temperature extremes}
-
-To complement \cref{fig:local_era5}, we compute local return periods using station data from the GHCN \cite{Menne:2012hk}.
-\Cref{fig:local_ghcnd} shows the return periods of the February 2021 cold snap for 1, 2, 3, and 4 day durations.
-Only stations with at least 60 years of data are considered, and since the locations of these stations are not chosen at random, this does not constitute a representative sample of all points across Texas.
-However, the spatial pattern matches that of \cref{fig:local_era5}, with a band of severe cold stretching from south-central to eastern Texas and in the Texas Panhandle.
-
-\subsection{Inferred heating demand per capita}
-
-To complement our analysis of inferred heating demand per capita, we consider how results change as a function of two modeling decisions.
-First, we consider what happens if the spatial field demand for heating is aggregated using grid cell area rather than population density.
-Next, we compute return periods using an estimator based on the method of $L$-moments.
-Although $L$-moment estimators for the generalized extreme value distribution are not unbiased, they are popular in the statistical hydrology literature for their stability \cite{hosking_gev:1985,martins_gev:2001,morrison_gev:2002}.
-
-We draw two conclusions from these plots.
-First, the 2021 event appears more severe if grid cells are weighted by population density (\cref{fig:idf_weighted,fig:idf_lmoments_weighted}) than if they are weighted only by area \cref{fig:idf_unweighted,fig:idf_lmoments_unweighted}).
-This is consistent with our observation of a correspondence between the most extreme temperatures in February 2021 and population density (\cref{fig:local_era5}).
-By contrast, the 2011 event appears more extreme when grid cells are weighted by area, which is consistent with \cref{fig:historic_tx,fig:historic_era5,fig:historic_bk} showing the coldest temperatures in relatively less populated West Texas.
-Second, the $L$-moment estimators (\cref{fig:idf_lmoments_weighted,fig:idf_lmoments_unweighted}) assign a lower return period to the 1989 and 2021 events than the maximum likelihood estimators (\cref{fig:idf_unweighted,fig:idf_weighted}).
-
-To provide some context for our inferred demand for heating metric, \cref{fig:hdd_ts} plots its time series during the peak of the February 2021 cold snap.
-This reveals a rise from approximately \SI{10}{\degree F} to nearly \SI{60}{\degree F} during the peak of the February 2021 cold snap.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{local_rt_ghcnd.pdf}
-  \caption{
-    As \cref{fig:local_era5} but return periods are calculated using station data from the GHCN data set \cite{Menne:2012hk}.
-    Black circles indicate that a station exceeded its own record for a particular duration.
-  }\label{fig:local_ghcnd}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{ERCOT_HDD_IDF_MLE_unweighted.pdf}
-  \caption{
-    As \cref{fig:idf_unweighted} but grid cells are weighted by area $A=\cos(\phi)$ where $\phi$ is latitude.
-  }\label{fig:idf_unweighted}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{ERCOT_HDD_IDF_plotpos_popweighted.pdf}
-  \caption{
-    As \cref{fig:idf_unweighted} but return periods are calculated using the $L$-moments estimator.
-  }\label{fig:idf_lmoments_weighted}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{ERCOT_HDD_IDF_plotpos_unweighted.pdf}
-  \caption{
-    As \cref{fig:idf_unweighted} but return periods are calculated using the $L$-moments estimator.
-  }\label{fig:idf_lmoments_unweighted}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{HDD_pop_weighted_ts.pdf}
-  \caption{
-    A time series of inferred heating demand per capita over the Texas Interconnection during the February 2021 cold snap.
-  }\label{fig:hdd_ts}
-\end{figure}
-
-\begin{figure}
-  \includegraphics[width=\textwidth]{EGOVA.pdf}
-  \caption{
-    ERCOT generator outages from 12:00AM on 14 February 2021 to 12:00AM on 18 February 2021.
-    Map produced using ERCOT's Generator Outage/Derate Visualization App (EGOVA) available at \url{https://bit.ly/EGOVA}.
-    Dot sizes describe the capacity reduction for each outage or derate event, which may not be the same as the difference between the power produced and what ERCOT anticipated, particularly for intermittent resources \cite{ercotpublic_outagesv2:2021}.
-  }\label{fig:egova}
-\end{figure}
-
 \end{document}

### 2021-12-06 doc/agu21/at-a-glance.tex
Some clarification and wording changes

@@
     \begin{itemize}
-        \item Electricity demand would have surpassed the all-season record had generation kept up \cite{ercotpublic_outagesv1:2021}
-        \item Over \SI{30000}{\mega\watt} of lost output (mostly natural gas; \cite{busby_cascadingrisks:2021}) in TX for $> \SI{2}{days}$
+        \item Electricity demand would have broken all-season record \cite{ercotpublic_outagesv1:2021}
+        \item Over \SI{30000}{\mega\watt} of lost output (mostly natural gas; see ref.~\cite{ercotpublic_outagesv2:2021})
         \item Grid within minutes of catastrophic failure (\cref{fig:ercot-frequency})
@@
         \item Estimated \$130 billion damages \cite{busby_cascadingrisks:2021}
-        \item Electricity outages caused cascading failures of water supply and other critical infrastructure
+        \item Cascading failures of water supply and other critical infrastructure
         \item Marginalized communities disproportionately affected \cite{dobbins_blackoutdisparity:2021}
@@
             \centering
-            \includegraphics[width=\textwidth]{magness_frequency.pdf}
+            \includegraphics[width=0.8\textwidth]{magness_frequency.pdf}
             \caption{

### 2021-12-06 doc/agu21/how-cold.tex
Some clarification and wording changes

@@
             \centering
-            \includegraphics[width=\textwidth]{historic_events_era5.pdf}
+            \includegraphics[width=\textwidth, trim = 0.25in 0 0 0.75in, clip]{historic_events_era5.pdf}
             \caption{
-                Anomalies of daily mean temperatures from the ERA-5 reanalysis \cite{hersbach_era5:2020} during previous cold snaps demonstrate a qualitative precedent for severe cold.
+                \textbf{Previous cold snaps demonstrate a qualitative precedent for severe cold.}
+                Data shows anomalies of daily mean temperatures from the ERA-5 reanalysis \cite{hersbach_era5:2020}.
             }\label{fig:historic_era5}

### 2021-12-06 doc/agu21/key-findings.tex
Some clarification and wording changes

@@
         \item Had they occurred today, the December 1989 cold snap would have caused more demand for heating than the February 2021 storm, and several other storms would have caused at least 90\% as much
-        \item The Texas grid is designed for summer peak demands, but is highly vulnerable to rare winter freezes
-        \item Population growth and electrification necessitate investment in Texas's energy supply; these must be made with system resilience in mind
+        \item The Texas grid is designed for summer peaks, but is highly vulnerable to rare winter freezes
+        \item Weather at least as cold as that observed during February 2011, which formed and forms the basis for season-ahead preparation \cite{ercotpublic_sarawinter:2020,ercot_sarawinter:2021}, should be expected with probability $\approx$ 1/10
+        \item Population growth and electrification necessitate investment in Texas's energy supply; these must consider system resilience
     \end{enumerate}

### 2021-12-07 doc/agu21/aggregate_rt.tex
clarify what return periods are / mean

@@
 \begin{block}{Spatially Aggregated Temperature Extremes}
-    Analyzing our ``inferred heating demand per capita'' (see Methods and Data) informs the question ``what would the aggregate demand for heating have been had historic cold snaps occurred with 2020's population?''
+    We analyze \emph{inferred heating demand per capita} to understand the potential impacts of historic cold snaps if they occurred with today's population.
     \begin{framed}
@@
             \caption{
-                \textbf{The inferred heating demand per capita induced by the February 2021 cold snap is not unprecedented.}
-                (a): time series of annual maximum inferred heating demand per capita.
-                (b): the intensity-duration-frequency intervals estimated using 1950-2020 data (i.e., not using the 2021 event), overlaid by the annual maxima from the 1989, 2011, and 2021 events.
+                \textbf{The \emph{inferred heating demand per capita} induced by the February 2021 cold snap is not unprecedented.}
+                (a): time series of annual maximum \emph{inferred heating demand per capita}.
+                (b): the intensity-duration-frequency (IDF) intervals, estimated using 1950-2020 data, overlaid by annual maxima from 1989, 2011, and 2021.
+                Each dashed line shows a single return level (2, 5, 10, 25, 50, 100 year) for \emph{inferred heating demand per capita} averaged over a given duration ($x$-axis).
                 Gray dashed lines indicate 2, 5, 10, 25, 50, and 100 year return levels.

### 2021-12-07 doc/agu21/inferred-heating.tex
clarify what return periods are / mean

@@
     Engineers often use ``heating degree days'' to quantify the effect of cold temperature on people and buildings.
-    We develop a spatially aggregated time series, which has the straightforward interpretation as the average heating demand experienced by a Texas resident, called ``\textbf{inferred heating demand per capita}.''
+    We develop a spatially aggregated time series, which has the straightforward interpretation as the average heating demand experienced by a Texas resident, called ``\textit{\textbf{inferred heating demand per capita}}.''
     \begin{enumerate}
-        \item Take gridded (\SI{0.25}{\degree}) hourly temperature data from ERA-5 reanalysis \cite{hersbach_era5:2020} (validated using other datasets -- see online supporting information)
+        \item Gridded (\SI{0.25}{\degree}) hourly temperature data from ERA-5 reanalysis \cite{hersbach_era5:2020} (validated using other datasets -- see online supporting information)
         \item Gridded 2020 population density from GPWv4 dataset \cite{ciesin_gpwv4:2016}
